---
title: "Linear Models, GLS, and Seasonal Indicator Variables"
subtitle: "Chapter 5: Lesson 1 Handout"
format: html
editor: source
sidebar: false
---

```{r}
#| include: false
source("common_functions.R")

library(plotly)
library(ggplot2)
```

```{=html}
<script type="text/javascript">
 function showhide(id) {
    var e = document.getElementById(id);
    e.style.display = (e.style.display == 'block') ? 'none' : 'block';
 }
 
 function openTab(evt, tabName) {
    var i, tabcontent, tablinks;
    tabcontent = document.getElementsByClassName("tabcontent");
    for (i = 0; i < tabcontent.length; i++) {
        tabcontent[i].style.display = "none";
    }
    tablinks = document.getElementsByClassName("tablinks");
    for (i = 0; i < tablinks.length; i++) {
        tablinks[i].className = tablinks[i].className.replace(" active", "");
    }
    document.getElementById(tabName).style.display = "block";
    evt.currentTarget.className += " active";
 }    
</script>
```




```{r}
# Read in Women's Clothing Retail Sales data
retail_ts <- rio::import("data/retail_by_business_type.parquet") |>
  filter(naics == 44812) |>
  filter(month >= yearmonth(my("Jan 2004")) & month <= yearmonth(my("Dec 2006"))) 
```


```{r, fig.width=6, fig.asp=0.6}
#| echo: false

# Read in Women's Clothing Retail Sales data
retail_ts <- rio::import("data/retail_by_business_type.parquet") |>
  filter(naics == 44812) |>
  filter(month >= yearmonth(my("Jan 2004")) & month <= yearmonth(my("Dec 2006"))) |>
  mutate(month_seq = 1:n()) |>
  mutate(year = year(month)) |>
  mutate(month_num = month(month)) |>
  as_tsibble(index = month)

# Number of years of data we are considering...it would be better if this was computed from the data
number_of_years <- 3

retail_scatter_plot <- ggplot(retail_ts, aes(month_seq, sales_millions)) +
  # data points
  geom_point(color = "#56B4E9", size = 2) +
  # x-axis
  geom_segment(x = 0, xend = number_of_years * 12 + 0.75, y = 0, yend = 0,
               arrow = arrow(length = unit(0.25, "cm"))) +
  geom_text(x = number_of_years * 12 + 0.75, y = 200, label = "x") +
  # scale_x_continuous(breaks = 1:(number_of_years * 12)) +
  scale_x_continuous(breaks = seq(0, (number_of_years * 12), by = 2)) +
  # y-axis
  geom_segment(x = 0, xend = 0, y = 0, yend = 5400,
               arrow = arrow(length = unit(0.25, "cm"))) +
  geom_text(x = 0.5, y = 5500, label = "y") +
  ylim(0,5500) +
  # labels and formatting
  labs(
    x = "Month, t",
    y = "Sales (in Millions of U.S. Dollars)",
    title = paste0(
      "Retail Sales: Women's Clothing in Millions (",
      min(retail_ts$year),
      "-",
      max(retail_ts$year),
      ")")
  ) +
  theme_minimal() +
  theme(panel.grid.minor.x = element_blank()) +
  theme(plot.title = element_text(hjust = 0.5))

# Fit regression model
model <- lm(sales_millions ~ month_seq, data = retail_ts)

# Create data frame for the regression line
reg_line_df <- data.frame(month_seq = c(0, number_of_years * 12 + 1)) |>
  mutate(sales_millions = coef(model)[1] + coef(model)[2] * month_seq)

# Create data frame with line that goes through origin
# and is parallel to regression line
zero_int_reg_line_df <- data.frame(month_seq = c(0, number_of_years * 12 + 1)) |>
  mutate(sales_millions = coef(model)[2] * month_seq)

arrow_df0 <- retail_ts |>
  mutate(
    arrow_start = coef(model)[2] * month_seq,
    deviation = sales_millions - arrow_start
  ) |>
  as_tibble()

arrow_df <- arrow_df0 |>
  group_by(month_num) |>
  summarize(
    mean_deviation = mean(deviation),
    .groups = 'drop'
  ) %>%
  right_join(arrow_df0, by = "month_num") |>
  arrange(month)

# Display scatterplot with regression line
p <- retail_scatter_plot +
  # regression line
  geom_line(data = reg_line_df, linetype = "dashed", linewidth = 1, color = "#F0E442") 
  # # line through the origin, parallel to regression line
  # geom_line(data = zero_int_reg_line_df, linewidth = 1, color = "#D55E00") +
  # # Vertical arrows
  # geom_segment(aes(x = month_seq, xend = month_seq,
  #                  y = arrow_df$arrow_start, yend = arrow_df$arrow_start + arrow_df$mean_deviation),
  #              arrow = arrow(length = unit(0.25, "cm")),
  #              color = "#009E73")

y_intercept_retail <- coef(model)[1] |> round(0)
slope_retail <- coef(model)[2] |> round(0)

ggplotly(p)
```

Use the estimated regression equation 
$$
  \hat x_t = `r y_intercept_retail` + `r slope_retail` ~ t
$$
to do the following:

-   Find the equation for the line parallel to the regression line that passes through the origin.
-   Compute the deviation of each observed point from the line you just obtained.
-   Compute the value of $\hat \beta_i$ for each month, where $i = 1, 2, \ldots, 12$.
-   Compute the estimate of the time series for $t = 1, 2, \ldots, 36$.
-   Predict the value of the time series for the next six months.




::: {.callout-tip appearance="minimal"}

```{r}
#| results: asis
#| echo: false

retail_df1 <- retail_ts |>
  as_tibble() |>
  mutate(month_text = format(month, "%b %Y")) |>
  rename(t = month_seq) |>
  dplyr::select(month_text, month_num, t, sales_millions) |>
  mutate(
    alpha1_t = slope_retail * t,
    deviation = sales_millions - alpha1_t
  )

retail_df2 <- retail_df1 |>
  group_by(month_num) |>
  summarize(
    mean_deviation = mean(deviation) |> round(2),
    .groups = 'drop'
  ) |>
  right_join(retail_df1, by = "month_num") |>
  arrange(t)

retail_pred_df <- rio::import("data/retail_by_business_type.parquet") |>
  filter(naics == 44812) |>
  filter(month >= yearmonth(my("Jan 2007")) & month <= yearmonth(my("Jun 2007"))) |>
  mutate(month_seq = (1:n()) + nrow(retail_df1)) |>
  mutate(year = year(month)) |>
  mutate(month_num = month(month)) |>
  mutate(month_text = format(month, "%b %Y")) |>
  as_tibble() |>
  mutate(month_text = format(month, "%b %Y")) |>
  rename(t = month_seq) |>
  dplyr::select(month_text, month_num, t, sales_millions) |>
  mutate(
    alpha1_t = slope_retail * t,
    deviation = sales_millions - alpha1_t
  ) |> 
  mutate(
    sales_millions = as.integer(NA),
    deviation = as.numeric(NA)
  ) |>
  left_join(retail_df2 |> select(month_num, mean_deviation) |> unique(), by = join_by(month_num))  |>
  dplyr::select(month_text, t, #month_num, 
                sales_millions, alpha1_t, deviation, mean_deviation)

retail_df3 <- retail_df2 |>
  dplyr::select(month_text, t, #month_num, 
                sales_millions, alpha1_t, deviation, mean_deviation) |>
  bind_rows(retail_pred_df) |>
  mutate(estimate = alpha1_t + mean_deviation) |>
  replace_na_with_char(emdash) |>
  rename(
    "$$Date$$" = month_text,
    "$$t$$" = t,
    # "$$i$$" = month_num,
    "$$x_t$$" = sales_millions,
    "$$\\hat \\alpha_1 t$$" = alpha1_t,
    "$$x_t - \\hat \\alpha_1 t$$" = deviation,
    "$$\\hat\\beta_i$$" = mean_deviation,
    "$$\\hat x_t$$" = estimate
  )

retail_df4 <- retail_df3 |>
  replace_cells_with_char(rows = c(2, 14, 26), cols = 6:7)  |>
  replace_cells_with_char(rows = c(3:5, 15:17, 27:29), cols = 4:7) |>
  replace_cells_with_char(rows = 38, cols = 6:7) |>
  replace_cells_with_char(rows = 39:41, cols = c(4,6:7))

retail_df4 |>
  head(24) |>
  display_table() |>
  column_spec(1:3, width_min = "0.35in") |>
  column_spec(4:ncol(retail_df3), width_min = "1in")
  
retail_df4 |>
  tail(12+6) |>
  display_table() |>
  column_spec(1:3, width_min = "0.35in") |>
  column_spec(4:7, width_min = "1in")
  

```
:::