---
title: "Additive Models"
subtitle: "Chapter 1: Lesson 4"
format: html
editor: source
sidebar: false
---

```{r}
#| include: false
source("common_functions.R")

library(plotly)
library(ggplot2)
```

```{=html}
<script type="text/javascript">
 function showhide(id) {
    var e = document.getElementById(id);
    e.style.display = (e.style.display == 'block') ? 'none' : 'block';
 }
 
 function openTab(evt, tabName) {
    var i, tabcontent, tablinks;
    tabcontent = document.getElementsByClassName("tabcontent");
    for (i = 0; i < tabcontent.length; i++) {
        tabcontent[i].style.display = "none";
    }
    tablinks = document.getElementsByClassName("tablinks");
    for (i = 0; i < tablinks.length; i++) {
        tablinks[i].className = tablinks[i].className.replace(" active", "");
    }
    document.getElementById(tabName).style.display = "block";
    evt.currentTarget.className += " active";
 }    
</script>
```

## Learning Outcomes

{{< include outcomes/chapter_1_lesson_4_outcomes.qmd >}}



## Preparation

-   Read Sections 1.5.4-1.5.5 and 1.6


## Learning Journal Exchange (10 min)

-   Review another student's journal
-   What would you add to your learning journal after reading your partner's?
-   What would you recommend your partner add to their learning journal?
-   Sign the Learning Journal review sheet for your peer






<!-- ## Random walks (5-10 min) -->

<!-- ```          -->
<!-- -   Definition of a random walk -->
<!-- -   Example: tossing a coin and moving left or right one integer on a number line -->
<!-- ``` -->

<!-- ::: callout-solution -->
<!-- Please add simulation of random walk here -->
<!-- ::: -->









## Converting from a Data File to a Tsibble (5 min)

This is a demonstration of two ways to convert an Excel 
or csv data file to a tsibble.

```{r}
deaths_df <- rio::import("https://byuistats.github.io/timeseries/data/traffic_deaths.xlsx")

# Method 1: Create date from scratch based on pattern of rows
# This only works if the data are in ascending order with no missing values
# Note: This file is not in the right order, so this code gives the wrong tsibble
# unless you sort the Excel file before proceeding.
start_date <- lubridate::ymd("2017-01-01")
date_seq <- seq(start_date,
                start_date + months(nrow(deaths_df)-1),
                by = "1 months")
deaths_tibble <- tibble(
  dates = date_seq,
  year = lubridate::year(date_seq),
  month = lubridate::month(date_seq),
  value = pull(deaths_df, Deaths)
)

# Method 2: Build using the date information in the Excel file
deaths_tibble <- deaths_df |>
  mutate(
    date_str = paste("1", Month, Year),
    dates = dmy(date_str),
    year = lubridate::year(dates),
    month = lubridate::month(dates),
    value = Deaths
  ) |>
  dplyr::select(dates, year, month, value)  |> 
  tibble()

# Create the index variable and convert to a tsibble
deaths_ts <- deaths_tibble |>
  mutate(index = tsibble::yearmonth(dates)) |>
  as_tsibble(index = index) |>
  dplyr::select(index, dates, year, month, value) |>
  rename(deaths = value) # rename value to emphasize data context
```

This results in a tsibble. The first few rows are given here:

```{r}
#| echo: false
deaths_ts |> 
  head()
```

## Data Visualizations (5 min)

The following time plot illustrates the data in this time series.

### Time Series Plot

::: panel-tabset
#### Default y-axis

This is the plot R creates by default

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false

ggplotly(autoplot(deaths_ts, .vars = deaths) +  
  labs(
    x = "Month",
    y = "Traffic Fatalities",
    title = "Traffic Fatalities in the United States (by Month)"
  ) +
  theme(plot.title = element_text(hjust = 0.5)))
```



#### Adjusted y-axis

The vertical axis was adjusted in this plot, so it would begin at 0.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false

ggplotly(autoplot(deaths_ts, .vars = deaths) +
  labs(
    x = "Month",
    y = "Traffic Fatalities",
    title = "Traffic Fatalities in the United States (by Month)"
  ) +
  coord_cartesian(ylim = c(0, 4500)) +
  theme(plot.title = element_text(hjust = 0.5)))
```

:::



::: {.callout-tip icon=false title="Check Your Understanding"}

-   What do you notice?
    -   Does it seem like there is a trend in the time series?
    -   Is there evidence of a seasonal effect? If so, when is the number of fatalities particularly high? particularly low?

-   Which of the two time plots above is superior? Why?

:::



### Visualization of Seasonal Effects: Side-by-Side Box Plots

These side-by-side box plots illustrate the seasonal effect present in the data.

```{r}
#| code-fold: true
#| code-summary: "Show the code"

ggplotly(ggplot(deaths_ts, aes(x = factor(month), y = deaths)) +
    geom_boxplot() +
  labs(
    x = "Month Number",
    y = "Deaths",
    title = "Boxplots of Traffic Deaths by Month"
  ) +
  theme(plot.title = element_text(hjust = 0.5)))
```

::: {.callout-tip icon=false title="Check Your Understanding"}

-   How do the positions of the box plots correspond to the times of the year where you noted more (or less) deaths occurring?

:::

## Computing the Seasonally Adjusted Series (25 min)

Our objective is to find an estimate for the time series that does not fluctuate with the seasons. This is called the seasonally adjusted series.

### Centered Moving Average

First, we compute the centered moving average, $\hat m_t$. This computation was explored in detail in the previous lesson. This code can be used to compute the 12-month centered moving average.

```{r}
# computes the 12-month centered moving average (m_hat)
deaths_ts <- deaths_ts |> 
  mutate(
    m_hat = (
          (1/2) * lag(deaths, 6)
          + lag(deaths, 5)
          + lag(deaths, 4)
          + lag(deaths, 3)
          + lag(deaths, 2)
          + lag(deaths, 1)
          + deaths
          + lead(deaths, 1)
          + lead(deaths, 2)
          + lead(deaths, 3)
          + lead(deaths, 4)
          + lead(deaths, 5)
          + (1/2) * lead(deaths, 6)
        ) / 12
  )
```

To emphasize the computation of the centered moving average, the observed data values that were used to find $\hat m_t$ for December 2017 are shown in <span style="color:#009E73;">green</span> in the table below. 

### Estimated Monthly Additive Effect

The centered moving average, $\hat m_t$, is then used to compute the monthly additive effect, $\hat s_t$:

$$
  \hat s_t = x_t - \hat m_t
$$

#### Table 1: Computation of the Centered Moving Average, $\hat m_t$, and the Estimated Monthly Additive Effect, $\hat s_t$

```{r}
#| echo: false

# Constants defining the number of rows included in the sample tables
num_blank_rows <- 14
num_addl_rows <- 2


deaths_shat_ts <- deaths_ts |>
  dplyr::select(index, month, deaths, m_hat) |>
  mutate(s_hat = deaths - m_hat)

deaths_shat_ts |>
  round_df(1) |>
  mutate(
    m_hat = as.character(m_hat),
    s_hat = as.character(s_hat),
  ) |> 
  mutate(
    s_hat = ifelse(row_number() <= num_blank_rows, "______", s_hat),
    deaths = as.character(deaths)
  ) |>
  dplyr::select(-month) |>
  rename(
    Month = index,
    "Deaths $$x_t$$" = deaths,
    "$$ \\hat m $$" = m_hat,
    "$$ \\hat s $$" = s_hat
  ) |>
  color_specific_cell(row_num = 12, col_num = 3, color = "#009E73") |>
  color_specific_cell(row_num = 6, col_num = 2, color = "#009E73") |>
  color_specific_cell(row_num = 7, col_num = 2, color = "#009E73") |>
  color_specific_cell(row_num = 8, col_num = 2, color = "#009E73") |>
  color_specific_cell(row_num = 9, col_num = 2, color = "#009E73") |>
  color_specific_cell(row_num = 10, col_num = 2, color = "#009E73") |>
  color_specific_cell(row_num = 11, col_num = 2, color = "#009E73") |>
  color_specific_cell(row_num = 12, col_num = 2, color = "#009E73") |>
  color_specific_cell(row_num = 13, col_num = 2, color = "#009E73") |>
  color_specific_cell(row_num = 14, col_num = 2, color = "#009E73") |>
  color_specific_cell(row_num = 15, col_num = 2, color = "#009E73") |>
  color_specific_cell(row_num = 16, col_num = 2, color = "#009E73") |>
  color_specific_cell(row_num = 17, col_num = 2, color = "#009E73") |>
  color_specific_cell(row_num = 18, col_num = 2, color = "#009E73") |>
  display_table()
```


::: {.callout-tip icon=false title="Check Your Understanding"}

-   Working with your assigned partner, fill in the missing values of $\hat s_t$ in the table above. 

:::

### Seasonally Adjusted Means


Next, we need to compute the mean (across years) of $\hat s_t$ by month. To compute this, it can be convenient to organize the values of $\hat s_t$ in a table, where the rows give the year and the columns give the month. We will calculate the mean of the $\hat s_t$ values for each month. We will call this $\bar {\hat s_t}$, the unadjusted monthly additive components.

The overall mean of these unadjusted monthly additive components $\left( \bar {\bar {\hat s_t}} \right)$ will be reasonably close to, but not exactly zero. We adjust these values by subtracting their overall mean, $\bar{\bar {\hat s_t}}$, from from each of them:

$$
  \bar s_t = \bar {\hat s_t} - \bar{\bar {\hat s_t}}
$$

where $\bar {\hat s_t}$ is the mean of the $\hat s_t$ values corresponding to month $t$, and $\bar{\bar {\hat s_t}}$ is the mean of the $\bar {\hat s_t}$ values.

This yields $\bar s_t$, the seasonally adjusted mean for each month.

#### Table 2: Computation of $\bar s_t$

```{r}
#| echo: false

# Compute s_hat
deaths_shat_df <- deaths_ts |>
  data.frame() |>
  mutate(month = month(dates, label=TRUE)) |>
  round_df(1) |> # Round df to make the computations simpler
  mutate(s_hat = deaths - m_hat)

wider_df <- deaths_shat_df |>
  dplyr::select(year, month, s_hat) |> 
  pivot_wider(values_from = "s_hat", names_from = "month")

wider_df2 <- wider_df %>% 
  bind_rows(colMeans(wider_df[ , -c(1)], na.rm = TRUE)) |>
  round_df(1) |>
  mutate(
    Jan = round_as_text(Jan),
    Feb = round_as_text(Feb),
    Mar = round_as_text(Mar),
    Apr = round_as_text(Apr),
    May = round_as_text(May),
    Jun = round_as_text(Jun),
    Jul = round_as_text(Jul),
    Aug = round_as_text(Aug),
    Sep = round_as_text(Sep),
    Oct = round_as_text(Oct),
    Nov = round_as_text(Nov),
    Dec = round_as_text(Dec)
  ) |>
  mutate(year = ifelse(row_number() == n(), "Mean", year))

wider_df2 |>
  # Hide bar_s_t values for November and December
  mutate(Jul = ifelse(row_number() == 1, "______", Jul)) |>
  mutate(Aug = ifelse(row_number() == 1, "______", Aug)) |>
  mutate(Sep = ifelse(row_number() == 1, "______", Sep)) |>
  mutate(Oct = ifelse(row_number() == 1, "______", Oct)) |>
  mutate(Nov = ifelse(row_number() == 1, "______", Nov)) |>
  mutate(Dec = ifelse(row_number() == 1, "______", Dec)) |>
  mutate(Jan = ifelse(row_number() == 2, "______", Jan)) |>
  mutate(Feb = ifelse(row_number() == 2, "______", Feb)) |>
  mutate(Nov = ifelse(row_number() == n(), "______", Nov)) |>
  mutate(Dec = ifelse(row_number() == n(), "______", Dec)) |>
  rename(Year = year) |>
  color_specific_cell(1, 8, "#0072B2") |>
  color_specific_cell(1, 9, "#0072B2") |>
  color_specific_cell(1, 10, "#0072B2") |>
  color_specific_cell(1, 11, "#0072B2") |>
  color_specific_cell(1, 12, "#0072B2") |>
  color_specific_cell(1, 13, "#0072B2") |>
  color_specific_cell(2, 2, "#0072B2") |>
  color_specific_cell(2, 3, "#0072B2") |>
  color_last_row2("#0072B2") |>
  rbind(c("$$ \\bar s_t $$",rep("$$~$$ ______",12))) |>
  color_last_row2("#0072B2") |>
  display_table()
```

::: {.callout-tip icon=false title="Check Your Understanding"}

-   The table above gives the values of $\hat s_t$. Fill in the missing values in the first two rows of the table above. (Note you already computed these.)
-   The second-to-last row (labeled "Mean") gives the values of $\bar s_t$. Fill in the two missing numbers. We will call these 12 means $\bar {\hat s_t}$.
-   Compute the mean of the $\bar {\hat s_t}$ values. Call this value $\bar {\bar {\hat s_t}}$ (This number should be relatively close to 0.)
-   Subtract the grand mean $\bar {\bar {\hat s_t}}$ from the column means $\bar {\hat s_t}$ to get $\bar s_t$, the seasonally adjusted mean for month $t$.  (Note that the mean of the $\bar s_t$ values is 0.)

$$ \bar s_t = \bar {\hat s_t} - \bar {\bar {\hat s_t}} $$

$\bar s_t$ is the seasonally adjusted mean for month $t$. 

:::


### Computing the Random Component and the Seasonally Adjusted Series

We calculate the random component by subtracting the trend and seasonally adjusted mean from the time series:

$$
  \text{random component} = x_t - \hat m_t - \bar s_t
$$

The seasonally adjusted series is computed by subtracting $\bar s_t$ from each of the observed values:

$$
  \text{seasonally adjusted series} = x_t - \bar s_t
$$

Compute the values missing from the table below.

#### Table 3: Computation of $\bar s$, the Random Component, and the Seasonally Adjusted Time Series

```{r}
#| echo: false

# THIS CODE IS DUPLICATED BELOW, BUT IT IS NEEDED HERE TO GET THE adjusted_ts
# Compute s_hat
deaths_ts2 <- deaths_ts |>
  mutate(s_hat = deaths - m_hat)

# Compute the unadjusted_s_bar and s_bar
s_bar_df <- deaths_ts2 |>
  data.frame() |>
  group_by(month) |>
  summarize(unadjusted_s_bar = mean(s_hat, na.rm = TRUE)) |>
  mutate(s_bar_bar = mean(unadjusted_s_bar)) |>
  mutate(s_bar = unadjusted_s_bar - s_bar_bar) |>
  dplyr::select(month, s_bar, s_bar_bar)

# Get seasonally adjusted time series
adjusted_ts <- deaths_ts2 |>
  left_join(s_bar_df, by = join_by(month)) |>
  mutate(random = deaths - m_hat - s_bar) |>
  mutate(seasonally_adjusted_x = deaths - s_bar) |>
  dplyr::select(-s_bar_bar)

additional_rows_from_adjusted_ts <- head(adjusted_ts, num_blank_rows + num_addl_rows) |> 
  tail(num_addl_rows) |> 
  convert_df_to_char(1)

adjusted_ts |>
  data.frame() |>
  round_df(1) |>
  filter(row_number() <= num_blank_rows) |>
  mutate(
    s_hat = cell_spec(s_hat, color = ""),
    s_bar = "______",
    random = "______",
    seasonally_adjusted_x = "______"
  ) |>
  convert_df_to_char() |>
  bind_rows(additional_rows_from_adjusted_ts) |>
  dplyr::select(-dates, -year, -month) |>
  rename(
    Month = index,
    "Deaths $$x_t$$" = deaths,
    "$$ \\hat m $$" = m_hat,
    "$$ \\hat s $$" = s_hat,
    "$$ \\bar s $$" = s_bar,
    Random = random,
    "Seasonally Adjusted $$x_t$$" = seasonally_adjusted_x
  ) |>
  display_partial_table(num_blank_rows + num_addl_rows, 0)
```



## Class Activity: Computing the Additive Decomposition in R (10 min)

<!-- ### Visualizing the Seasonally Adjusted Series -->

This code calculates the decomposition, including the seasonally adjusted time series, beginning with the tsibble deaths_ts.

#### Table 4: First Few Rows of the Decomposition of the Traffic Deaths Time Series

```{r}
# Compute the additive decomposition for deaths_ts
deaths_decompose <- deaths_ts |>
  model(feasts::classical_decomposition(deaths,
          type = "add"))  |>
  components()
```

First few rows of the `deaths_decompose` tsibble:

```{r}
#| echo: false

deaths_decompose |>
  display_partial_table(14, 0)
```


<!-- Check Your Understanding -->

::: {.callout-tip icon=false title="Check Your Understanding"}

-   How do these values from R compare to the ones you computed above?

:::



```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false

ggplotly(autoplot(deaths_decompose))
```


The figure below illustrates the original time series (in black), the centered moving average $\hat m_t$ (in blue), and the seasonally adjusted series (in red).

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false

p <- deaths_decompose |>
  ggplot() +
  geom_line(data = deaths_decompose, aes(x = index, y = deaths), color = "black") +
  geom_line(data = deaths_decompose, aes(x = index, y = season_adjust), color = "#D55E00") +
  geom_line(data = deaths_decompose, aes(x = index, y = trend), color = "#0072B2") +
  labs(
    x = "Month",
    y = "Traffic Fatalities",
    title = "Traffic Fatalities in the United States (by Month)"
  ) +
  coord_cartesian(ylim = c(0,4500)) +
  theme(plot.title = element_text(hjust = 0.5))


ggplotly(p)
```

::: {.callout-tip icon=false title="Check Your Understanding"}

-   Do you observe a trend in the time series?
    -   What does this trend suggest?
-   In February 2020, there is a spike in the seasonally adjusted time series. Refer to the values in the original time series to explain why this occurred.
-   In April 2020, there is a dip in the seasonally adjusted time series. Refer to the values in the original time series to explain why this occurred. 
    -   What would explain this phenomenon?

:::



## Homework Preview (5 min)

-   Review upcoming homework assignment
-   Clarify questions






## Homework

::: {.callout-note icon="false"}
## Download Assignment

<!-- ## need to update href link to correct files when we get them -->

<a href="https://byuistats.github.io/timeseries/homework/homework_1_4.qmd" download="homework_1_4.qmd"> homework_1_4.qmd </a>
:::






<!-- Solutions  -->


<a href="javascript:showhide('Solutions')"
style="font-size:.8em;">Team Activity: Computational Practice</a>

::: {#Solutions style="display:none;"}

Solutions for the computations from this lesson

#### Table 3: (Solutions)

```{r}
#| echo: false

# THIS IS THE HARD WAY. JUST USE THE DECOMPOSITION FUNCTION ABOVE...

# Compute s_hat
deaths_ts2 <- deaths_ts |>
  mutate(s_hat = deaths - m_hat)

# Compute the unadjusted_s_bar and s_bar
s_bar_df <- deaths_ts2 |>
  data.frame() |>
  group_by(month) |>
  summarize(unadjusted_s_bar = mean(s_hat, na.rm = TRUE)) |>
  mutate(s_bar_bar = mean(unadjusted_s_bar)) |>
  mutate(s_bar = unadjusted_s_bar - s_bar_bar) |>
  dplyr::select(month, s_bar, s_bar_bar)

# Get seasonally adjusted time series
adjusted_ts <- deaths_ts2 |>
  left_join(s_bar_df, by = join_by(month)) |>
  mutate(random = deaths - m_hat - s_bar) |>
  mutate(seasonally_adjusted_x = deaths - s_bar) |>
  dplyr::select(-s_bar_bar)

adjusted_ts |> 
  round_df(1) |>
  filter(row_number() <= num_blank_rows) |>
  mutate(
    s_hat = cell_spec(s_hat, color = "#E69F00"),
    s_bar = cell_spec(s_bar, color = "#009E73"),
    random = cell_spec(random, color = "#009E73"),
    seasonally_adjusted_x = cell_spec(seasonally_adjusted_x, color = "#009E73")
  ) |>
  convert_df_to_char() |>
  bind_rows(additional_rows_from_adjusted_ts) |>
  select(-dates, -year, -month) |>
  rename(
    Month = index, 
    "Deaths $$x_t$$" = deaths,
    "$$ \\hat m_t $$" = m_hat, 
    "$$ \\hat s_t $$" = s_hat, 
    "$$ \\bar s_t $$" = s_bar,
    Random = random,
    "Seasonally Adjusted $$x_t$$" = seasonally_adjusted_x
  ) |>
  display_partial_table(num_blank_rows + num_addl_rows, 0)
```

#### Table 2: (Solutions)

```{r}
#| echo: false

s_bar_bar_value <- s_bar_df %>% 
  summarize(s_bar_bar = round(mean(s_bar_bar), 1)) |>
  dplyr::select(s_bar_bar) |>
  pull()

wider_df3 <- wider_df2 |>
  bind_rows(
    wider_df2 |> tail(1)
  ) 

# Subtract the mean of \bar s_t to make the mean of the adjusted values be zero.
for (j in 2:ncol(wider_df3)) {
  wider_df3[nrow(wider_df3), j] = 
    as.character(
      as.numeric(wider_df3[nrow(wider_df3), j]) - s_bar_bar_value
    )
}

wider_df3 |> 
  mutate(year = ifelse(row_number() == n(), "$$ \\bar s_t $$", year)) |>
  rename(Year = year) |>
  color_specific_cell(1, 8, "#0072B2") |>
  color_specific_cell(1, 9, "#0072B2") |>
  color_specific_cell(1, 10, "#0072B2") |>
  color_specific_cell(1, 11, "#0072B2") |>
  color_specific_cell(1, 12, "#0072B2") |>
  color_specific_cell(1, 13, "#0072B2") |>
  color_specific_cell(2, 2, "#0072B2") |>
  color_specific_cell(2, 3, "#0072B2") |>
  color_last_row2("#0072B2") |>
  color_2nd_to_last_row2("#0072B2") |>
  display_table()
```

:::