---
title: "Covariance and Correlation"
subtitle: "Chapter 2: Lesson 1"
format: html
editor: source
sidebar: false
---

```{r}
#| include: false
source("common_functions.R")

library(ggplot2)
library(plotly)
```

```{=html}
<script type="text/javascript">
 function showhide(id) {
    var e = document.getElementById(id);
    e.style.display = (e.style.display == 'block') ? 'none' : 'block';
 }
 
 function openTab(evt, tabName) {
    var i, tabcontent, tablinks;
    tabcontent = document.getElementsByClassName("tabcontent");
    for (i = 0; i < tabcontent.length; i++) {
        tabcontent[i].style.display = "none";
    }
    tablinks = document.getElementsByClassName("tablinks");
    for (i = 0; i < tablinks.length; i++) {
        tablinks[i].className = tablinks[i].className.replace(" active", "");
    }
    document.getElementById(tabName).style.display = "block";
    evt.currentTarget.className += " active";
 }    
</script>
```
## Learning Outcomes

{{< include outcomes/chapter_2_lesson_1_outcomes.qmd >}}



## Preparation

-   Read Sections 2.1-2.2.2 and 2.2.4

## Learning Journal Exchange (10 min)

-   Review another student's journal
-   What would you add to your learning journal after reading your partner's?
-   What would you recommend your partner add to their learning journal?
-   Sign the Learning Journal review sheet for your peer

## Class Activity: Variance and Standard Deviation (10 min)

We will explore the variance and standard deviation in this section.

::: {.callout-tip icon=false title="Check Your Understanding"}

-   What do the standard deviation and the variance measure?

:::

The following code simulates observations of a random variable. We will use these data to explore the variance and standard deviation.

```{r}
# Set random seed
set.seed(2412)

# Specify means and standard deviation
n <- 5        # number of points
mu <- 10      # mean
sigma <- 3    # standard deviation

# Simulate normal data
sim_data <- data.frame(x = round(rnorm(n, mu, sigma), 1)) |> 
  arrange(x)
```

```{r}
#| echo: false

# Convert to character string
# Initialize empty character vector
five_vals <- sim_data$x[1]
# Simulate normal data
for(t in 2:n){
  five_vals <- paste(five_vals, sim_data$x[t], sep=", ")
}
```

The data simulated by this process are:

<center>`r five_vals`</center>

::: {.callout-tip icon=false title="Check Your Understanding"}

-   Find the sample mean of these numbers. <!-- `r mean(sim_data$x)`. -->
-   What are some ways to interpret the mean?

:::

The variance and standard deviation are individual numbers that summarize how far the data are from the mean. We first compute the deviations from the mean, $x - \bar x$. This is the directed distance from the mean to each data point.

```{r}
#| echo: false
#| warning: false
temp <- sim_data |> 
  mutate(deviations = x - mean(x)) |> 
  arrange(desc(x))

mean_x <- mean(temp$x)
min_x <- floor(min(temp$x))
max_x <- ceiling(max(temp$x))
range <- max_x - min_x
lower <- min_x - range / 10
upper <- max_x + range / 10

ticks <- ceiling(lower):floor(upper)
ticks_df <- data.frame(x = ticks, y = -1)


# Plot deviations from the mean
ggplot(temp, aes(x = x, y = 0)) +
  # x-axis
  annotate("segment", x = lower, xend = upper, y = -1, yend = -1, colour = "black", linewidth = 1, arrow = arrow(length = unit(0.3,"cm"))) +
  
  # Add tick marks and labels          
  annotate("segment", x = ticks, xend = ticks, y = -1.25, yend = -0.75, colour = "black", linewidth = 0.5) +

  geom_text(aes(x = upper, y = -1, label = "x"), size = 4, hjust = -1, vjust = 0, color = "black") +
  
  geom_text(data = ticks_df, aes(x = x, y = y, label = x), size = 4, vjust = 2, color = "black") +
  
  # Deviations from the mean arrows and lines
  geom_segment(aes(x = mean_x, xend = x, y = 1:n, yend = 1:n), colour = okabeito_colors_list[2], linewidth = 1, arrow = arrow(length = unit(0.3,"cm"))) +
  geom_segment(aes(x = x, xend = x, y = 0.25, yend = 1:n - 0.25), colour = okabeito_colors_list[2], linewidth = 0.5, linetype = "dashed") +
  geom_text(aes(x = (mean_x + x)/2, y = 1:n, label = round(deviations, 2)), size = 3, vjust = -0.5) +
  
  # Marker for the mean
  annotate("segment", x = mean_x, xend = mean_x, y = -2.5, yend = -1.25, colour = okabeito_colors_list[1], linewidth = 1, arrow = arrow(length = unit(0.3,"cm"))) +
  geom_segment(aes(x = mean_x, xend = mean_x, y = -0.75, yend = n + 1), colour = okabeito_colors_list[1], linewidth = 0.5, linetype = "dashed") +
  # Add xbar
  geom_label(
    label=expression(bar(x) == " "),
    x=mean_x-0.1,
    y=-2.6,
    color = okabeito_colors_list[1],parse = TRUE,label.size = NA, fill=NA)+
  # add mean value
  geom_label(
    label=paste0(round(mean_x, 2)),
    x=mean_x+0.3,
    y=-2.6,
    color = okabeito_colors_list[1],label.size = NA, fill=NA)+
  geom_point(size = 3, color = okabeito_colors_list[2]) + 
  geom_text(aes(x = x, y = rep(0,n), label = x), size = 3, vjust = 1.75) +
  
  # theme
  theme_void() +
  theme(axis.title.y = element_blank()) +
  theme(plot.title = element_text(hjust = 0.5)) +
     
  theme(aspect.ratio = 0.4) +
  labs(title = "Deviations from the Mean", 
       x = "Value",
       y = "")
```

We can summarize this information in a table:

#### Table 1: Deviations from the mean

```{r}
#| echo: false
sim_data |> 
  mutate(
    xx = x - mean(x),
    extra1 = " ",
    extra2 = " ",
    extra3 = " ",
    extra4 = " "
    ) |> 
  rename(
    "$$x_t$$" = x,
    "$$x_t-\\bar x$$" = xx,
    " " = extra1,
    "  " = extra2,
    "   " = extra3,
    "    " = extra4
  ) |>
  display_table("0.75in")
```

::: {.callout-tip icon=false title="Check Your Understanding"}

How can we obtain one number that summarizes how spread out the data are from the mean? We may try averaging the deviations from the mean.

-   What is the average deviation from the mean?
-   Will we get the same value with other data sets, or is this just a coincidence?
-   What could you do to prevent this from happening?
-   Apply your idea. Compute the resulting value that summarizes the spread. What do you get?
-   What is the relationship between the sample variance and the sample standard deviation?
-   Use a table like the one above to verify that the sample variance is `r var(sim_data$x)`.
-   Show that the sample standard deviation is `r sd(sim_data$x) |> round(4)`.

:::

## Class Activity: Covariance and Correlation (15 min)

```{=html}
 <iframe id="CoAndCo" src="https://posit.byui.edu/content/564c2e71-3d0b-43a6-8c6f-d402125c8b28" style="border: none; width: 100%; height: 2200px" frameborder="0"></iframe>
```

::: {.callout-tip icon=false title="Check Your Understanding"}

-   What do you get if you multiply the equations for $r$, $s_x$, and $s_y$ together?

:::

$$
  r \cdot s_x \cdot s_y 
    =
      \frac{\sum\limits_{t=1}^n (x - \bar x)(y - \bar y)}{\sqrt{\sum\limits_{t=1}^n (x - \bar x)^2} \sqrt{\sum\limits_{t=1}^n (y - \bar y)^2}} 
      \cdot 
      \sqrt{ \frac{\sum\limits_{t=1}^n (x - \bar x)^2}{n-1} }
      \cdot 
      \sqrt{ \frac{\sum\limits_{t=1}^n (y - \bar y)^2}{n-1} }
    =
      ?
$$

::: {.callout-tip icon=false title="Check Your Understanding"}

-   Use the numerical values above to confirm your result. Any discrepancy is due to roundoff error.

:::


## Team Activity: Computational Practice (15 min)

```{r}
#| echo: false

# Set random seed
set.seed(300)

# Specify means and correlation coefficient
n <- 6              # number of points
mu <- c(3, 1)       # mean vector (mu_x, mu_y)
sigma_x <- 3.5      # standard deviation x
sigma_y <- 2        # standard deviation y
rho <- -0.85          # correlation coefficient

# Define variance-covariance matrix
sigma <- matrix(
  c(sigma_x^2,
    rho*sigma_x*sigma_y,
    rho*sigma_x*sigma_y,
    sigma_y^2),
  nrow = 2)

# Simulate bivariate normal data
mvn_data_6 <- MASS::mvrnorm(n, mu, sigma) |> 
  data.frame() |> 
  rename(x = X1, y = X2) |> 
  round_df(1)
```

#### Table 3: Computational Practice

The table below contains values of two time series $\{x_t\}$ and $\{y_t\}$ observed at times $t = 1, 2, \ldots, 6$. We will use these values to practice finding the means, standard deviations, correlation coefficient, and covariance without using built-in R functions.

```{r}
#| echo: false

cov_dat <- mvn_data_6 |> 
  mutate(t = row_number()) |> 
  dplyr::select(t, x, y) |> 
  mutate(
    xx = x - mean(x),
    xx2 = xx^2,
    yy = y - mean(y),
    yy2 = yy^2,
    xy = (x - mean(x)) * (y - mean(y))
    ) 

cov_dat_summary <- cov_dat |> 
  summarize(
    x = sum(x),
    y = sum(y),
    xx = sum(xx),
    xx2 = sum(xx2),
    yy = sum(yy),
    yy2 = sum(yy2),
    xy = sum(xy)
  ) |> 
  round_df(5) |>
  mutate(across(everything(), as.character)) |> 
  mutate(t = "sum")

temp <- cov_dat |> 
  round_df(5) |>
  mutate(across(everything(), as.character)) |> 
  bind_rows(cov_dat_summary)

temp |> 
  blank_out_cells_in_df(ncols_to_keep = 3, nrows_to_keep = 1) |>
  bind_rows(temp |> tail(1) |> blank_out_cells_in_df(ncols_to_keep = 0, nrows_to_keep = 0) |> mutate(t = "$$~$$")) |> 
  rename(
    "$$t$$" = t,
    "$$x_t$$" = x,
    "$$y_t$$" = y,
    "$$x_t-\\bar x$$" = xx, 
    "$$(x_t - \\bar x)^2$$" = xx2, 
    "$$y_t-\\bar y$$" = yy,
    "$$(y_t-\\bar y)^2$$" = yy2, 
    "$$(x_t - \\bar x)(y_t-\\bar y)$$" = xy
  ) |>
  display_table()
```

Use the table above to determine these values:

::: columns
::: {.column width="30%"}

-   $\bar x =$

-   $\bar y =$

:::

::: {.column width="5%"}
<!-- empty column to create gap -->
:::

::: {.column width="30%"}

-   $s_x =$

-   $s_y =$

:::

::: {.column width="5%"}
<!-- empty column to create gap -->
:::

::: {.column width="30%"}

-   $r =$

-   $\\cov(x,y) =$

:::
:::

Here is a scatterplot of the data.

<center>

```{r fig.asp=0.75}
#| echo: false
#| warning: false

cov_dat <- cov_dat |> 
  mutate(
    sign = case_when(
      xy > 0 ~ "positive", 
      xy < 0 ~ "negative",
      TRUE ~ "zero")
  ) 
  
ggplotly(ggplot(cov_dat, aes(x = x, y = y)) +
  geom_point(size = 2) +
  labs(x="x", y="y") +
  theme_bw() +  
  ggtitle(paste0("Data for Computational Practice (n = ",n,")")) + 
  theme(plot.title = element_text(hjust = 0.5)))
```

</center>



### Summary

::: {.callout-tip icon=false title="Check Your Understanding"}

Working with your partner, prepare to explain the following concepts to the class:

-   Variance
-   Standard deviation
-   Correlation
-   Covariance

:::

## Computations in R (5 min)

Use these commands to load the data from the previous activity into R.
```{r}
#| echo: false
x <- mvn_data_6$x
y <- mvn_data_6$y
cat("x <- c(", paste(mvn_data_6$x, collapse = ", "),")")
cat("y <- c(", paste(mvn_data_6$y, collapse = ", "),")")
```

We can use R to compute the mean, variance, standard deviation, correlation coefficient, and covariance.

#### Mean, $\bar x$
```{r}
mean(x)
```


#### Variance, $s_x^2$
```{r}
var(x)
```


#### Standard Deviation, $s_x$
```{r}
sd(x)
```


#### Correlation Coefficient, $r$
```{r}
cor(x, y)
```


#### Covariance, $\\cov(x,y)$
```{r}
cov(x, y)
```




## Homework Preview (5 min)

-   Review upcoming homework assignment
-   Clarify questions



## Homework

::: {.callout-note icon=false}

## Download Homework

<a href="https://byuistats.github.io/timeseries/homework/homework_2_1.qmd" download="homework_2_1.qmd"> homework_2_1.qmd </a>

:::


<a href="javascript:showhide('Solutions')"
style="font-size:.8em;">Class Activity: Variance and Standard Deviation</a>

::: {#Solutions style="display:none;"}
Solutions to Class Activity: Variance and Standard Deviation

```{r}
#| echo: false
temp <- sim_data |> 
  mutate(
    xx = x - mean(x),
    xx2 = (x - mean(x))^2,
    ) 

temp2 <- temp |> 
  bind_rows(colSums(temp)) |>
  round_df(5) |>
  mutate(Solution = ifelse(row_number() == n(), "Sum", "")) |>
  dplyr::select(Solution, x, xx, xx2) |>
  data.frame()

ssx <- temp2[nrow(temp2), ncol(temp2)]

temp2 |> 
  rename(
    "$$x_t$$" = x,
    "$$x_t-\\bar x$$" = xx, 
    "$$(x_t-\\bar x)^2$$" = xx2
  ) |>
  display_table()
```



The variance of these values is $s^2 = \frac{`r ssx`}{`r nrow(temp)` - 1} = `r var(x)`$.

The standard deviation is $s = \sqrt{s^2} = \sqrt{`r var(x)`} = `r sd(x) |> round(3)`$.
:::

<a href="javascript:showhide('Solutions2')"
style="font-size:.8em;">Team Activity: Computational Practice</a>

::: {#Solutions2 style="display:none;"}
Solutions to Team Activity: Computational Practice

#### Table 3: Computational Practice

```{r}
#| echo: false

cov_dat <- mvn_data_6 |> 
  mutate(t = row_number()) |> 
  dplyr::select(t, x, y) |> 
  mutate(
    xx = x - mean(x),
    xx2 = xx^2,
    yy = y - mean(y),
    yy2 = yy^2,
    xy = (x - mean(x)) * (y - mean(y))
    ) 

cov_dat_summary <- cov_dat |> 
  summarize(
    x = sum(x),
    y = sum(y),
    xx = sum(xx),
    xx2 = sum(xx2),
    yy = sum(yy),
    yy2 = sum(yy2),
    xy = sum(xy)
  ) |> 
  round_df(5) |>
  mutate(across(everything(), as.character)) |> 
  mutate(t = "sum")

temp <- cov_dat |> 
  round_df(5) |>
  mutate(across(everything(), as.character)) |> 
  bind_rows(cov_dat_summary)

temp |> 
  # blank_out_cells_in_df(ncols_to_keep = 3, nrows_to_keep = 1) |>
  # bind_rows(temp |> tail(1) |> blank_out_cells_in_df(ncols_to_keep = 0, nrows_to_keep = 0) |> mutate(t = "$$~$$")) |> 
  rename(
    "$$t$$" = t,
    "$$x_t$$" = x,
    "$$y_t$$" = y,
    "$$x_t-\\bar x$$" = xx, 
    "$$(x_t - \\bar x)^2$$" = xx2, 
    "$$y_t-\\bar y$$" = yy,
    "$$(y_t-\\bar y)^2$$" = yy2, 
    "$$(x_t - \\bar x)(y_t-\\bar y)$$" = xy
  ) |>
  display_table()
```


::: columns
::: {.column width="30%"}

-   $\bar x = `r mean(mvn_data_6$x)`$

-   $\bar y = `r mean(mvn_data_6$y)`$

:::

::: {.column width="5%"}
<!-- empty column to create gap -->
:::

::: {.column width="30%"}

-   $s_x = `r sd(mvn_data_6$x)`$

-   $s_y = `r sd(mvn_data_6$y)`$

:::

::: {.column width="5%"}
<!-- empty column to create gap -->
:::

::: {.column width="30%"}

-   $r = `r cor(mvn_data_6$x, mvn_data_6$y)`$

-   $\\cov(x,y) = `r cov(mvn_data_6$x, mvn_data_6$y)`$

:::
:::


:::

