---
title: "Time Series Homework: Chapter 1 Lesson 1"
subtitle: "Please_put_your_name_here"
format: 
  html:
    embed-resources: true
    toc: true
editor: source
---

```{r}
#| message: false
#| echo: false

# Loading R packages
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, 
               tsibble, fable,
               feasts, tsibbledata,
               fable.prophet,
               patchwork,
               lubridate, 
               rio 
)
```

## Learning Outcomes

{{< include ../outcomes/chapter_1_lesson_2_outcomes.qmd >}}

## Questions

### Question 1 (5 points)

What is something unusual or unique about you?

::: {.callout-note icon="false" title="Answer"}
<!-- Delete this and put your answer here. The three colons mark the end of your answer. -->
:::

### Question 2 (5 points)

What do you hope to gain from this class?

::: {.callout-note icon="false" title="Answer"}
<!-- Delete this and put your answer here. The three colons mark the end of your answer. -->
:::

<!-- The following tip can help you complete Exercise 3 -->

::: {.callout-tip title="Tip for Exercise 3"}
In a qmd file, you can do all sorts of fun things. If you type `<ctrl> <alt> i`, RStudio will insert an R code chunk, where you can run any R code. The code below generates 10 uniform random variables between 0 and 1.

```{r}
#| message: false

# Load required R packages
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, lubridate)
               
# Set the random number seed
set.seed(100)

# Define the number of values to simulate
reps <- 4

# Create a data frame with a counter variable, t, 
# and the simulated values, called x
df1 <- data.frame(t = 1:reps, x = runif(reps, min = 0, max = 1))
df1
```

You can use inline R code as well. For example, the mean of the n=`r reps` simulated x values is: `r mean(df1$x)`.

If you do not know how to use any command such as `rnorm` in R, you can type the name of the command after a question mark (e.g. `?rnorm`) in the console area, and the help file will appear on the right.
:::

### Question 3 (10 points)

Use the `rnorm` command to simulate 100,000 normally-distributed random variables with a mean of 50 and a standard deviation of 10. Please do not list all the numbers. Instead, give a histogram of the data.\*

::: {.callout-note icon="false" title="Answer"}
<!-- Delete this and put your answer here. The three colons mark the end of your answer. -->
:::

### Question 4 (10 points)

Suppose your pulse was measured at the start of every hour today and the values are given in the data frame `pulse_df`.

```{r}
set.seed(123)
pulse_df <- data.frame(
  times = paste0(as.Date(substr(now(),1,10)), " ", c(0:23), ":00"),
  value = sample(70:100, size = 24, replace = TRUE)
)
```

We can convert a character representation of a date to a date-time object. The Lubridate package contains commands such as `mdy("12/31/2024")` which converts this value to: `r mdy("12/31/2024")`. There are other variations of this command such as `dmy_hms("31/12/2024 15:16:47")` which gives us: `r dmy_hms("31/12/2024 15:16:47")`.

Use the command `ymd_hm()` to convert the `times` variable into a date-type variable. Then filter the four observations from noon to 3 PM (15:00).

There are many ways to accomplish this but here is one simple example:

```{r message=FALSE}
pulse_df$times <- ymd_hm(pulse_df$times) # convert times to date-type data
pulse_df_filtered <- pulse_df %>% filter((times) >= 12 & hour(times) <= 15) # using the hour() function from lubridate we look at just the hour section of the 'times' variable and then filter to values between 12 and 15
```

Then, compute the mean of the observed value for these four times. Write your code in the R chunk below. Write an English sentence giving this mean and its interpretation in terms of your pulse rate.

::: {.callout-note icon="false" title="Answer"}
```{r}
# put your code here
pulse_df |>
  head() # lists the first 6 values - delete this sample command
```

<!-- Write an English sentence interpreting the mean, which includes an inline R chunk that gives its value. -->
:::

### Question 5 (5 points)

Do you have any concerns about your ability to succeed in this class? If so, please share them.

::: {.callout-note icon="false" title="Answer"}
<!-- Delete this and put your answer here. The three colons mark the end of your answer. -->
:::

### Rubric 

+----------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **Criteria**                           | **Mastery (5)**                                                                                                                                                                                                                                                                                                                                                          | **Incomplete (0)**                                                                                                                                                                                                                                                                                                     |
+----------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| #### **Question 1: Something Unusual** | Offers a solid and well-elaborated description of unique or unusual aspects, demonstrating a good level of self-awareness and the ability to express distinctive qualities effectively.                                                                                                                                                                                  | Gives a limited or unclear description of unique or unusual aspects, lacking essential details. Demonstrates a minimal level of self-awareness.                                                                                                                                                                        |
+----------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
|                                        | **Mastery (5)**                                                                                                                                                                                                                                                                                                                                                          | **Incomplete (0)**                                                                                                                                                                                                                                                                                                     |
+----------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| #### **Question 2: Class Goals**       | States clear and specific goals for the class, demonstrating a good level of understanding of what the individual aims to gain.                                                                                                                                                                                                                                          | States goals with limited clarity or specificity, lacking detail and depth. The understanding of what is hoped to be gained is minimal.                                                                                                                                                                                |
+----------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
|                                        | **Mastery (10)**                                                                                                                                                                                                                                                                                                                                                         | **Incomplete (0)**                                                                                                                                                                                                                                                                                                     |
+----------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| #### **Question 3: Data Generation**   | Utilizes the rnorm command accurately, generating 100,000 normally-distributed random variables with a mean of 50 and a standard deviation of 10. Demonstrates a precise understanding of the command and its parameters.                                                                                                                                                | Fails to use the rnorm command to generate normally-distributed random variables, leaving the question unanswered or without the required simulation.                                                                                                                                                                  |
+----------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
|                                        | **Mastery (10)**                                                                                                                                                                                                                                                                                                                                                         | **Incomplete (0)**                                                                                                                                                                                                                                                                                                     |
+----------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| #### **Question 4: Pulse**             | Successfully applies the ymd_hm() command to convert the times variable into a date-type variable. Accurately filters the four observations from noon to 3 PM (15:00). Accurately computes the mean of the observed values for the specified times and provides a clear English sentence interpreting the mean in terms of pulse rate. The interpretation is insightful. | Fails to provide any code for applying the ymd_hm() command and filtering, leaving the R chunk incomplete or without the required transformation. Attempts to compute the mean and provide an interpretation but with significant inaccuracies or lack of clarity. The interpretation may be incorrect or incomplete.  |
+----------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
|                                        | **Mastery (5)**                                                                                                                                                                                                                                                                                                                                                          | **Incomplete (0)**                                                                                                                                                                                                                                                                                                     |
+----------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| #### **Question 5: Concern**           | States concerns with clarity and provides a good level of detail in explaining potential challenges to success in the class. The response is clear and relevant.                                                                                                                                                                                                         | States concerns with limited clarity or detail, lacking in-depth explanations. Demonstrates a minimal understanding of potential challenges to success.                                                                                                                                                                |
+----------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **Total Points**                       | **35**                                                                                                                                                                                                                                                                                                                                                                   |                                                                                                                                                                                                                                                                                                                        |
+----------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
