---
title: "Time Series Homework: Chapter 2 Lesson 2"
subtitle: "Please_put_your_name_here"
format: 
  html:
    embed-resources: true
    toc: true
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}

# Loading R packages
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, 
               tsibble, fable,
               feasts, tsibbledata,
               fable.prophet,
               patchwork,
               lubridate, 
               rio,
               ggplot2,
               kableExtra
               )
```

## Data

```{r, message=FALSE}

manu_inv <- rio::import("https://byuistats.github.io/timeseries/data/manu_mat_invent.csv")

```

## Questions

### Question 1 - Context and Measurement (10 points)

The first part of any time series analysis is context. You cannot properly analyze data without knowing what the data is measuring. Without context, the most simple features of data can be obscure and inscrutable. This homework assignment will center around the series below.

Please research the time series. In the spaces below, give the data collection process, unit of analysis, and meaning of each observation for the series.

#### **Manufacturers' Materials and Supplies Inventories**

<https://fred.stlouisfed.org/series/UMTMMI>

::: {.callout-note title="Answer" icon="false"}
<!-- Delete this and put your answer here. The three colons mark the end of your answer. -->
:::

### Question 2 - Key Definitions (10 points)

Define the terms below and include any mathematical expressions or illustrations that may accompany the definition if available.

::: {.callout-note title="Answer" icon="false"}
-   **Ensemble of a time series**: <!-- Delete this and put your answer here. -->

-   **Sample estimate of the population mean of a time series**: <!-- Delete this and put your answer here. -->

-   **Constant variance estimator for a time series model**: <!-- Delete this and put your answer here. -->

-   **Stationarity**: <!-- Delete this and put your answer here. -->

-   **Lag k Autocovariance function**: <!-- Delete this and put your answer here. -->

-   **Lag k Autocorrelation function**: <!-- Delete this and put your answer here. -->
:::

### Question 3 - Manufacturer's Inventory: Autocorrelation and autocovariance (20 points)

#### a)  Please calculate the list of autocorrelation and autocovariance values for the Manufacturer's Inventory series.

::: {.callout-note title="Answer" icon="false"}
```{r}



```
:::

#### b)  If autocovariance and autocorrelation are trying to evaluate the temporal relationship of observations of our time series, explain why we get different values for the same lag of the autocorrelation and autocovariance estimates.

::: {.callout-note title="Answer" icon="false"}
<!-- Delete this and put your answer here. The three colons mark the end of your answer. -->
:::

### Question 4 - Manufacturer's Inventory: Stationary (40 points)

Weak stationarity is a form of stationarity important for the analysis of time series data. A time series is said to be weakly stationary if its statistical properties such as mean, variance, and autocovariance are constant over time. Here are the key components of weak stationarity:

*Constant Mean:* The mean of the time series remains constant over time. This doesn't necessarily mean that the time series is centered around zero; it just implies that the average value remains the same throughout the observed period.

*Constant Variance:* The variance of the time series is uniform across all time points. Like the mean, this doesn't imply that the variance must be zero, just that it doesn't change systematically with time.

*Constant Autocovariance:* The autocovariance between any two observations of the time series depends only on the time lag between them and not on the absolute positions of the observations in time. This implies that the dependence structure of the time series remains constant over time.

#### a)  Please split the time series into two parts, the earlier half of the data and the latter part of the data. Calculate the mean, variance, and autocovariance for each half. *Note:* *(it doesn't really matter if it's precisely half. An approximate middle is sufficient.)*

::: {.callout-note title="Answer" icon="false"}
```{r}



```
:::

#### b)  Is there evidence to suggest that the Manufacturer's Inventory series is weakly stationary?

::: {.callout-note title="Answer" icon="false"}
<!-- Delete this and put your answer here. The three colons mark the end of your answer. -->
:::

### Rubric

+-----------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **Criteria**                                        | **Mastery (10)**                                                                                                                                                                                                                                                                                                                                                                                                                        | **Incomplete (0)**                                                                                                                                                                                                                                                                                                                                                                                                                                                           |
+-----------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| #### **Question 1: Context and Measurement**        | The student thoroughly researches the data collection process, unit of analysis, and meaning of each observation for both the requested time series. Clear and comprehensive explanations are provided.                                                                                                                                                                                                                                 | The student does not adequately research or provide information on the data collection process, unit of analysis, and meaning of each observation for the specified series.                                                                                                                                                                                                                                                                                                  |
+-----------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
|                                                     | **Mastery (10)**                                                                                                                                                                                                                                                                                                                                                                                                                        | **Incomplete (0)**                                                                                                                                                                                                                                                                                                                                                                                                                                                           |
+-----------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| #### **Question 2: Definitions**                    | The student correctly defined each of the terms and included mathematical expressions or illustration if available in the text or the Time Series Notebook                                                                                                                                                                                                                                                                              | The student did not provide a correct definition for one or more of the terms.                                                                                                                                                                                                                                                                                                                                                                                               |
+-----------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
|                                                     | **Mastery (10)**                                                                                                                                                                                                                                                                                                                                                                                                                        | **Incomplete (0)**                                                                                                                                                                                                                                                                                                                                                                                                                                                           |
+-----------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| #### **Question 3: Autocorrelation and Covariance** | The student correctly computes the autocorrelation and autocovariance values for the Manufacturer's Inventory series using R.The R code is well-commented and structured, facilitating understanding of each step in the calculation process. Results are presented clearly.                                                                                                                                                            | The student attempts to compute autocorrelation and autocovariance values for the Manufacturer's Inventory series, but significant errors are present in the computations. The R code lacks clear documentation, with unclear or missing comments that hinder comprehension of the calculation process. Presentation of results may be confusing or incomplete, making it challenging to interpret the autocorrelation and autocovariance values accurately.                 |
+-----------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
|                                                     | **Mastery (10)**                                                                                                                                                                                                                                                                                                                                                                                                                        | **Incomplete (0)**                                                                                                                                                                                                                                                                                                                                                                                                                                                           |
+-----------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| #### **Question 3:Theoretical understanding**       | The student provides a clear and accurate explanation of why different values are obtained for the same lag of the autocorrelation and autocovariance estimates. The explanation demonstrates a solid understanding of the underlying concepts.                                                                                                                                                                                         | The student attempts to explain why different values are obtained for the same lag of the autocorrelation and autocovariance estimates but does so with significant inaccuracies or lack of clarity. The explanation may lack coherence or fail to address key differences between autocorrelation and autocovariance adequately.                                                                                                                                            |
+-----------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
|                                                     | **Mastery (20)**                                                                                                                                                                                                                                                                                                                                                                                                                        | **Incomplete (0)**                                                                                                                                                                                                                                                                                                                                                                                                                                                           |
+-----------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| #### **Question 4: Stationarity Calculations**      | The student accurately splits the dataset into two parts and calculates the mean, variance, and autocovariance for each part using R. The R code is well-commented, providing clear explanations of the steps taken to perform the analysis. The calculated statistics are presented clearly, aiding interpretation of the results, and the student shows a solid understanding of the concepts involved in analyzing time series data. | The student attempts to split the dataset into two parts and calculate the mean, variance, and autocovariance for each part using R, but does so with significant errors or inaccuracies. The R code lacks clear and sufficient commenting, making it difficult to understand the steps taken in the analysis. The calculated statistics may be presented poorly or inaccurately, indicating a limited understanding of the concepts involved in analyzing time series data. |
+-----------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
|                                                     | **Mastery (20)**                                                                                                                                                                                                                                                                                                                                                                                                                        | **Incomplete (0)**                                                                                                                                                                                                                                                                                                                                                                                                                                                           |
+-----------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| #### **Question 4: Evaluation**                     | The student accurately assesses whether there is evidence to suggest that the Manufacturer's Inventory series is weakly stationary. The analysis is supported by clear and concise explanations, demonstrating a solid understanding of the concept of weak stationarity.                                                                                                                                                               | The student attempts to assess whether the Manufacturer's Inventory series is weakly stationary but does so with significant errors or lacks clarity in their analysis. There may be inaccuracies in the methodology or misinterpretation of results, indicating a limited understanding of weak stationarity                                                                                                                                                                |
+-----------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **Total Points**                                    | **80**                                                                                                                                                                                                                                                                                                                                                                                                                                  |                                                                                                                                                                                                                                                                                                                                                                                                                                                                              |
+-----------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
