---
title: "Time Series Homework: Chapter 2 Lesson 1"
subtitle: "Please_put_your_name_here"
format: 
  html:
    embed-resources: true
    toc: true
    code-fold: true
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}

# Loading R packages
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, 
               tsibble, fable,
               feasts, tsibbledata,
               fable.prophet,
               patchwork,
               lubridate, 
               rio,
               ggplot2,
               kableExtra
               )
```

## Data

```{r, message=FALSE}

okuns <- rio::import("https://byuistats.github.io/timeseries/data/outputgap_and_cyclical_unemp.xlsx")

gs_night <- rio::import("https://byuistats.github.io/timeseries/data/nightstand-sweat.csv")

```

## Questions

### Question 1 - Context and Measurement (10 points)

The first part of any time series analysis is context. You cannot properly analyze data without knowing what the data is measuring. Without context, the most simple features of data can be obscure and inscrutable. This homework assignment will center around the series below.

Please research the time series. In the spaces below, give the data collection process, unit of analysis, and meaning of each observation for the series.

#### a) **Output Gap**

<https://chat.openai.com/share/122aaad9-2be6-43ec-b58a-e1858305b401>

::: {.callout-note title="Answer" icon="false"}
<!-- Delete this and put your answer here. The three colons mark the end of your answer. -->
:::

#### b) **Cyclical Unemployment**

<https://chat.openai.com/share/7d6bf187-41d0-42c3-98bc-d02ea1bd5b80>

::: {.callout-note title="Answer" icon="false"}
<!-- Delete this and put your answer here. The three colons mark the end of your answer. -->
:::

### Question 2 - Key Definitions (10 points)

Define the terms below and include any mathematical expressions or illustrations that may accompany the definition if available.

::: {.callout-note title="Answer" icon="false"}
-   **Sample mean**: <!-- Delete this and put your answer here. -->

-   **Sample variance**: <!-- Delete this and put your answer here. -->

-   **Sample standard deviation**: <!-- Delete this and put your answer here. -->

-   **Sample covariance**: <!-- Delete this and put your answer here. -->

-   **Sample correlation coefficient**: <!-- Delete this and put your answer here. -->
:::

### Question 3 - Covariance and Correlation: Okun's Law (45 points)

Okun's Law is an empirical relationship defined as a negative correlation between the Output Gap and Cyclical Unemployment. If the economy is expanding, businesses are producing more, and unemployment tends to decrease. Conversely, during economic contractions or recessions, output shrinks, leading to an increase in unemployment.

Please use the data ***okuns*** to test whether Okun's Law applies to the US from 1960 to 2021.

##### a)  Please create a scatter plot of the Output Gap in the x-axis and Cyclical Unemployment in the y-axis.

::: {.callout-note title="Answer" icon="false"}
```{r}

```
:::

##### b)  Please calculate the covariance and correlation coefficient between the Output Gap and Cyclical Unemployment.

::: {.callout-note title="Answer" icon="false"}
```{r}

```
:::

##### c)  Evaluate Okun's law. Is there sufficient evidence to suggest an empirical relationship between the Output Gap and Cyclical Unemployment ?

::: {.callout-note title="Answer" icon="false"}
<!-- Delete this and put your answer here. The three colons mark the end of your answer. -->
:::

### Question 4 - Correlation vs Causation: Spurious Searches (45 points)

::: quote
*"Every single person who confuses correlation and causation ends up dying."*

Hannah Fry
:::

Please use the data ***gs_night*** to evaluate the empirical relationship between the Google search terms for *night sweats* and *nightstand.*

##### a)  Please create a scatter plot of the night sweat search index in the x-axis and nightstand search index in the y-axis.

::: {.callout-note title="Answer" icon="false"}
```{r}

```
:::

##### b)  Please calculate the covariance and correlation between the night sweat search index and the nightstand search index.

::: {.callout-note title="Answer" icon="false"}
```{r}

```
:::

##### c)  Please suggest a scenario where an increase in the frequency of searches for nightstand will cause higher searches for the term night sweats?

::: {.callout-note title="Answer" icon="false"}
<!-- Delete this and put your answer here. The three colons mark the end of your answer. -->
:::

##### d)  Please suggest a scenario where an increase in the frequency of searches for night sweats will cause higher searches for the term nightstand?

::: {.callout-note title="Answer" icon="false"}
<!-- Delete this and put your answer here. The three colons mark the end of your answer. -->
:::

##### e)  Please suggest an event or reason that would result in the correlation between searches for night sweats and searches for the term nightstand?

::: {.callout-note title="Answer" icon="false"}
<!-- Delete this and put your answer here. The three colons mark the end of your answer. -->
:::

##### f)  Are any of your suggestions plausible? What does that teach you about the difference between correlation and causation? Please elaborate.

::: {.callout-note title="Answer" icon="false"}
<!-- Delete this and put your answer here. The three colons mark the end of your answer. -->
:::

### Rubric

+-------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **Criteria**                                    | **Mastery (10)**                                                                                                                                                                                                                                                                                                                                                 | **Incomplete (0)**                                                                                                                                                                                                                                                                                                                                                                                                                                      |
+-------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| #### **Question 1: Context and Measurement**    | The student thoroughly researches the data collection process, unit of analysis, and meaning of each observation for both the requested time series. Clear and comprehensive explanations are provided.                                                                                                                                                          | The student does not adequately research or provide information on the data collection process, unit of analysis, and meaning of each observation for the specified series.                                                                                                                                                                                                                                                                             |
+-------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
|                                                 | **Mastery (10)**                                                                                                                                                                                                                                                                                                                                                 | **Incomplete (0)**                                                                                                                                                                                                                                                                                                                                                                                                                                      |
+-------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| #### **Question 2: Definitions**                | The student correctly defined each of the terms and included mathematical expressions or illustration if available in the text or the Time Series Notebook                                                                                                                                                                                                       | The student did not provide a correct definition for one or more of the terms.                                                                                                                                                                                                                                                                                                                                                                          |
+-------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
|                                                 | **Mastery (15)**                                                                                                                                                                                                                                                                                                                                                 | **Incomplete (0)**                                                                                                                                                                                                                                                                                                                                                                                                                                      |
+-------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| #### **Question 3: Scatter Plot**               | The scatter plot is appropriately titled, and all elements, including the plot itself, axis labels, and title, are clearly labeled. The scatter plot matches or exceeds the quality of the scatter plots shown in lecture.                                                                                                                                       | The scatter plot doesn't meet professional standards                                                                                                                                                                                                                                                                                                                                                                                                    |
+-------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
|                                                 | **Mastery (10)**                                                                                                                                                                                                                                                                                                                                                 | **Incomplete (0)**                                                                                                                                                                                                                                                                                                                                                                                                                                      |
+-------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| #### **Question 3:Correlation and Covariance**  | The student effectively utilizes R to correctly calculate covariance and correlation.                                                                                                                                                                                                                                                                            | The student does not use R to describe key features of time series data or does not demonstrate proficiency in one or more specified R functions.                                                                                                                                                                                                                                                                                                       |
+-------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
|                                                 | **Mastery (20)**                                                                                                                                                                                                                                                                                                                                                 | **Incomplete (0)**                                                                                                                                                                                                                                                                                                                                                                                                                                      |
+-------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| #### **Question 3: Interpretation**             | The student provides a clear and accurate interpretation of the correlation between the variables. The response demonstrates understanding of how the correlation coefficient quantifies the strength and direction of the relationship between the variables. The student demonstrates mastery of the definitions and their application to the current context. | The student's interpretation of the correlation between the variables lacks clarity and accuracy. The response demonstrates confusion or misunderstanding of how the correlation coefficient quantifies the strength and direction of the relationship between the variables. The student's grasp of the definitions and their application to the current context appears limited, reflecting a need for improvement in comprehension and articulation. |
+-------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
|                                                 | **Mastery (15)**                                                                                                                                                                                                                                                                                                                                                 | **Incomplete (0)**                                                                                                                                                                                                                                                                                                                                                                                                                                      |
+-------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| #### **Question 4: Scatter Plot**               | The scatter plot is appropriately titled, and all elements, including the plot itself, axis labels, and title, are clearly labeled. The scatter plot matches or exceeds the quality of the scatter plots shown in lecture.                                                                                                                                       | The scatter plot doesn't meet professional standards                                                                                                                                                                                                                                                                                                                                                                                                    |
+-------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
|                                                 | **Mastery (10)**                                                                                                                                                                                                                                                                                                                                                 | **Incomplete (0)**                                                                                                                                                                                                                                                                                                                                                                                                                                      |
+-------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| #### **Question 4: Correlation and Covariance** | The student effectively utilizes R to correctly calculate covariance and correlation.                                                                                                                                                                                                                                                                            | The student does not use R to describe key features of time series data or does not demonstrate proficiency in one or more specified R functions.                                                                                                                                                                                                                                                                                                       |
+-------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
|                                                 | **Mastery (20)**                                                                                                                                                                                                                                                                                                                                                 | **Incomplete (0)**                                                                                                                                                                                                                                                                                                                                                                                                                                      |
+-------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| #### **Question 4: Plausiblity**                | The student provides a critical evaluation regarding the plausibility of the suggested scenarios and events. The evaluation includes a clear and comprehensive explanation on the difference between correlation and causation.                                                                                                                                  | The student's critical evaluation regarding the plausibility of the suggested scenarios and events lacks depth and clarity. The explanation on the difference between correlation and causation is unclear and lacks comprehensiveness. There is limited evidence of a clear understanding of the nuanced relationship between correlation and causation, indicating a need for improvement in both critical analysis and conceptual understanding.     |
+-------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **Total Points**                                | **110**                                                                                                                                                                                                                                                                                                                                                          |                                                                                                                                                                                                                                                                                                                                                                                                                                                         |
+-------------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
