---
title: "Time Series Homework: Chapter 1 Lesson 5"
subtitle: "Please_put_your_name_here"
format: 
  html:
    embed-resources: true
    toc: true
---

```{r, echo=FALSE, message=FALSE}

# Loading R packages
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, 
               tsibble, fable,
               feasts, tsibbledata,
               fable.prophet,
               patchwork,
               lubridate, 
               rio,
               ggplot2
               )
```

## Data

```{r}
vessels <- rio::import("https://byuistats.github.io/timeseries/data/Vessels_Trade_US.csv")
```

## Questions

### Question 1: Context and Measurement (10 points)

The first part of any time series analysis is context. You cannot properly analyze data without knowing what the data is measuring. Without context, the most simple features of data can be obscure and inscrutable. This homework assignment will center around the series below.

Please research the time series. In the spaces below, give the data collection process, unit of analysis, and meaning of each observation for the series.

#### a) **Vessels Cleared in Foreign Trade for United States**

<https://fred.stlouisfed.org/series/M03022USM583NNBR>

::: {.callout-note title="Answer" icon="false"}
<!-- Delete this and put your answer here. The three colons mark the end of your answer. -->
:::

### Question 2: Key Definitions (10 points)

Define the terms "sampling interval," "time series trend," "seasonal variation," "cycle," "deterministic trend," and "stochastic trend."

::: {.callout-note title="Answer" icon="false"}
-   **Multiplicative Decomposition**: 
<!-- Delete this and put your answer here. -->

-   **Estimators of seasonal variation for a multiplicative model**: 
<!-- Delete this and put your answer here. -->

:::

### Question 3: Seasonally Adjusted Series - Calculation and Plot (40 points)

Calculate the seasonally adjusted series for the Vessels Cleared in Foreign Trade series using the multiplicative decomposition model. Illustrate the original and the seasonally adjusted series in the same plot. Use the appropriate axis labels, units, and captions.

::: {.callout-note title="Answer" icon="false"}
<!-- Delete this and put your answer here. -->

:::

### Question 4: Seasonally Adjusted Series - Analysis (20 points)

Justify why we use the multiplicative decomposition model instead of the additive model to seasonally adjust the series.

::: {.callout-note title="Answer" icon="false"}
<!-- Delete this and put your answer here. -->

:::

## Rubric

+----------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **Criteria**                                 | **Mastery (10)**                                                                                                                                                                                                                         | **Incomplete (0)**                                                                                                                                                                                                                               |
+----------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| #### **Question 1: Context and Measurement** | The student thoroughly researches the data collection process, unit of analysis, and meaning of each observation for both the requested time series. Clear and comprehensive explanations are provided.                                  | The student does not adequately research or provide information on the data collection process, unit of analysis, and meaning of each observation for the specified series.                                                                      |
+----------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
|                                              | **Mastery (10)**                                                                                                                                                                                                                         | **Incomplete (0)**                                                                                                                                                                                                                               |
+----------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| #### **Question 2: Definitions**             | The student correctly defined each of the terms and included mathematical expressions or illustration if available in the text or the Time Series Notebook                                                                               | The student did not provide a correct definition for one or more of the terms.                                                                                                                                                                   |
+----------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
|                                              | **Mastery (20)**                                                                                                                                                                                                                         | **Incomplete (0)**                                                                                                                                                                                                                               |
+----------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| #### **Question 3: Code**                    | Correctly calculates the seasonally adjusted series for the Vessels Cleared in Foreign Trade series using the multiplicative decomposition model. The code is clear, well-commented, and easy to understand.                             | Attempts to calculate the seasonally adjusted series but with significant inaccuracies in the methodology. The code is unclear or poorly commented. Shows a limited understanding of the mathematical process.                                   |
+----------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
|                                              | **Mastery (20)**                                                                                                                                                                                                                         | **Incomplete (0)**                                                                                                                                                                                                                               |
+----------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| #### **Question 3: Plot**                    | Creates a clear and well-labeled plot illustrating both the original and seasonally adjusted series. Provides accurate axis labels, units, and a caption. The plot effectively communicates the results.                                 | Attempts to create a plot, but with significant inaccuracies or lack of clarity. The plot may not effectively communicate the results due to labeling or presentation issues.                                                                    |
+----------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
|                                              | **Mastery (20)**                                                                                                                                                                                                                         | **Incomplete (0)**                                                                                                                                                                                                                               |
+----------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| #### **Question 4: Analysis**                | Offers a solid explanation for choosing the multiplicative decomposition model, demonstrating a good understanding of the data characteristics and the rationale behind the selection. The explanation is clear and logically presented. | Attempts to justify the use of the multiplicative decomposition model but with significant inaccuracies or unclear reasoning. Shows a limited understanding of the data characteristics and the rationale for choosing the multiplicative model. |
+----------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **Total Points**                             | **80**                                                                                                                                                                                                                                   |                                                                                                                                                                                                                                                  |
+----------------------------------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
