---
title: "Holt-Winters Method (Multiplicative Models)"
subtitle: "Chapter 3: Lesson 5"
format: html
editor: source
sidebar: false
---

```{r}
#| include: false
source("common_functions.R")
```

```{=html}
<script type="text/javascript">
 function showhide(id) {
    var e = document.getElementById(id);
    e.style.display = (e.style.display == 'block') ? 'none' : 'block';
 }
 
 function openTab(evt, tabName) {
    var i, tabcontent, tablinks;
    tabcontent = document.getElementsByClassName("tabcontent");
    for (i = 0; i < tabcontent.length; i++) {
        tabcontent[i].style.display = "none";
    }
    tablinks = document.getElementsByClassName("tablinks");
    for (i = 0; i < tablinks.length; i++) {
        tablinks[i].className = tablinks[i].className.replace(" active", "");
    }
    document.getElementById(tabName).style.display = "block";
    evt.currentTarget.className += " active";
 }    
</script>
```

```{r}
#| echo: false

# Rounds the values at each step

hw_additive_slope_multiplicative_seasonal_rounded <- function(df, date_var, value_var, p = 12, predict_periods = 18, alpha = 0.2, beta = 0.2, gamma = 0.2, s_initial = rep(1,p)) {

  # Get expanded data frame
  df <- df |> expand_holt_winters_df_old(date_var, value_var, p, predict_periods) |>
    mutate(x_t = round(x_t, 1))

  # Fill in prior belief about s_t
  for (t in 1:p) {
    df$s_t[t] <- s_initial[t]
  }

  # Fill in first row of values
  offset <- p # number of header rows to skip
  df$a_t[1 + offset] <- df$x_t[1 + offset]
  df$b_t[1 + offset] <- round( (1 / p) * mean(df$x_t[(p + 1 + offset):(2 * p + offset)] - df$x_t[(1 + offset):(p + offset)]), 3)
  df$s_t[1 + offset] <- df$s_t[1]

  # Fill in remaining rows of body of df with values
  for (t in (2 + offset):(nrow(df) - predict_periods) ) {
    df$a_t[t] = round( alpha * (df$x_t[t] / df$s_t[t-p]) + (1 - alpha) * (df$a_t[t-1] + df$b_t[t-1]), 1)
    df$b_t[t] = round( beta * (df$a_t[t] - df$a_t[t-1]) + (1 - beta) * df$b_t[t-1], 3)
    df$s_t[t] = round( gamma * (df$x_t[t] / df$a_t[t]) + (1 - gamma) * df$s_t[t-p], 2)
  }

  df <- df |>
    mutate(k = ifelse(row_number() >= nrow(df) - predict_periods, row_number() - (nrow(df) - predict_periods), NA))

  # Fill in forecasted values
  offset <- nrow(df) - predict_periods
  for (t in (offset+1):nrow(df)) {
    df$s_t[t] = df$s_t[t - p]
    df$xhat_t[t] = round( (df$a_t[offset] + df$k[t] * df$b_t[offset]) * df$s_t[t - p], 1)
  }
  df$xhat_t[offset] = round( (df$a_t[offset] + df$k[offset] * df$b_t[offset]) * df$s_t[offset], 1) #### NOTE THIS ISSUE!!!

  # Delete temporary variable k
  df <- df |> select(-k)

  return(df)
}
```



## Learning Outcomes

{{< include outcomes/_chapter_3_lesson_5_outcomes.qmd >}}




## Preparation

-   Read Sections 3.4.2-3.4.3, 3.5



## Learning Journal Exchange (10 min)

-   Review another student's journal

-   What would you add to your learning journal after reading another student's?

-   What would you recommend the other student add to their learning journal?

-   Sign the Learning Journal review sheet for your peer




<!-- Great reference: -->
<!-- https://medium.com/analytics-vidhya/a-thorough-introduction-to-holt-winters-forecasting-c21810b8c0e6 -->



## Class Discussion: Multiplicative Seasonality (10 min)

We can assume either additive or multiplicative seasonality. In the previous two lessons, we explored additive seasonality. In this lesson, we consider the case where the seasonality is multiplicative. 

Additive seasonality is appropriate when the variation in the time series is roughly constant for any level. We assume multiplicative seasonality when the variation gets larger as the level increases.

### Forecasting 

Here are the forecasting equations we use, based on the model that is appropriate for the time series.



|                          | **Additive Seasonality**                                                               | **Multiplicative Seasonality**                                                              |
|------------------|----------------------------|----------------------------|
| **Additive Slope**       | $$ \hat x_{n+k \mid n} = \left( a_n + k \cdot b_n \right) + s_{n+k-p} $$                          | $$ \hat x_{n+k \mid n} = \left( a_n + k \cdot b_n \right) \cdot s_{n+k-p} $$                               |


<!-- Additional Row for the table:  -->
<!-- | **Multiplicative Slope** | $$ \hat x_{n+k \mid n} = \left[ a_n \cdot ( b_n )^k \right]+s_{n+k-p} $$ | $$ \hat x_{n+k \mid n} = \left[ a_n \cdot ( b_n )^k \right]\cdot s_{n+k-p} $$ | -->


<!-- Check Your Understanding -->

::: {.callout-tip icon=false title="Check Your Understanding"}

-   With your partner, for the forecasting equations above, identify where the additive or multiplicative terms are represented for both the slope and the seasonality.
    -   How is an additive slope represented in the forecasting equation?
    <!-- -   How is a multiplicative slope represented in the forecasting equation? -->
    -   How is additive seasonality represented in the forecasting equation?
    -   How is multiplicative seasonality represented in the forecasting equation?
:::


### Update Equations (Multiplicative Seasonals)

The update equations for the seasonals are:

\begin{align*}
  a_t &= \alpha \left( \frac{x_t}{s_{t-p}} \right) + (1-\alpha) \left( a_{t-1} + b_{t-1} \right) && \text{Level} \\
  b_t &= \beta \left( a_t - a_{t-1} \right) + (1-\beta) b_{t-1} && \text{Slope} \\
  s_t &= \gamma \left( \frac{x_t}{a_{t}} \right) + (1-\gamma) s_{t-p} && \text{Seasonal}
\end{align*}


Note that when the seasonal effect is additive, we subtract it from the time series to remove its effect. If the seasonal effect is multiplicative, we divide.


<!-- Check Your Understanding -->

::: {.callout-tip icon=false title="Check Your Understanding"}

Work with your partner to answer the following questions about the update equations.

$$
a_t = \alpha \cdot \underbrace{ \left( \frac{x_t}{s_{t-p}} \right) }_{A} + (1-\alpha) \cdot  \underbrace{ \left( a_{t-1} + b_{t-1} \right) }_{B}
~~~~~~~~~~~~~~~~~~~~ \text{Level}
$$


-   Interpret the term $A = \dfrac{x_t}{s_{t-p}}$.
-   Interpret the term $B = a_{t-1} - b_{t-1}$.
-   Explain why this expression for $a_t$ estimates the level of the time series at time $t$.

$$
b_t = \beta \cdot \underbrace{ \left( a_t - a_{t-1} \right) }_{C} + (1-\beta) \cdot \underbrace{ b_{t-1} }_{D}
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ \text{Slope}
$$

-   Interpret the term $C = a_t - a_{t-1}$.
-   Interpret the term $D = b_{t-1}$.
-   Explain why this expression for $b_t$ estimates the slope of the time series at time $t$.

$$
s_t = \gamma \cdot \underbrace{ \left( \frac{x_t}{a_{t}} \right) }_{E} + (1-\gamma) \cdot \underbrace{ s_{t-p} }_{F}
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~ \text{Seasonal}
$$

-   Interpret the term $E = \dfrac{x_t}{a_{t}}$.
-   Interpret the term $F = s_{t-p}$.
-   Explain why this expression for $s_t$ estimates the seasonal component of the time series at time $t$.
-   When the seasonal component appears on the right-hand side of the update equations, it always given as $s_{t-p}$. Why do we use the estimate of the seasonal effect $p$ periods ago? Why not apply a more recent value?
-   What do the following sets of terms have in common?
    -   $\{A, C, E \}$
    -   $\{B, D, F \}$
-   Explain why the Holt-Winters method for multiplicative seasonals works.

:::






## Small Group Activity: Holt-Winters Model for BYU-Idaho Enrollment Data (25 min)

We will now apply Holt-Winters filtering to the BYU-Idaho Enrollment data. First, we examine the time plot in @fig-enrollment-ts.

```{r}
#| label: fig-enrollment-ts
#| fig-cap: "Time plot of BYU-Idaho campus enrollments"
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false

# read in the data from a csv and make the tsibble
byui_enrollment_ts <- rio::import("https://byuistats.github.io/timeseries/data/byui_enrollment_2012.csv") |>
  rename(
    semester = "TermCode",
    year = "Year",
    enrollment = "On Campus Enrollment (Campus HC)"
  ) |>
  mutate(
    term = 
      case_when(
        left(semester, 2) == "WI" ~ 1,
        left(semester, 2) == "SP" ~ 2,
        left(semester, 2) == "FA" ~ 3,
        TRUE ~ NA
      )
  ) |>
  filter(!is.na(term)) |>
  mutate(dates = yearmonth( ym( paste(year, term * 4 - 3) ) ) ) |>
  mutate(enrollment_1000 = enrollment / 1000) |>
  dplyr::select(semester, dates, enrollment_1000) |>
  as_tsibble(index = dates) 

byui_enrollment_ts_expanded <- byui_enrollment_ts |>
  as_tibble() |>
  hw_additive_slope_multiplicative_seasonal_rounded("dates", "enrollment_1000", p = 3, predict_periods = 9) |>
  mutate(xhat_t = ifelse(t %in% c(1:36), round(a_t * s_t, 1), xhat_t)) |>
  select(-dates, -enrollment_1000)

# This is hard-coded for this data set, because I am in a hurry to get done.
# This revises the semester codes
byui_enrollment_ts_expanded$semester[1:3] <- c("SP11", "FA11", "WI12")
byui_enrollment_ts_expanded$semester[40:48] <- c("SP24", "FA24", "WI25",
                                                 "SP25", "FA25", "WI26",
                                                 "SP26", "FA26", "WI27")

byui_enrollment_ts_expanded |>
  tail(nrow(byui_enrollment_ts_expanded) - 3) |>
  ggplot(aes(x = date, y = x_t)) +
    geom_line(color = "black", linewidth = 1) +
    coord_cartesian(ylim = c(0, 22.5)) +
    labs(
      x = "Date",
      y = "Enrollment (in Thousands)",
      title = "BYU-Idaho On-Campus Enrollments (in Thousands)",
      color = "Components"
    ) +
    theme_minimal() +
    theme(legend.position = "none") +
    theme(plot.title = element_text(hjust = 0.5))
```

<!-- Check Your Understanding -->

::: {.callout-tip icon=false title="Check Your Understanding"}

-   Which of the Holt-Winter models described above is appropriate for this situation? Justify your answer.

:::


@tbl-enrollment-table summarizes the intermediate values for Holt-Winters filtering with multiplicative seasonals.

```{r}
#| label: tbl-enrollment-table
#| tbl-cap: "Holt-Winters smoothing for BYU-Idaho campus enrollments"
#| warning: false
#| echo: false

byui_enrollment_ts_expanded |>
  as_tibble() |>
  select(-date) |>
  rename(
    "$$Semester$$" = semester,
    "$$t$$" = t,
    "$$x_t$$" = x_t,
    "$$a_t$$" = a_t,
    "$$b_t$$" = b_t,
    "$$s_t$$" = s_t,
    "$$\\hat x_t$$" = xhat_t
  ) |>
  replace_cells_with_char(rows = 1:3, cols = 3:5, new_char = emdash) |>
  replace_cells_with_char(rows = 1:3, cols = 6, new_char = "") |>
  replace_cells_with_char(rows = 1:3, cols = 7, new_char = emdash) |>
  replace_cells_with_char(rows = 4, cols = 4:7, new_char = "") |>
  replace_cells_with_char(rows = 9:10, cols = 4:7, new_char = "") |>
  replace_cells_with_char(rows = 40:48, cols = 3:5, new_char = emdash) |>
  replace_cells_with_char(rows = 40:43, cols = 6:7, new_char = "") |>
  replace_na_with_char() |>
  head(nrow(byui_enrollment_ts_expanded) - 3) |>
  display_partial_table(14, 14, min_col_width = "0.75in")
```

<!-- Check Your Understanding -->

::: {.callout-tip icon=false title="Check Your Understanding"}

Apply Holt-Winters filtering to these data.

$$
\alpha = 0.2, \beta = 0.2, \gamma = 0.2
$$

-   Identify the value of $p$
-   Let $a_1 = x_1$
-   Compute 
$$
  b_1 =
    \frac{
      \left(
        \dfrac{x_{p+1} - x_{1}}{p} +
        \dfrac{x_{p+2} - x_{2}}{p} +
        \cdots +
        \dfrac{x_{2p} - x_{p}}{p}
      \right)
    }{p}
$$
-   Let $s_{1-p} = s_{2-p} = \cdots = s_{3-p} = 1$, and set $s_1 = s_{1-p}$.
-   Compute the values of $a_t$, $b_t$, and $s_t$ for all rows with observed time series data.
-   Find $\hat x_t = a_t \cdot s_t$ for all rows with data. 
-   Compute the prediction $\hat x_{n+k|n} = \left( a_t + k \cdot b_n \right) \cdot s_{n+k-p}$ for the future values.
-   Superimpose a sketch of your Holt-Winters filter and the associated forecast on @fig-enrollment-ts.

:::



## Small Group Activity: Applying Holt-Winters in R to the Apple Quarterly Revenue Data (10 min)

Recall the Apple, Inc., revenue values reported by Bloomberg:

```{r}
#| code-fold: true
#| code-summary: "Show the code"

apple_ts <- rio::import("https://byuistats.github.io/timeseries/data/apple_revenue.csv") |>
  mutate(
    dates = mdy(date),
    year = lubridate::year(dates),
    quarter = lubridate::quarter(dates),
    value = revenue_billions
  ) |>
  dplyr::select(dates, year, quarter, value)  |> 
  arrange(dates) |>
  mutate(index = tsibble::yearquarter(dates)) |>
  as_tsibble(index = index) |>
  dplyr::select(index, dates, year, quarter, value) |>
  rename(revenue = value) # rename value to emphasize data context

apple_ts |>
  autoplot(.vars = revenue) +
  labs(
    x = "Quarter",
    y = "Apple Revenue, Billions $US",
    title = "Apple's Quarterly Revenue, Billions of U.S. Dollars"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

Here are a few rows of the summarized data.

```{r}
#| echo: false

# View data
apple_ts |>
  display_partial_table(6,3) 
```


<!-- Check Your Understanding -->

::: {.callout-tip icon=false title="Check Your Understanding"}

-   What do you notice about this time plot?
    - Describe the trend
    - Is there evidence of seasonality?
    - Is the additive or multiplicative model appropriate?

:::

We apply Holt-Winters filtering to the quarterly Apple revenue data with a multiplicative model:


```{r}
#| code-fold: true
#| code-summary: "Show the code"

apple_hw <- apple_ts |>
  # tsibble::fill_gaps() |>
  model(Additive = ETS(revenue ~
        trend("A") +
        error("A") +
        season("M"),
        opt_crit = "amse", nmse = 1))
report(apple_hw)
```

We can compute some values to assess the fit of the model:
```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| eval: false

# SS of random terms
sum(components(apple_hw)$remainder^2, na.rm = T)

# RMSE
forecast::accuracy(apple_hw)$RMSE

# Standard devation of the quarterly revenues
sd(apple_ts$revenue)
```

-   The sum of the square of the random terms is: `r sum(components(apple_hw)$remainder^2, na.rm = T)`.
-   The root mean square error (RMSE) is: `r forecast::accuracy(apple_hw)$RMSE`.
-   The standard deviation of the number of incidents each month is `r sd(apple_ts$revenue)`.

@fig-apple-decomp illustrates the Holt-Winters decomposition of the Apple revenue data.

```{r}
#| label: fig-apple-decomp
#| fig-cap: "Apple, Inc., Quarterly Revenue (in Billions)"
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false

autoplot(components(apple_hw))
```

In @fig-apple-hw, we can observe the relationship between the Holt-Winters filter and the Apple revenue time series.

```{r}
#| label: fig-apple-hw
#| fig-cap: "Superimposed plots of the Apple revenue and the Holt-Winters filter"
#| code-fold: true
#| code-summary: "Show the code"

augment(apple_hw) |>
  ggplot(aes(x = index, y = revenue)) +
    geom_line() +
    geom_line(aes(y = .fitted, color = "Fitted")) +
    labs(color = "")
```

@fig-apple-hw-forecast contains the information from @fig-apple-hw, with the addition of an additional four years of forecasted values. The light blue bands give a 95% prediction bands for the forecast.

```{r}
#| label: fig-apple-hw-forecast
#| fig-cap: "Superimposed plots of Apple's quarterly revenue and the Holt-Winters filter, with four additional years forecasted"
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false

apple_forecast <- apple_hw |>
  forecast(h = "4 years") 

apple_forecast |>
  autoplot(apple_ts, level = 95) +
  geom_line(aes(y = .fitted, color = "Fitted"),
    data = augment(apple_hw)) +
  scale_color_discrete(name = "")
```



## Homework Preview (5 min)

-   Review upcoming homework assignment
-   Clarify questions



::: {.callout-note icon=false}

## Download Homework

<a href="https://byuistats.github.io/timeseries/homework/homework_3_5.qmd" download="homework_3_5.qmd"> homework_3_5.qmd </a>

:::




<a href="javascript:showhide('Solutions1')"
style="font-size:.8em;">BYU-Idaho Enrollment</a>

::: {#Solutions1 style="display:none;"}


```{r}
#| echo: false
#| warning: false

## Holt-Winters Multiplicative Model - Plot
byui_enrollment_ts |> 
  as_tibble() |>
  hw_additive_slope_multiplicative_seasonal_rounded("dates", "enrollment_1000", p = 3, predict_periods = 9) |>
  as_tsibble(index = date) |>
  tail(-3) |>
  ggplot(aes(x = date)) +
    geom_line(aes(y = x_t), color = "black", linewidth = 1) +
    geom_line(aes(y = a_t * s_t, color = "Combined", alpha=0.5), linewidth = 1) +
    geom_line(aes(y = xhat_t, color = "Combined", alpha=0.5), linetype = "dashed", linewidth = 1) +
    coord_cartesian(ylim = c(12, 22.5)) +
    labs(
      x = "Date",
      y = "Enrollment (in Thousands)",
      title = "BYU-Idaho Enrollments with Holt-Winters Forecast",
      color = "Components"
    ) +
    theme_minimal() +
    theme(legend.position = "none") +
    theme(
      plot.title = element_text(hjust = 0.5)
    )
```


#### Table 1: Holt-Winters smoothing for BYU-Idaho campus enrollments

```{r}
#| warning: false
#| echo: false

# This is hard-coded for this data set, because I am in a hurry to get done.
# Revise the semester codes
byui_enrollment_ts_expanded$semester[1:3] <- c("SP11", "FA11", "WI12")
byui_enrollment_ts_expanded$semester[1:3] <- c("SP11", "FA11", "WI12")
byui_enrollment_ts_expanded$semester[40:48] <- c("SP24", "FA24", "WI25",
                                                 "SP25", "FA25", "WI26",
                                                 "SP26", "FA26", "WI27")
# ,
#                                                  "SP27", "FA27", "WI28")

byui_enrollment_ts_expanded |>
  as_tibble() |>
  select(-date) |>
  rename(
    "$$Semester$$" = semester,
    "$$t$$" = t,
    "$$x_t$$" = x_t,
    "$$a_t$$" = a_t,
    "$$b_t$$" = b_t,
    "$$s_t$$" = s_t,
    "$$\\hat x_t$$" = xhat_t
  ) |>
  replace_cells_with_char(rows = 1:3, cols = 3:5, new_char = emdash) |>
  # replace_cells_with_char(rows = 1:3, cols = 6, new_char = "") |>
  replace_cells_with_char(rows = 1:3, cols = 7, new_char = emdash) |>
  # replace_cells_with_char(rows = 4, cols = 4:7, new_char = "") |>
  # replace_cells_with_char(rows = 9:10, cols = 4:7, new_char = "") |>
  replace_cells_with_char(rows = 40:48, cols = 3:5, new_char = emdash) |>
  # replace_cells_with_char(rows = 40:43, cols = 6:7, new_char = "") |>
  replace_na_with_char() |>
  display_partial_table(14, 14, min_col_width = "0.75in")
```

:::




## References

-   C. C. Holt (1957) Forecasting seasonals and trends by exponentially weighted moving averages, ONR Research Memorandum, Carnegie Institute of Technology 52. (Reprint at [https://doi.org/10.1016/j.ijforecast.2003.09.015](https://doi.org/10.1016/j.ijforecast.2003.09.015)).
-   P. R. Winters (1960). Forecasting sales by exponentially weighted moving averages. Management Science, 6, 324--342. (Reprint at [https://doi.org/10.1287/mnsc.6.3.324](https://doi.org/10.1287/mnsc.6.3.324).)