---
title: "Exploration of Autocorrelation Concepts"
subtitle: "Chapter 2: Lesson 3"
format: html
editor: source
sidebar: false
---

```{r}
#| include: false
source("common_functions.R")
library(tidyverse)
```

```{=html}
<script type="text/javascript">
 function showhide(id) {
    var e = document.getElementById(id);
    e.style.display = (e.style.display == 'block') ? 'none' : 'block';
 }
 
 function openTab(evt, tabName) {
    var i, tabcontent, tablinks;
    tabcontent = document.getElementsByClassName("tabcontent");
    for (i = 0; i < tabcontent.length; i++) {
        tabcontent[i].style.display = "none";
    }
    tablinks = document.getElementsByClassName("tablinks");
    for (i = 0; i < tablinks.length; i++) {
        tablinks[i].className = tablinks[i].className.replace(" active", "");
    }
    document.getElementById(tabName).style.display = "block";
    evt.currentTarget.className += " active";
 }    
</script>
```

```{r}
#| include: false

get_data_for_cov_table <- function(offset = 1) {
  x1 <- get_toy_data()
  
  # build a data frame
  df <- data.frame(t = 1:length(x1),
                   x = x1,
                   y = lead(x1, offset)) |>
    mutate(
      xx = x - mean(x),
      xx2 = xx^2,
      yy = y - mean(x),
      # yy2 = yy^2,
      xy = xx * yy
    ) |> 
    dplyr::select(t, x, y, xx, xx2, yy, xy)
  
  return(df)
}

make_cov_table_df <- function(df, offset=1, decimals_1st_order = 5, decimals_2nd_order = 5) {
  # Color vector
  oi_colors <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#F5C710", "#CC79A7", "#999999")
  
  df_summary <- df |> 
    summarize(
      x = sum(x),
      y = sum(y, na.rm = TRUE),
      xx = sum(xx),
      yy = sum(yy, na.rm = TRUE),
      xx2 = sum(xx2),
      # yy2 = sum(yy2, na.rm = TRUE),
      xy = sum(xy, na.rm = TRUE)
    ) |> 
    # round_df(3) |> 
    mutate(t = paste0("sum")) |> 
    mutate(
      t = paste0("sum"),
      x = round_to_places(x, 1),
      y = round_to_places(y, 1),
      xx = round_to_places(xx, decimals_1st_order),
      xx2 = round_to_places(xx2, decimals_2nd_order),
      yy = round_to_places(yy, decimals_1st_order),
      # yy2 = round_to_places(yy2, decimals_2nd_order),
      xy = round_to_places(xy, decimals_2nd_order)
    ) |> 
    dplyr::select(t, x, y, xx, xx2, yy, xy)
  
  out <- df |>
    mutate(
      t = as.character(t),
      x = round_to_places(x, 1),
      y = round_to_places(y, 1),
      xx = round_to_places(xx, decimals_1st_order),
      xx2 = round_to_places(xx2, decimals_2nd_order),
      yy = round_to_places(yy, decimals_1st_order),
      # yy2 = round_to_places(yy2, decimals_2nd_order),
      xy = round_to_places(xy, decimals_2nd_order)
    ) |> 
    mutate(
      x = cell_spec(x, 
                    color = case_when(
                      is.na(x) ~ "#999999",
                      TRUE ~ oi_colors[( row_number() + 0 ) %% 9 + 1]
                    )
      ),
      y = cell_spec(y, 
                    color = case_when(
                      is.na(y) ~ "#999999",
                      TRUE ~ oi_colors[( row_number() + offset ) %% 9 + 1]
                    )
      )
    ) |> 
    mutate(
      # x = ifelse(row_number() > nrow(.) - offset, paste0("[",x,"]"), x),
      y = ifelse(row_number() > nrow(.) - offset, NA, y),
    ) |> 
    replace(., is.na(.), "—") |>
    bind_rows(df_summary) |> 
    rename(
      "x_t" = x,
      "x_{t+k}" = y,
      # paste0("x_{t+", offset, "}") = y,
      "x_t-mean(x)" = xx, 
      "(x_t-mean(x))^2" = xx2, 
      "x_{t+k}-mean(x)" = yy,
      # "(x_{t+k}-mean(x))^2" = yy2, 
      "(x-mean(x))(x_{t+k}-mean(x))" = xy
    )
  
  return(out)
}

# Compute summary values
compute_summaries <- function(df, digits = 4) {
  df |> 
    summarize(
      mean_x = mean(x),
      mean_y = mean(y, na.rm = TRUE),
      ss_x = sum(xx2),
      # ss_y = sum(yy2, na.rm = TRUE),
      ss_xy = sum(xy, na.rm = TRUE),
      c_0 = sum(xx2) / nrow(.),
      c_k = sum(xy, na.rm = TRUE) / nrow(.),
      r_k = c_k / c_0
    ) |> 
    round_df(digits)
}

```


## Learning Outcomes

{{< include outcomes/_chapter_2_lesson_3_outcomes.qmd >}}





## Preparation

-   Read Sections 2.2.5 and 2.3-2.5 (No new reading assignment)



## Learning Journal Exchange (10 min)

-   Review another student's journal
-   What would you add to your learning journal after reading your partner's?
-   What would you recommend your partner add to their learning journal?
-   Sign the Learning Journal review sheet for your peer


## Correlograms (10 min)

In the previous lesson, we used the following time series as an example.
Here are the values in that time series:

```{r}
#| echo: false

# simulate correlated normal random data
x1 <- get_toy_data()

cat("x <- c(",paste(x1, collapse = ", "),")")
```

-   The table below gives the sample autocorrelation function, acf, for this data set. You may recognize some of these values from the previous lesson. 

```{r}
#| echo: false

df <- data.frame(x = x1)

z <- acf(df$x, plot=FALSE, type = "correlation") 
# acf(df$x, plot=TRUE, type = "correlation") ## Solution

z$acf |> 
  data.frame() |> 
  round_df(3) |> 
  rename("acf" = "z.acf") |> 
  mutate(
    k = row_number() - 1,
    acf = as.character(acf)
  ) |> 
  pivot_wider(names_from = k, values_from = acf) |> 
  # mutate("4" = "_____") |> 
  display_table()
```

::: {.callout-tip icon=false title="Check Your Understanding"}

-   Use the acf values to sketch the correlogram for these data in your Learning Journal. The figure below can help you begin.

```{r}
#| echo: false

ggplot(data = df, aes(x = seq_along(x), y = acf(x, plot = FALSE)$acf)) +
  # geom_col() +
  ylim(-1, 1) +
  scale_x_continuous(breaks = 0:9) + 
  geom_segment(aes(x = 0, y = 0, xend = 0, yend = 1)) + 
  geom_segment(aes(x = 0, y = 0, xend = 9, yend = 0)) + ## Hack
  geom_hline(yintercept = 0, linetype = "solid", linewidth=1, color = "black") +
  geom_hline(yintercept = (2.6/4.2), linetype = "dashed", linewidth=1, color = "#0072B2") +  # Texbooks says these lines should be at (-0.1 +/- 2/sqrt(10)). Used +/-(2.6/4.2), based on measurements made visually with a ruler from the figure generated by R.
  geom_hline(yintercept = (-2.6/4.2), linetype = "dashed", linewidth=1, color = "#0072B2") +
  labs(x = "Lag", y = "ACF") +
  # theme_bw()   
  # theme(panel.grid.major.x = element_blank(), panel.grid.major.y = element_blank())
  theme_bw() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()
  )
```

<!-- -   Finish a partially-created correlogram with the r_k values computed -->

-   Are any of the autocorrelations statistically significant? If so, which one(s)?

:::

## Application: Chocolate Search Trends (10 min)

Recall the Google Trends data for the term "chocolate" from the last lesson.
The cleaned data are available in the file <a href="data/chocolate.csv" download>chocolate.csv</a>.

### Import the chocolate search data and convert to tsibble format

Use the code below to import the data and convert it into a time series (tsibble) object.

```{r}
#| warning: false

# load packages
if (!require("pacman")) install.packages("pacman")
pacman::p_load("tsibble", "fable",
               "feasts", "tsibbledata",
               "fable.prophet", "tidyverse",
               "patchwork", "rio")

# read in the data from a csv and make the tsibble
# change the line below to include your file path
chocolate_month_ts <- rio::import("https://byuistats.github.io/timeseries/data/chocolate.csv") |>
  mutate(
    dates = yearmonth(ym(Month)),
    month = month(dates),
    year = year(dates),
    value = chocolate
  ) |> 
  dplyr::select(dates, month, year, value) |>
  as_tsibble(index = dates)

choc_decompose <- chocolate_month_ts |>
    model(feasts::classical_decomposition(value,
        type = "add"))  |>
    components()

autoplot(choc_decompose)
```

Here are the values of the acf for the chocolate search data:

```{r}
acf(chocolate_month_ts$value, plot=FALSE, type = "correlation", lag.max = 25)
```

Here is the associated correlogram:

```{r}
acf(chocolate_month_ts$value, plot=TRUE, type = "correlation", lag.max = 25)
```

<!-- Check Your Understanding -->

::: {.callout-tip icon=false title="Check Your Understanding"}

-   What does the information displayed in this correlogram suggest?

:::

If we consider only the random component of this time series, the correlogram is:

```{r}
acf(choc_decompose$random |> na.omit(), plot=TRUE, type = "correlation", lag.max = 25)
```

::: {.callout-tip icon=false title="Check Your Understanding"}

-   What do the spikes in the correlogram tell us about this time series?
-   Is there evidence of autocorrelation in the data after removing the trend and seasonal variation?

:::


## Small Group Activity: BYU-Idaho On-Campus Enrollment (25 min)

The official number of on-campus BYU-Idaho students each semester is given in the file <a href="https://byuistats.github.io/timeseries/data/byui_enrollment.csv" download>byui_enrollment.csv</a>. 



::: {.callout-tip icon=false title="Check Your Understanding"}

Do the following:

-   Create a tsibble with the BYU-Idaho enrollment data. (Hint: There are three semesters in a year, so treat the enrollments as observations taken every four months in January, May, and September.)
-   Plot the decomposition of this time series.
-   Describe the trend.
-   Describe the seasonal component.
-   Is there evidence of seasonal variation? If so, propose an explanation for the seasonal variation.
-   Create the correlogram for these data. 
    -   What do you observe?
    -   Does the correlogram support the statement you made about the seasonal component?
-   Is there evidence of autocorrelation in the data after removing the trend and seasonal variation?

:::





## Homework Preview (5 min)

-   Review upcoming homework assignment
-   Clarify questions



## Homework

::: {.callout-note icon=false}

## Download Homework

<a href="https://byuistats.github.io/timeseries/homework/homework_2_3.qmd" download="homework_2_3.qmd"> homework_2_3.qmd </a>

:::

<a href="javascript:showhide('Solutions')"
style="font-size:.8em;">Correlograms</a>

::: {#Solutions style="display:none;"}

Solutions to correlogram activity

```{r}
x <- c( 4.4, 4.2, 4.2, 4, 4.4, 4.7, 4.9, 5.3, 5.4, 5.5 )
acf(x, plot=FALSE, type = "correlation")
acf(x, plot=TRUE, type = "correlation")
```

:::


<a href="javascript:showhide('Solutions2')"
style="font-size:.8em;">BYU-Idaho Enrollment</a>

::: {#Solutions2 style="display:none;"}

Solutions to BYU-Idaho Enrollment Activity


```{r}
#| warning: false

# read in the data from a csv and make the tsibble

# Method 1:
enrollment_df <- rio::import("https://byuistats.github.io/timeseries/data/byui_enrollment.csv")
start_date <- lubridate::ymd("2019-05-01")
date_seq <- seq(start_date,
                start_date + months(nrow(enrollment_df)-1) * 4,
                by = "4 months")
enrollment_ts <- tibble(
    dates = tsibble::yearmonth(date_seq),
    semester = pull(enrollment_df, semester),
    enrollment = pull(enrollment_df, enrollment)
  ) |>
  dplyr::select(semester, dates, enrollment) |>
  as_tsibble(index = dates)

# Method 2:
enrollment_ts <- rio::import("https://byuistats.github.io/timeseries/data/byui_enrollment.csv") |>
  mutate(
    dates = yearmonth(ym(paste(year, term * 4 - 3)))
  ) |>
  dplyr::select(semester, dates, enrollment) |>
  as_tsibble(index = dates) 

# Compute and plot the decomposition
enrollment_decompose <- enrollment_ts |>
    model(feasts::classical_decomposition(enrollment,
        type = "add"))  |>
    components()
autoplot(enrollment_decompose)
```


```{r}
acf(enrollment_decompose$enrollment, type = "correlation")
```


```{r}
acf(enrollment_decompose$enrollment, plot=FALSE, type = "correlation")
```


```{r}
acf(enrollment_decompose$enrollment, plot=TRUE, type = "correlation")
```


```{r}
acf(enrollment_decompose$random |> na.omit(), plot=TRUE, type = "correlation")
```

:::
