---
title: "Plots Trends, and Seasonal Variation"
subtitle: "Chapter 1: Lesson 2"
format: html
editor: source
sidebar: false
filters:
  - shinylive
---

```{r}
# install.packages("pak")
#pak::pak("posit-dev/r-shinylive")
```

```{shinylive-r}
# #| standalone: true
# #| viewerHeight: 500
#library(shiny)
#library(shinylive)
#library(see)

```



```{r}
#| include: false
source("common_functions.R")
```

```{=html}
<script type="text/javascript">
 function showhide(id) {
    var e = document.getElementById(id);
    e.style.display = (e.style.display == 'block') ? 'none' : 'block';
 }
 
 function openTab(evt, tabName) {
    var i, tabcontent, tablinks;
    tabcontent = document.getElementsByClassName("tabcontent");
    for (i = 0; i < tabcontent.length; i++) {
        tabcontent[i].style.display = "none";
    }
    tablinks = document.getElementsByClassName("tablinks");
    for (i = 0; i < tablinks.length; i++) {
        tablinks[i].className = tablinks[i].className.replace(" active", "");
    }
    document.getElementById(tabName).style.display = "block";
    evt.currentTarget.className += " active";
 }    
</script>
```

## Learning Outcomes

<details>
<summary>Use technical language to describe the main features of time series data</summary>
    -   Define time series analysis
    -   Define time series
    -   Define sampling interval
    -   Define serial dependence or autocorrelation
    -   Define a time series trend
    -   Define seasonal variation
    -   Define cycle
    -   Differentiate between deterministic and stochastic trends
</details>

<details>
<summary>Use R to describe key features of time series data</summary>
    -   Create a seasonal summmary of a time series
</details>
<details>
<summary>Plot time series data to visualize trends, seasonal patterns, and potential outliers</summary>
    -   Plot a "ts" object
    -   Combine multiple "ts" objects into a single plot ????????????????????????????????????????????????????????????????????????-TIED-TO-CORRELATION-IS-NOT-CAUSATION-?????????????????? When?
    -   Make boxplots to examine distribution of values within each season
</details>


## Preparation

-   Read Sections 1.1-1.4 and 1.5.1-1.5.3
 
## Learning Journal Exchange (15 min)

-   Review another student's journal
-   What would you add to your learning journal after reading your partner's?
-   What would you recommend your partner add to their learning journal?
-   Sign the Learning Journal review sheet for your peer


## Vocabulary and Momenclature Matching Activity (15 min)

::: {.callout-tip icon=false title="Check Your Understanding"}

Working with a partner, match the definitions on the left with the terms on the right.

<!-- https://bookdown.org/yihui/rmarkdown-cookbook/multi-column.html -->

:::: {style="display: flex;"}
<div>

1.  A figure with time on the horizontal axis and the value of a random variable on the vertical axis
2.  A systematic change in a time series that does not appear to be periodic
3.  Repeated pattern within each year (or any other time period)
4.  Repeated pattern that does not correspond to some fixed natural period
5.  Observations in which values are related to lagged observations of the same
6.  Random trend that does not follow a discernible or predictable pattern
7.  Can be modeled with mathematical functions, facilitating the long-term prediction of the behavior
8.  Number of observations of a time series
9.  Discrete observations of a time series, taken at times $1, 2, \ldots, n$.
10. Lead time
11. Forecast made at time $t$ for a future value $k$ time units in the future
12. Additive decomposition
13. Additive decomposition model after taking the logarithm
14. Multiplicative model
15. Centered moving average
16. Estimate of monthly additive effect
17. Estimate of monthly multipliciative effect
18. Seasonally adjusted mean for the month corresponding to time $t$
19. Seasonal adjusted series (additive seasonal effect)
20. Seasonal adjusted series (additive seasonal effect)
21. The trend as observed at time $t$
22. The seasonal effect, as observed at time $t$
23. The error term (a sequence of correlated random variables with mean zero), as observed at time $t$
</div>

<div>
A.  Cycles
B.  Correlated (Serially Dependent) Data
C.  Deterministic Trend
D.  Seasonal Variations
E.  Stochastic Trend
F.  Time Plot
G.  Trends
H.  $n$
I.  $\{x_t\}$
J.  $k$
K.  $\hat x_{t+k \mid t}$
L.  $m_t$
M.  $\hat m_t$
N.  $s_t$
O.  $\hat s_t = x_t - \hat m_t$
P.  $\hat s_t = \dfrac{x_t}{\hat m_t}$
Q.  $\bar s_t$
R.  $z_t$
S.  $x_t = m_t + s_t + z_t$
T.  $x_t = m_t \cdot s_t + z_t$
U.  $\log(x_t) = m_t + s_t + z_t$
V.  $x_t - \bar s_t$
W.  $\frac{x_t}{\bar s_t}$
</div>

::::

where $\hat m_t = \dfrac{\frac{1}{2}x_{t-6} + x_{t-5} + \cdots + x_{t-1} + x_t + x_{t+1} + \cdot + x_{t+5} + \frac{1}{2} x_{t+6}}{12}$.

:::

## Comparison of Deterministic and Stochastic Time Series (5 min)

### Stochastic Time Series

The following app illustrates a few realizations of a stochastic time series. 

```{=html}
 <iframe id="stochastic" src="https://posit.byui.edu/content/728a7ddc-97d2-44ee-9977-fc648102b986" style="border: none; width: 100%; height: 720px" frameborder="0"></iframe>
```


-   If a stochastic time series displays an upward trend, can we conclude that trend will continue in the same direction? Why or why not?


### Deterministic Time Series

The figure below illustrates realizations of a deterministic time series. The data fluctuate around a sine curve.


```{=html}
 <iframe id="deterministic" src="https://posit.byui.edu/content/4d25268b-d562-455c-ac47-5a9871770e7c" style="border: none; width: 100%; height: 720px" frameborder="0"></iframe>
```


## Hands-on Exercise: Exploring a time series in R (Google Trends: Chocolate) (20 min)

Recall the Google Trends data for the term "chocolate" from the last lesson.
The cleaned data are available in the file <a href="data/chocolate.csv" download>chocolate.csv</a>.

### Import the chocolate search data and convert to tsibble format

Use the code below to import the data and convert it into a time series (tsibble) object.

```{r}

# load packages
if (!require("pacman")) install.packages("pacman")
pacman::p_load("tsibble", "fable",
               "feasts", "tsibbledata",
               "fable.prophet", "tidyverse",
               "patchwork", "rio")

# read in the data from a csv and make the tsibble
# change the line below to include your file path
chocolate_month <- rio::import("data/chocolate.csv")
start_date <- lubridate::ymd("2004-01-01")
date_seq <- seq(start_date,
                start_date + months(nrow(chocolate_month)-1),
                by = "1 months")
chocolate_tibble <- tibble(
  dates = date_seq,
  year = lubridate::year(date_seq),
  month = lubridate::month(date_seq),
  value = pull(chocolate_month, chocolate)
)
chocolate_month_ts <- chocolate_tibble |>
  mutate(index = tsibble::yearmonth(dates)) |>
  as_tsibble(index = index)

# generate the plot
choc_plot <- autoplot(chocolate_month_ts, .vars = value) +
  labs(
    x = "Month",
    y = "Searches",
    title = "Relative Number of Google Searches for 'Chocolate'"
  ) +
  theme(plot.title = element_text(hjust = 0.5))

choc_plot
```

### Explore R commands summarizing time series data

::: {.callout-tip icon=false title="Check Your Understanding"}

-   What does each of the following R commands give us?
    -   class(chocolate_month_ts)
    -   head(chocolate_month_ts, 1)
    -   tail(chocolate_month_ts, 1)
    -   guess_frequency(chocolate_month_ts$index)

:::

### Aggregate to the annual level

To help visualize what is happening with a time series, we often aggregate the data in the time series to the annual level. This can make it easier to spot a trend.
The first plot is the time series plot of the raw data, and the second plot is a time series plot of the annual means.

```{r}
chocolate_annual_ts <- summarise(index_by(chocolate_month_ts, year), value = mean(value))

# monthly plot
mp <- autoplot(chocolate_month_ts, .vars = value) +
  labs(
    x = "Month",
    y = "Searches",
    title = "Relative Number of Google Searches for 'Chocolate'"
  ) +
  theme(plot.title = element_text(hjust = 0.5))

# yearly plot
yp <- autoplot(chocolate_annual_ts, .vars = value) +
  labs(
    x = "Year",
    y = "Searches",
    title = "Mean Annual Google Searches for 'Chocolate'"
  ) +
  scale_x_continuous(breaks = seq(2004, max(chocolate_month_ts$year), by = 2)) +
  theme(plot.title = element_text(hjust = 0.5))

# display both simultaneously
mp / yp
```

::: {.callout-tip icon=false title="Check Your Understanding"}

-   What do you observe in the trend?
-   What do you suspect is causing this trend?

:::

### Boxplot of Seasonal Values

To better visualize the effect of seasonal variation, we can make box plots by month.

```{r}
pbox <- ggplot(chocolate_month_ts, aes(x = factor(month), y = value)) +
    geom_boxplot() +
  labs(
    x = "Month Number",
    y = "Searches",
    title = "Boxplots of Google Searches for 'Chocolate' by Month"
  ) +
  theme(plot.title = element_text(hjust = 0.5))

pbox
```

::: {.callout-tip icon=false title="Check Your Understanding"}

-   What do you observe? 
-   Which months tend to have the most searches?  Which months tend to have the fewest seraches?
    -   Can you provide an explanation for this?

:::

## Additive Decomposition 

The additive model for a time series, $\{x_t\}$, is:

$$
  x_t = m_t + s_t + z_t
$$

where $x_t$ is the observed series, $m_t$ is the trend, $s_t$ is the seasonal effect, $z_t$ is random variation, and $t = 1, 2, 3, \ldots, n$. We consider $z_t$ as a sequence of correlated random variables with mean 0.

The residual is the part of the time series that is not explained by the trend or the seasonal effect. This is an estimate of $
For an additive model, the residual is computed as:
$$
  Residual = Value - Trend - Seasonal
$$

The code below produces a plot showing the original data in the time series, the trend, the seasonal effect, and the residual (random component). 

```{r}
#| warning: false

choc_decompose <- chocolate_month_ts |>
    model(feasts::classical_decomposition(value,
        type = "add"))  |>
    components()

autoplot(choc_decompose)
```


::: {.callout-tip icon=false title="Check Your Understanding"}

-   How does the trend and seasonality compare to the plots made previously?
-   What do you observe in the residual? 
    -   Are there any unusual spikes? Use Google to try to determine what may have caused the usual spikes.

:::

## Hands-on Exercise: Exploring simulated time series data (10 min)

### Additive Model

The following code simulates 10 years of a monthly time series with a linear trend and seasonal variation with an additive model.

```{r}
#| warning: false

# Set random seed for reproducibility
set.seed(20) 

# Set parameters & initialize vectors
num_years <- 10
n <- 12 * num_years
sigma <- .75
a <- 0.05
b <- 1
c <- 0.5
trend <- seasonal <- x <- rep(0,n)
time_seq <- seq(1,n)

# Generate correlated error terms
w <- rnorm(n + 4, 0, 1)
z = w + lead(w,1) + lead(w,2) + lead(w,3) + lead(w,4)
z  = head(z, n) / 2

# Get date
year_seq <- lubridate::year(today()) - num_years  + (time_seq - 1) %/% 12
month_seq <- (time_seq - 1) %% 12 + 1
date_seq <- ymd(paste0(year_seq,"-",month_seq,"-01"))

# Get data
for (t in 1:n) {
  trend[t] <- a * t + 10
  seasonal[t] <- b * sin(t / 12 * 2 * pi * 1)  + c * cos(t / 12 * 2 * pi * 3)
  x[t] <- trend[t] + seasonal[t] + z[t]
}

x_df <- data.frame(x = x, trend = trend, seasonal = seasonal)

start_year <- lubridate::year(today()) - num_years
start_date <- lubridate::ymd(paste0(start_year,"-01-01"))

# start_date <- lubridate::ymd("1958-01-01")
date_seq <- seq(start_date,
    start_date + months(nrow(x_df)-1),
    by = "1 months")

x_df_ts <- x_df |>
    mutate(
        date = date_seq,
        Month = tsibble::yearmonth(date)) |>
    as_tsibble(index = Month)
```


The code above has generated simulated data, where the trend is linear with equation

$$
  m_t = \frac{t}{20}
$$

and the seasonal effect follows the function

$$
  s_t = \sin \left( \frac{t\pi}{6} \right) + \frac{1}{2}\cos\left(\frac{t \pi}{18} \right) + 5
$$

Letting $t$ represent the month number across 10 years, we simulate a time series. Click on the tabs below to compare the actual construction of the time series (using the components generated in the code above) to the decomposition in R.

::: panel-tabset
#### Actual Construction

Here is a plot of the components of the simulated data.

```{r}
#| code-fold: true
#| code-summary: "Show the code"

trend_plot <- ggplot(x_df_ts, aes(x=Month, y=trend)) + 
  geom_line() +
  labs(
    title="Plot of Trend", 
    x="Month", 
    y="Trend"
    ) +
  theme(plot.title = element_text(hjust = 0.5))

seasonal_plot <- ggplot(x_df_ts, aes(x=Month, y=seasonal)) + 
  geom_line() +
  labs(
    title="Plot of Seasonal Effect", 
    x="Month", 
    y="Seasonal"
    ) +
  theme(plot.title = element_text(hjust = 0.5))

# trend_plot / seasonal_plot


error_plot <- ggplot(x_df_ts, aes(x = Month, y = x - trend - seasonal)) + 
  geom_line() +
  labs(
    title="Plot of Random Error Term", 
    x="Month", 
    y="Random"
  ) +
  theme(plot.title = element_text(hjust = 0.5))

x_plot <- ggplot(x_df_ts, aes(x=Month, y=x)) + 
  geom_line() +
  labs(
    title="Plot of Simulated Time Series", 
    x="Month", 
    y="x"
  ) +
  theme(plot.title = element_text(hjust = 0.5))


# error_plot / x_plot

x_plot <- x_plot  + labs(title = "True (Simulated) Values", x = NULL)
trend_plot <- trend_plot + labs(title = NULL, x = NULL)
seasonal_plot <- seasonal_plot + labs(title = NULL, x = NULL)
error_plot <- error_plot + labs(title = NULL)


x_plot / trend_plot / seasonal_plot / error_plot 
```


#### Decomposition

Now, we use R to decompose the time series $\{x_t\}$.

```{r}
#| warning: false
#| code-fold: true
#| code-summary: "Show the code"

x_decompose <- x_df_ts |>
    model(feasts::classical_decomposition(x,
        type = "add"))  |>
    components()

autoplot(x_decompose)
```

```{r}
#| include: false
i <- c(0:n)
# a <- 0
x <- a*i + b*sin(i/12*2*pi) + c*cos(i/12*2*pi*3)

df <- data.frame(i, x)

ggplot(df, aes(x=i, y=x)) + 
  geom_line() +
  labs(title="Plot of Function", 
       x="i", 
       y="x")
```
:::

::: {.callout-tip icon=false title="Check Your Understanding"}

-   How does the (estimated) decomposition compare to the theoretical values above?
    -   How well is the trend estimated?
    -   How well is the seasonal effect estimated?
    -   How well is the random effect estimated?
-   Make changes to the simulated data and observe the effect on the plots

:::

### Note on Additive and Multiplicative Models in the Textbook and R

Both the textbook and R use the same model in the additive case:

$$
  x_t = m_t + s_t + z_t
$$

However, there is a discrepancy in the definitions for the mulitplicative models. The textbook defines the multiplicative model as

$$
  x_t = m_t \cdot s_t + z_t
$$
but R defines the multiplicative model as 

$$
  x_t = m_t \cdot s_t \cdot z_t
$$
You can investigate R's definition by executing this command in RStudio.

```{r}
#| eval: false

?classical_decomposition
```


### Multiplicative Model

We use the same simulated data as before, but this time we apply R's multiplicative model. This implies that the error term, $z_t$, has a mean of 1, rather than 0.

```{r}
# Set random seed for reproducibility
set.seed(123) 

# Set parameters & initialize vectors
num_years <- 10
n <- 12 * num_years
sigma <- .75
a <- 0.03
b <- 1
c <- 0.5 
trend <- seasonal <- x <- rep(0,n)
time_seq <- seq(1,n)

# Generate correlated error terms
w <- rnorm(n + 4, 0.2, 0.1) # Changed to a mean of 1 and sd of 0.03
z = w + lead(w,1) + lead(w,2) + lead(w,3) + lead(w,4)
z  = head(z, n)

# Get date
year_seq <- lubridate::year(today()) - num_years  + (time_seq - 1) %/% 12
month_seq <- (time_seq - 1) %% 12 + 1
date_seq <- ymd(paste0(year_seq,"-",month_seq,"-01"))

# Get data
for (t in 1:n) {
  trend[t] <- exp(a * t)
  seasonal[t] <- b * sin(t / 12 * 2 * pi * 1)  + c * cos(t / 12 * 2 * pi * 3) + 5
  x[t] <- trend[t] * seasonal[t] * z[t] # Note R's definition of the mult. model
}

x_df <- data.frame(x = x, trend = trend, seasonal = seasonal)

start_year <- lubridate::year(today()) - num_years
start_date <- lubridate::ymd(paste0(start_year,"-01-01"))

# start_date <- lubridate::ymd("1958-01-01")
date_seq <- seq(start_date,
    start_date + months(nrow(x_df)-1),
    by = "1 months")

x_df_ts <- x_df |>
    mutate(
        date = date_seq,
        Month = tsibble::yearmonth(date)) |>
    as_tsibble(index = Month)
```


The code above simulated data, where the trend is expoential with equation

$$
  m_t = e^{0.03 t}
$$

and the seasonal effect follows the function

$$
  s_t = \sin \left( \frac{t\pi}{6} \right) + \frac{1}{2}\cos\left(\frac{t \pi}{18} \right) + 5
$$

Letting $t$ represent the month number across 10 years, we get:

```{r}
#| echo: false

trend_plot <- ggplot(x_df_ts, aes(x=Month, y=trend)) + 
  geom_line() +
  labs(
    title="Plot of Trend", 
    x="Month", 
    y="Trend"
    ) +
  theme(plot.title = element_text(hjust = 0.5))

seasonal_plot <- ggplot(x_df_ts, aes(x=Month, y=seasonal)) + 
  geom_line() +
  labs(
    title="Plot of Seasonal Effect", 
    x="Month", 
    y="Seasonal"
    ) +
  theme(plot.title = element_text(hjust = 0.5))

# trend_plot / seasonal_plot


error_plot <- ggplot(x_df_ts, aes(x = Month, y = x / trend / seasonal)) + 
  geom_line() +
  labs(
    title="Plot of Random Error Term", 
    x="Month", 
    y="Random"
  ) +
  theme(plot.title = element_text(hjust = 0.5))

x_plot <- ggplot(x_df_ts, aes(x=Month, y=x)) + 
  geom_line() +
  labs(
    title="Plot of Simulated Time Series", 
    x="Month", 
    y="x"
  ) +
  theme(plot.title = element_text(hjust = 0.5))


# error_plot / x_plot

x_plot <- x_plot  + labs(title = "True (Simulated) Values", x = NULL)
trend_plot <- trend_plot + labs(title = NULL, x = NULL)
seasonal_plot <- seasonal_plot + labs(title = NULL, x = NULL)
error_plot <- error_plot + labs(title = NULL)


x_plot / trend_plot / seasonal_plot / error_plot 
```


Now, we use R to decompose the time series $\{x_t\}$.

```{r}
#| warning: false

x_decompose <- x_df_ts |>
    model(feasts::classical_decomposition(x,
        type = "mult"))  |>
    components()

autoplot(x_decompose)
```

```{r}
#| include: false
i <- c(0:n)
# a <- 0
x <- a*i * ( b*sin(i/12*2*pi) + c*cos(i/12*2*pi*3) )

df <- data.frame(i, x)

ggplot(df, aes(x=i, y=x)) + 
  geom_line() +
  labs(title="Plot of Function", 
       x="i", 
       y="x")
```

::: {.callout-tip icon=false title="Check Your Understanding"}

-   How does the (estimated) decomposition compare to the theoretical values above?
    -   How well is the trend estimated?
    -   How well is the seasonal effect estimated?
    -   How well is the random effect estimated?
-   Make changes to the simulated data and observe the effect on the plots

:::


### Which Model Should I Use? Additive or Multiplicative?

Compare the following two time series. 

```{r}
#| echo: false

# Read and clean rexburg weather data

# pacman::p_load("tsibble", "fable",
#                "feasts", "tsibbledata",
#                "fable.prophet", "tidyverse",
#                "patchwork",
#                "ggthemes", "see",   # for okabeito color scheme
#                "tidyverse",
#                "rio",
#                "lubridate"
# )

weather_dat <- rio::import("data/rexburg_weather.csv")

w2 <- weather_dat %>%
  dplyr::select("NAME", "DAY", "MONTH", "YEAR", "DATE", "TMAX") %>%
  pivot_wider(names_from = "NAME", values_from = c("TMAX")) %>%
  rename(
    "byui" = "REXBURG BYU IDAHO, ID US",
    "airport" = "REXBURG MADISON CO AIRPORT, ID US",
    "moody" = "MOODY IDAHO, ID US"
  )

w3 <- w2 %>%
  dplyr::select(byui, airport) %>%
  rename(x = byui, y = airport) %>%
  filter(abs(x - y)<50)
# plot(w3$x, w3$y)
lm_byu <- lm(y~x, data= w3)

w4 <- w2 %>%
  dplyr::select(moody, airport) %>%
  rename(x = moody, y = airport) %>%
  filter(abs(x - y)<50)
# plot(w4$x, w4$y)
lm_moody <- lm(y~x, data= w4)

weather_clean <- w2 %>%
  mutate(byui_pred = predict(lm_byu, data.frame(x=w2$byui))) %>%
  mutate(moody_pred = predict(lm_moody, data.frame(x=w2$moody))) %>%
  mutate(rexburg = airport) %>%
  mutate(rexburg =
           case_when(
             airport > 100 ~ byui_pred,
             !is.na(airport) ~ airport,
             !is.na(byui_pred) ~ byui_pred,
             !is.na(moody_pred) ~ moody_pred,
             TRUE ~ NA
           )
        ) %>%
  mutate(imputed = !(rexburg == round(rexburg, 0))) %>%
  mutate(rexburg = round(rexburg, 0)) %>%
  mutate(dates = mdy(DATE)) %>%
  filter(dates > mdy("08/31/1998")) %>%
  filter(!is.na(rexburg)) %>%
  rename(rexburg_airport_high = rexburg) %>%
  dplyr::select(dates, rexburg_airport_high, imputed) %>%
  filter(dates > mdy("01/01/1999"))

# weather_clean %>%
#   mutate(years = year(dates)) %>%
#   group_by(years) %>%
#   summarize(average = mean(rexburg_airport_high)) %>%
#   plot

rexburg_day <- weather_clean %>%
  mutate(date_seq = dates) %>%
  mutate(
  dates = date_seq,
  year = lubridate::year(date_seq),
  month = lubridate::month(date_seq),
  value = rexburg_airport_high
  ) %>%
  dplyr::select(-date_seq, -imputed) %>%
  tibble()

rexburg_daily_ts <- rexburg_day |>
  mutate(index = dates) |>
  as_tsibble(index = index)

rexburg_annual_ts <- summarise(index_by(rexburg_daily_ts, year), value = mean(value))

rexburg_weather_plot <- autoplot(rexburg_daily_ts, .vars = value) +
  labs(
    x = "Date",
    y = "High Temp (F)",
    title = "Daily High Temperature in Rexburg, Idaho"
  ) +
  theme(plot.title = element_text(hjust = 0.5))

# mp <- autoplot(rexburg_daily_ts, .vars = value) +
#   labs(y = "high temperature")
# yp <- autoplot(rexburg_annual_ts) +
#   labs(y = "high temperature")
# # +
# #   scale_x_continuous(breaks = seq(1900, 2010, by = 2))
# mp / yp



##################### S&P 500

replaceCommas<-function(x){
  x<-as.numeric(gsub("\\,", "", x))
}

sp500_dat <- rio::import("data/sp500.csv") |>
  mutate(dates = mdy(Date))

sp500_day <- sp500_dat |>
  mutate(date_seq = dates) |>
  mutate(
    dates = date_seq,
    year = lubridate::year(date_seq),
    month = lubridate::month(date_seq),
    value = replaceCommas(Close)
  ) |>
  dplyr::select(-date_seq) |>
  tibble()

sp500_ts <- sp500_day |>
  mutate(index = dates) |>
  as_tsibble(index = index)

# sp500_annual_ts <- summarise(index_by(sp500_ts, year), value = mean(value))
# 
# temp <- sp500_ts |> filter(month == 7 & day(dates) == 1) |> 
#   select(Date, year) 
# 
# temp2 <- sp500_annual_ts |> 
#   right_join(temp)

sp500_plot <- autoplot(sp500_ts, .vars = value) +
  labs(
    x = "Date",
    y = "Closing Price",
    title = "Daily Closing Price of the S&P 500 Stock Index"
  )  +
  theme(plot.title = element_text(hjust = 0.5))


rexburg_weather_plot / sp500_plot
```


::: {.callout-tip icon=false title="Check Your Understanding"}

Complete a table like the following in your Learning Journal to compare characteristics of these two time series.

|                                          | Rexburg Temperature | S&P 500 Closing Price |
|------------------------------------------|---------------------|-----------------------|
| Deterministic or stochastic trend?       |                     |                       | 
| Is there a seasonal effect?              |                     |                       | 
| Is there evidence of cycles?             |                     |                       | 
| Does the variation get bigger over time? |                     |                       | 
| Additive or multiplicative model?        |                     |                       | 

:::



<!-- ## POSSIBLY: DEMONSTRATE THAT TWO UNRELATED TIME SERIES WILL BE CORRELATED IF THEY BOTH CONTAIN A TREND---------- -->





<!-- Compare Rexburg weather and Basel, Switzerland Weather. (Better yet, Melbourne, Austrailia) -->





## Recap (5 min)

-   Review objectives and key concepts
-   Clarify questions on time series data


## Homework

::: {.callout-note icon=false}

## Download Assignment

<!-- ## need to update href link to correct files when we get them -->

<a href="https://byuistats.github.io/timeseries/homework/HW-Draft.qmd" download="homework_1_2.qmd"> homework_1_2.qmd </a>

:::

<a href="javascript:showhide('Solutions')"
style="font-size:.8em;">Which Model?</a>

::: {#Solutions style="display:none;"}

Solutions to Which Model Should I Use?:

|                                          | Rexburg Temperature | S&P 500 Closing Price |
|------------------------------------------|---------------------|-----------------------|
| Deterministic or stochastic trend?       |  Deterministic      |  Stochastic           | 
| Is there a seasonal effect?              |  Yes                |  No                   | 
| Is there evidence of cycles?             |  Yes                |  Yes                  |
| Does the variation get bigger over time? |  No                 |  Yes                  |  
| Additive or multiplicative model?        |  Additive           |  Multiplicative       |
:::


