---
# title: "Fitted Models"
subtitle: "Chapter 4: Lesson 4"
format: html
editor: source
sidebar: false
---

```{r}
#| include: false
source("common_functions.R")
```

```{=html}
<script type="text/javascript">
 function showhide(id) {
    var e = document.getElementById(id);
    e.style.display = (e.style.display == 'block') ? 'none' : 'block';
 }
 
 function openTab(evt, tabName) {
    var i, tabcontent, tablinks;
    tabcontent = document.getElementsByClassName("tabcontent");
    for (i = 0; i < tabcontent.length; i++) {
        tabcontent[i].style.display = "none";
    }
    tablinks = document.getElementsByClassName("tablinks");
    for (i = 0; i < tablinks.length; i++) {
        tablinks[i].className = tablinks[i].className.replace(" active", "");
    }
    document.getElementById(tabName).style.display = "block";
    evt.currentTarget.className += " active";
 }    
</script>
```


## Learning Outcomes

{{< include outcomes/chapter_4_lesson_4_outcomes.qmd >}}




## Preparation

-   Read Sections 4.6-4.7



## Learning Journal Exchange (10 mins)

-   Review another student's journal

-   What would you add to your learning journal after reading another student's?

-   What would you recommend the other student add to their learning journal?

-   Sign the Learning Journal review sheet for your peer


## Class Activity: Global Warming (15 min)

```{r}
temps_ts <- rio::import("data/global_temparature.csv") |>
  as_tsibble(index = year)

temps_ts |> autoplot(.vars = change) +
    labs(
      x = "Year",
      y = "Temperature (Celsius)",
      title = "Change in Mean Annual Global Temperature"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5)
    )

pacf(temps_ts$change)

global_ar <- temps_ts |>
    model(AR(change ~ order(1:9)))

mean(temps_ts$change)

tidy(global_ar) |> filter(str_detect(term, "ar")) |> nrow()

tidy(global_ar) |> filter(str_detect(term, "ar")) |> pull(estimate)

coeffs <- tidy(global_ar) |> filter(str_detect(term, "ar")) |> pull(estimate)
coeffs[1]

residuals(global_ar) |> ACF(lag_max = 50) |> autoplot(.vars = .resid)
```



## Small Group Activity: SectionTitle (xxx min)
## Class Activity: SectionTitle (xxx min)


<!-- Check your Understanding -->

::: {.callout-tip icon=false title="Check Your Understanding"}

-   Question1
-   Question2

:::





## Homework Preview (5 min)

-   Review upcoming homework assignment
-   Clarify questions


::: {.callout-note icon=false}

## Download Homework

<a href="https://byuistats.github.io/timeseries/homework/homework_4_4.qmd" download="homework_4_4.qmd"> homework_4_4.qmd </a>

:::


<a href="javascript:showhide('Solutions1')"
style="font-size:.8em;">Class Activity</a>
  
::: {#Solutions1 style="display:none;"}
    

:::




<a href="javascript:showhide('Solutions')"
style="font-size:.8em;">Class Activity</a>
  
::: {#Solutions2 style="display:none;"}
    

:::




<a href="javascript:showhide('Solutions3')"
style="font-size:.8em;">Class Activity</a>
  
::: {#Solutions3 style="display:none;"}
    

:::



