---
title: "White Noise and Random Walks: Part 2"
subtitle: "Chapter 4: Lesson 2"
format: html
editor: source
sidebar: false
---

```{r}
#| include: false
source("common_functions.R")
```

```{=html}
<script type="text/javascript">
 function showhide(id) {
    var e = document.getElementById(id);
    e.style.display = (e.style.display == 'block') ? 'none' : 'block';
 }
 
 function openTab(evt, tabName) {
    var i, tabcontent, tablinks;
    tabcontent = document.getElementsByClassName("tabcontent");
    for (i = 0; i < tabcontent.length; i++) {
        tabcontent[i].style.display = "none";
    }
    tablinks = document.getElementsByClassName("tablinks");
    for (i = 0; i < tablinks.length; i++) {
        tablinks[i].className = tablinks[i].className.replace(" active", "");
    }
    document.getElementById(tabName).style.display = "block";
    evt.currentTarget.className += " active";
 }    
</script>
```


## Learning Outcomes

{{< include outcomes/chapter_4_lesson_2_outcomes.qmd >}}




## Preparation

-   Read Sections 4.3.3-4.3.7 and 4.4



## Learning Journal Exchange (10 mins)

-   Review another student's journal

-   What would you add to your learning journal after reading another student's?

-   What would you recommend the other student add to their learning journal?

-   Sign the Learning Journal review sheet for your peer



## Class Activity: Random Walks Continued (15 min)

Recall the definition of a random walk:

$\{x_t\}$ is a **random walk** if it can be expressed as
$$
  x_{t} = x_{t-1} + w_{t}
$$
where $\{w_t\}$ is a white noise series.

There are other ways to represent this expression.

<!-- Check Your Understanding -->

::: {.callout-tip icon=false title="Check Your Understanding"}

-   Notice that $x_{t-1} = x_{t-2} + w_{t-1}$ and $x_{t-2} = x_{t-3} + w_{t-2}$. Use this to explain why it is possible to write $x_t$ as
    $$
      x_{t} = \sum\limits_{i=-\infty}^{t} w_{i} = \cdots + w_{t-3} + w_{t-2} + w_{t-1} + w_{t} 
    $$
    where $\{w_t\}$ is a DWN time series.
    
    If the time series is finite, we can write $x_t$ as:
    $$
      x_{t} = w_1 + w_2 + w_3 + \cdots + w_{t-3} + w_{t-2} + w_{t-1} + w_{t} 
    $$
    
:::


<!-- Note that this implies that: -->
<!-- \begin{align*} -->
<!--   x_{t} &= x_{t-1} + w_{t} \\ -->
<!--   x_{t-1} &= x_{t-2} + w_{t-1} \\ -->
<!--   x_{t-2} &= x_{t-3} + w_{t-2} \\ -->
<!--   x_{t-3} &= x_{t-4} + w_{t-3} \\ -->
<!--   \vdots ~~ &= ~~~~~~~~~~~~ \vdots \\ -->
<!--   x_{4} &= x_{3} + w_{4} \\ -->
<!--   x_{3} &= x_{2} + w_{3} \\ -->
<!--   x_{2} &= x_{1} + w_{2} \\ -->
<!--   x_{1} &= w_{1}  -->
<!-- \end{align*} -->

<!-- We can write this as -->

<!-- \begin{align*} -->
<!--   x_{t}  -->
<!--     &= x_{t-1} + w_{t} \\ -->
<!--     &= \underbrace{x_{t-2} + w_{t-1}} + w_{t} \\ -->
<!--     &= \underbrace{x_{t-3} + w_{t-2}} + w_{t-1} + w_{t} \\ -->
<!--     &= \underbrace{x_{t-4} + w_{t-3}} + w_{t-2} + w_{t-1} + w_{t} \\ -->
<!--     &= ~~~~ \vdots \\ -->
<!--     &= \underbrace{x_3 + w_4} + \cdots + w_{t-4} + w_{t-3} + w_{t-2} + w_{t-1} + w_{t} \\ -->
<!--     &= \underbrace{x_2 + w_3} + w_4 + \cdots + w_{t-4} + w_{t-3} + w_{t-2} + w_{t-1} + w_{t} \\ -->
<!--     &= \underbrace{x_1 + w_2} + w_3 + w_4 + \cdots + w_{t-4} + w_{t-3} + w_{t-2} + w_{t-1} + w_{t} \\ -->
<!--     &= w_1 + w_2 + w_3 + w_4 + \cdots + w_{t-4} + w_{t-3} + w_{t-2} + w_{t-1} + w_{t}  -->
<!-- \end{align*} -->


### Definition of the Backward Shift Operator

This process of *back substitution* is so common, we define notation to handle it.

::: {.callout-note icon=false title="Definition of the Backward Shift Operator"}

We define the **backward shift operator** or the **lag operator**, $\mathbf{B}$, as:
$$
  \mathbf{B} x_t = x_{t-1}
$$
where $\{x_t\}$ is any time series.

We can apply this operator repeatedly. We will use exponential notation to indicate this.

$$
  \mathbf{B}^2 x_t = \mathbf{B} \mathbf{B} x_t = \mathbf{B} ( \mathbf{B} x_t ) x_t = \mathbf{B} x_{t-1} = \mathbf{B} x_{t-2}
$$

In general, 
$$
  \mathbf{B}^n x_t = \underbrace{\mathbf{B} \cdot \mathbf{B} \cdot \cdots \cdot \mathbf{B}}_{n ~ \text{terms}} x_t = \mathbf{B}^{n-1} ( \mathbf{B} x_t ) = \mathbf{B}^{n-1} ( x_{t-1} ) = \mathbf{B}^{n-2} ( x_{t-2} ) = \cdots = \mathbf{B} x_{t-(n-1)} = x_{t-n}
$$

:::

Note that we can use the backwards shift operator to write a random walk in terms of the realizations of a white noise process. Consider the following process:
\begin{align*}
  x_t 
    &= \frac{1}{5} w_{t-2} + \frac{1}{3} w_{t-1} + w_{t} \\
    &= \frac{1}{5} \mathbf{B}^2 w_t + \frac{1}{3} \mathbf{B} w_t + w_{t} \\
    &= \left( \frac{1}{5} \mathbf{B}^2 + \frac{1}{3} \mathbf{B} + 1 \right) w_{t} \\
\end{align*}
<!-- Or more generally, -->
<!-- \begin{align*} -->
<!--   x_t  -->
<!--     &= w_1 + w_2 + w_3 + \cdots + w_{t-3} + w_{t-2} + w_{t-1} + w_{t} \\ -->
<!--     &= \mathbf{B}^{t-1} w_t + \mathbf{B}^{t-2} w_t + \mathbf{B}^{t-3} w_t + \cdots + \mathbf{B}^3 w_t + \mathbf{B}^2 w_t + \mathbf{B} w_t + w_{t} \\ -->
<!--     &= \left( \mathbf{B}^{t-1} + \mathbf{B}^{t-2} + \mathbf{B}^{t-3} + \cdots + \mathbf{B}^3 + \mathbf{B}^2 + \mathbf{B} + 1 \right) w_{t} \\ -->
<!-- \end{align*} -->

We will practice applying this operator.

<!-- Check Your Understanding -->

::: {.callout-tip icon=false title="Check Your Understanding"}

Let $\{x_t\}$ be a time series with the following values. 

<center>
```{r, results='asis'}
#| echo: false

set.seed(6)
n <- 8
d_operator <- data.frame(t = c(1:n), x = sample(1:15, n, replace = FALSE)) |>
  mutate(diff = t - n)

#cat( paste( paste0("$x_{t", ifelse(d_operator$t==n,"",d_operator$t-n), "} = ", d_operator$x, "$"), collapse = ",$~$ " ) ) 

cat( paste( paste0("$x_{", d_operator$t, "} = ", d_operator$x, "$"), collapse = ",$~$ " ) ) 

# Computes the value of the "power_on_d"^th difference from x_n
d_value <- function(power_on_d = 0) {
  out <- d_operator |> #### Note the use of this global variable
    filter(diff == -power_on_d) |>
    dplyr::select(x) |>
    pull()
  
  return(out)
}


ts_val <- function(t_value) {
  out <- d_operator |> #### Note the use of this global variable
    filter(t == t_value) |>
    dplyr::select(x) |>
    pull()
  
  return(out)
}
```
</center>
Evaluate the following. 

-   $\mathbf{B} x_8$
-   $\mathbf{B}^5 x_8$
-   $(\mathbf{B}^5 - \mathbf{B} ) x_8$
-   $( \mathbf{B}^2 - 6 \mathbf{B} + 9 ) x_8$
-   $( (\mathbf{B} - 6 )\mathbf{B} + 9 ) x_8$
-   $( \mathbf{B} - 3 )^2 x_8 =  ( \mathbf{B} - 3 ) \left[ ( \mathbf{B} - 3 ) x_8 \right]$
-   $( 1 - \frac{1}{2} \mathbf{B} - \frac{1}{4} \mathbf{B}^2 - \frac{1}{8} \mathbf{B}^3 ) x_8$

:::




### Simulation

The following simulation illustrates a random walk.

```{=html}
 <iframe id="Randomwalk" src="https://posit.byui.edu/content/f9a0690a-87ce-423c-9d9c-4dfd0188dd86" style="border: none; width: 100%; height: 1100px" frameborder="0"></iframe>
```

<!-- Check Your Understanding -->

::: {.callout-tip icon=false title="Check Your Understanding"}

-   What do you notice about this time series?
-   What characteristics do you observe in the correlogram?
-   How does this compare to the time series and correlogram for the DWN process?

:::

<!-- Move the stuff above this line into 4-1 -->

<!-- Move the stuff above this line into 4-1 -->

<!-- Move the stuff above this line into 4-1 -->

<!-- Move the stuff above this line into 4-1 -->

<!-- Move the stuff above this line into 4-1 -->

<!-- Move the stuff above this line into 4-1 -->

<!-- Move the stuff above this line into 4-1 -->

<!-- Move the stuff above this line into 4-1 -->

<!-- Move the stuff above this line into 4-1 -->

<!-- Move the stuff above this line into 4-1 -->

<!-- Move the stuff above this line into 4-1 -->

<!-- Move the stuff above this line into 4-1 -->

<!-- Move the stuff above this line into 4-1 -->

<!-- Move the stuff above this line into 4-1 -->

<!-- Move the stuff above this line into 4-1 -->

<!-- Move the stuff above this line into 4-1 -->

<!-- Move the stuff above this line into 4-1 -->

<!-- Move the stuff above this line into 4-1 -->

<!-- Move the stuff above this line into 4-1 -->

<!-- Move the stuff above this line into 4-1 -->

<!-- Move the stuff above this line into 4-1 -->

<!-- Move the stuff above this line into 4-1 -->

### Second-Order Properties of a Random Walk

The second-order properties of a random walk are summarized below. 

::: {.callout-note icon=false title="Second-Order Properties of a Random Walk"}
If $\{x_t\}_{t=1}^n$ is a random walk, then the population has the following properties. 

$$ \mu_x = 0 $$
and
$$
  cov(x_t, x_{t+k}) = t \sigma^2
$$

::: {.callout-tip title="Click here for a proof of the equation for $cov(x_t,x_{t+k})$" collapse=true}

Why is $cov(x_t, x_{t+k}) = t \sigma^2$?

First, note that that since the terms in the white noise series are independent, 

$$
cov ( w_i, w_j ) = 
  \begin{cases}
    \sigma^2, & \text{if } ~ i=j \\
    0, & \text{otherwise}
  \end{cases}
$$

Also, when random variables are independent, the covariance of a sum is the sum of the covariance.

Hence,
\begin{align*}
  cov(x_t, x_{t+k})
    &= cov ( \sum_{i=1}^t w_i, \sum_{j=1}^{t+K} w_j ) \\
    &= \sum_{i=j} cov ( w_i, w_j ) \\
    &= \sum_{i=1}^t \sigma^2 \\
    &= t \sigma^2
\end{align*}

:::

If $k>0$ and $t>0$, the correlation function is

$$
  \rho_k 
  = 
    \frac{
            cov(x_t, x_{t+k})
          }{
            \sqrt{var(x_t)} \sqrt{var(x_{t+k})}
          }
  =
    \frac{t \sigma^2}{\sqrt{t \sigma^2} \sqrt{(t+k) \sigma^2}}
  =
    \frac{1}{\sqrt{1+\frac{k}{t}}}
$$

:::

Note that the covariance of a random walk process depends on $t$. Hence, random walks are non-stationary. The variance is unbounded as $t$ increases. That implies a random walk will not provide good predictions in the long term.

Note that if $0 < k \ll t$, then $\rho_k \approx 1$. Because of this, a correlogram for a random walk will typically demonstrate positive autocorrelations that start near 1 and slowly decrease as $k$ increases.


### Differencing a Time Series 

<a id="differencing">Computing</a> the difference between successive terms of a random walk leads to a discrete white noise series. In many cases, differencing sequential terms of a non-stationary process can lead to a stationary process of differences.
We use the code below to obtain the daily closing stock prices for any publicly-traded company. 

```{r}
#| code-fold: true
#| code-summary: "Show the code"

# Set symbol and date range
symbol <- "MCD"
company <- "McDonald's"
date_start <- "2021-01-01"
date_end <- "2024-01-01"

# Fetch stock prices (can be used to get new data)
stock_df <- tq_get(symbol, from = date_start, to = date_end, get = "stock.prices")

# Store data in a static file
stock_df |> rio::export("data/stock_price_mcd.parquet")

# Retrieve static file
# stock_df <- rio::import("data/stock_price_mcd.parquet")

# Transform data into tibble
stock_ts <- stock_df %>%
  mutate(
    dates = date, 
    value = adjusted
  ) %>%
  dplyr::select(dates, value) %>%
  as_tibble() %>% 
  arrange(dates) |>
  mutate(diff = value - lag(value)) |>
  as_tsibble(index = dates, key = NULL)
```

Consider the closing price of `r company` stock illustrated in the time plot in @fig-example-stock-price-plot. 
@fig-example-stock-difference-plot gives the differences in the closing prices of the stock as a time series.


<!-- Beginning of two columns -->
::: columns
::: {.column width="45%"}


```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| label: fig-example-stock-price-plot
#| fig-cap: "Plot of the daily closing price of the stock"
#| fig.asp: 0.61

plot_ly(stock_ts, x = ~dates, y = ~value, type = 'scatter', mode = 'lines') %>%
  layout(
    xaxis = list(title = paste0("Dates (", format(ymd(date_start), "%d/%m/%Y"), " to ", format(ymd(date_end), "%d/%m/%Y"), ")" ) ),
    yaxis = list(title = "Closing Price (US$)"),
    title = paste0("Time Plot of ", symbol, " Daily Closing Price")
  )
```

:::

::: {.column width="10%"}
<!-- empty column to create gap -->
:::

::: {.column width="45%"}

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| label: fig-example-stock-difference-plot
#| fig-cap: "Plot of the stock price differences"
#| fig.asp: 0.61

# Generate time series plot using plot_ly
plot_ly(stock_ts, x = ~dates, y = ~diff, type = 'scatter', mode = 'lines') %>%
  layout(
    xaxis = list(title = paste0("Dates (", format(ymd(date_start), "%d/%m/%Y"), " to ", format(ymd(date_end), "%d/%m/%Y"), ")" ) ),
    yaxis = list(title = "Closing Price (US$)"),
    title = paste0("Difference of ", symbol, " Daily Closing Price")
  )

```

:::
:::
<!-- End of two columns -->

@fig-example-stock-correlogram is the correlogram for the original `r company` stock price time series.
@fig-example-diff-correlogram gives the correlogram for the differences in successive closing stock prices.

<!-- Beginning of two columns -->
::: columns
::: {.column width="45%"}

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| label: fig-example-stock-correlogram
#| fig-cap: "Correlogram of the stock prices"
#| fig.asp: 0.61

acf(stock_ts$value, plot=TRUE, type = "correlation", lag.max = 25)
```


:::

::: {.column width="10%"}
<!-- empty column to create gap -->
:::

::: {.column width="45%"}


```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| label: fig-example-diff-correlogram
#| fig-cap: "Correlogram of the stock prices"
#| fig.asp: 0.61

acf(stock_ts$diff |> na.omit(), plot=TRUE, type = "correlation", lag.max = 25)
```

:::
:::
<!-- End of two columns -->

@fig-example-stock-difference-histogram is a histogram of the differences.
On the right, we give the variance of the differences in the stock prices. This is a simple measure of the volatility of the stock, or in other words, how much the price changes in a day.

<!-- Beginning of two columns -->
::: columns
::: {.column width="45%"}

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| label: fig-example-stock-difference-histogram
#| fig-cap: "Histogram of the stock price differences"
#| warning: false
#| fig.asp: 0.61

# Hisogram of differences in stock prices
stock_ts |>
  mutate(
    density = dnorm(diff, mean(stock_ts$diff, na.rm = TRUE), sd(stock_ts$diff, na.rm = TRUE))
  ) |>
  ggplot(aes(x = diff)) +
    geom_histogram(aes(y = after_stat(density)),
        color = "white", fill = "#56B4E9", binwidth = 1) +
    geom_line(aes(x = diff, y = density)) +
    theme_bw() +
    labs(
      x = "Difference",
      y = "Frequency",
      title = "Histogram of Difference in the Closing Stock Prices"
    ) +
    theme(
      plot.title = element_text(hjust = 0.5)
    )
```

:::

::: {.column width="10%"}
<!-- empty column to create gap -->
:::

::: {.column width="45%"}

The variance of the differences is `r var(stock_ts$diff, na.rm = TRUE)|> round(3)`.

:::
:::
<!-- End of two columns -->




Notice that the values in the correlogram start at 1 and slowly decay as $k$ increases. This is exactly what we would expect from a random walk. It is also interesting that the differences are nearly normally distributed.

### Defintion of the Difference Operator

Differencing nonstationary time series often leads to a stationary series, so we will define a formal operator to express this process.

::: {.callout-note icon=false title="Definition of the Difference Operator"}

The **difference operator**, $\nabla$, is defined as:

$$\nabla x_t = x_t - x_{t-1} = (1-\mathbf{B}) x_t$$

Higher-order differencing can be denoted

$$\nabla^n x_t = (1-\mathbf{B})^n x_t$$

:::

To see what this expression gives us, note that $\nabla$ gives a new time series that is comprised of the differences between successive terms of the original time series. The operator $\nabla^2$ generates a time series that is comprised of the differences between successive terms of the *differenced* time series. It is the difference of the differences, or the second difference.

<!-- Check Your Understanding -->

::: {.callout-tip icon=false title="Check Your Understanding"}

Consider the following time series, where $n=`r n`$:

<center>
```{r, results='asis'}
#| echo: false

cat( paste( paste0("$x_{", d_operator$t, "} = ", d_operator$x, "$"), collapse = ",$~$ " ) ) 
```
</center>


<!-- Beginning of two columns -->
::: columns
::: {.column width="45%"}

-   Find the first differences, $\nabla x_t$
-   Find the second differences, $\nabla^2 x_t$.
-   Fill in the missing steps:
\begin{align*}
\nabla^2 x_8 &= (1-\mathbf{B} )^2 x_8 \\
  &= (1-\mathbf{B} ) \left[ (1-\mathbf{B} ) x_8 \right] \\
  &  ~~~~~~~~~~~~~~~~~~~~~~ ⋮ \\
  &= (x_8-x_7)-(x_7-x_6)
\end{align*}
and check that this is equal to the last term in the sequence of second differences.

:::

::: {.column width="10%"}
<!-- empty column to create gap -->
:::

::: {.column width="45%"}

```{r}
#| echo: false

d_operator |>
  dplyr::select(-diff) |>
  mutate(diff1 = x - lag(x)) |>
  mutate(diff2 = diff1 - lag(diff1)) |>
  rename(
    "$$t$$" = t,
    "$$x_t$$" = x,
    "$$\\nabla x_t$$" = diff1,
    "$$\\nabla^2 x_t$$" = diff2
  ) |>
  replace_na_with_char("") |>
  replace_cells_with_char(rows = 1:8, cols = 3:4) |>
  display_table("0.5in")
```

:::
:::
<!-- End of two columns -->

:::
<!-- End of Check your understanding -->


### Small-Group Activity: Computing Differences

The difference operator can be helpful in identifying the functional underpinnings of a trend.
If a function is linear, then the first differences of equally-spaced values will be constant. If a function is quadratic, then the second differences of equally-spaced values will be constant. If a function is cubic, then the third differences of equally-spaced values will be constant, and so on. 

Compute the differences specified below.




#### Linear 

<!-- Beginning of two columns -->
::: columns
::: {.column width="45%"}

```{r}
#| echo: false
#| fig-asp: 1
#| fig-width: 4

linear_fcn <- function(t) { return(5 + 2.5 * t) }

df_linear <- data.frame(t = 1:9) |>
  mutate(x = linear_fcn(t)) |>
  mutate(diff1 = x - lag(x)) |>
  mutate(diff2 = diff1 - lag(diff1)) |>
  mutate(diff3 = diff2 - lag(diff2)) 

df_linear |>
  ggplot(aes(x = t, y = x)) +
  geom_function(fun = linear_fcn, color = "#D55E00", xlim=c(0.8, 9.2)) +
  geom_point() +
  scale_x_continuous(breaks=seq(1, 9, by = 1)) +
  scale_y_continuous(breaks=seq(0, 30, by = 5)) +
  coord_cartesian(ylim = c(0,30)) +
  theme_bw() +
  theme(panel.grid.minor = element_blank()) +
  labs(
    title = "Linear Function",
    subtitle = "x = 2.5 t + 5"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )
```

:::

::: {.column width="10%"}
<!-- empty column to create gap -->
:::

::: {.column width="45%"}

```{r}
#| echo: false

df_linear |>
  rename(
    "$$t$$" = t,
    "$$x_t$$" = x,
    "$$\\nabla x_t$$" = diff1,
    " " = diff2, 
    "  " = diff3
  ) |>
  replace_na_with_char("") |>
  replace_cells_with_char(rows = 1:9, cols = 3:5, new_char = "") |>
  display_table("0.5in")
```

:::
:::
<!-- End of two columns -->




 
#### Quadratic 


<!-- Beginning of two columns -->
::: columns
::: {.column width="45%"}

```{r}
#| echo: false
#| fig-asp: 1
#| fig-width: 4

quadratic_fcn <- function(t) { return(2 * (t - 4.75)^2 - 13.125) }

df_quadratic <- data.frame(t = 1:9) |>
  mutate(x = quadratic_fcn(t)) |>
  mutate(diff1 = x - lag(x)) |>
  mutate(diff2 = diff1 - lag(diff1)) |>
  mutate(diff3 = diff2 - lag(diff2))

df_quadratic |>
  ggplot(aes(x = t, y = x)) +
  geom_function(fun = quadratic_fcn, color = "#D55E00", xlim=c(0.8, 9.2)) +
  geom_point() +
  scale_x_continuous(breaks=seq(1, 9, by = 1)) +
  scale_y_continuous(breaks=seq(-15, 30, by = 5)) +
  coord_cartesian(ylim = c(-14,26)) +
  theme_bw() +
  theme(panel.grid.minor = element_blank()) +
  labs(
    title = "Quadratic Function",
    subtitle = "x = 2 (t - 4.75)^2 - 13.125"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )
```

:::

::: {.column width="10%"}
<!-- empty column to create gap -->
:::

::: {.column width="45%"}

```{r}
#| echo: false

df_quadratic |>
  rename(
    "$$t$$" = t,
    "$$x_t$$" = x,
    "$$\\nabla x_t$$" = diff1,
    "$$\\nabla^2 x_t$$" = diff2,
    "  " = diff3
  ) |>
  replace_na_with_char("") |>
  replace_cells_with_char(rows = 1:9, cols = 3:5, new_char = "") |>
  display_table("0.5in")
```

:::
:::
<!-- End of two columns -->



#### Cubic 


<!-- Beginning of two columns -->
::: columns
::: {.column width="45%"}

```{r}
#| echo: false
#| fig-asp: 1
#| fig-width: 4

cubic_fcn <- function(t) { return((t - 5.5) * (t - 2) * (t - 7) / 10) }

df_cubic <- data.frame(t = 1:9) |>
  mutate(x = cubic_fcn(t)) |>
  mutate(diff1 = x - lag(x)) |>
  mutate(diff2 = diff1 - lag(diff1)) |>
  mutate(diff3 = diff2 - lag(diff2)) 

df_cubic |>
  ggplot(aes(x = t, y = x)) +
  geom_function(fun = cubic_fcn, color = "#D55E00", xlim=c(0.8, 9.2)) +
  geom_point() +
  scale_x_continuous(breaks=seq(1, 9, by = 1)) +
  scale_y_continuous(breaks=seq(-5, 8, by = 1)) +
  theme_bw() +
  theme(panel.grid.minor = element_blank()) +
  labs(
    title = "Cubic Function",
    subtitle = "x = (t - 2)(t - 7)(t - 5.5)/10"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )
```

:::

::: {.column width="10%"}
<!-- empty column to create gap -->
:::

::: {.column width="45%"}

```{r}
#| echo: false

df_cubic |>
  rename(
    "$$t$$" = t,
    "$$x_t$$" = x,
    "$$\\nabla x_t$$" = diff1,
    "$$\\nabla^2 x_t$$" = diff2,
    "$$\\nabla^3 x_t$$" = diff3
  ) |>
  replace_na_with_char("") |>
  replace_cells_with_char(rows = 1:9, cols = 3:5, new_char = "") |>
  display_table("0.5in")
```

:::
:::
<!-- End of two columns -->




## Small-Group Activity: Differencing Stock Prices (15 min)

In this activity, you will apply what you have learned to a new stock.

<!-- Check Your Understanding -->

::: {.callout-tip icon=false title="Check Your Understanding"}

Modify the <a href="#differencing">code used to get the prices of `r company` stock</a> to download closing stock prices for a different publicly-traded company over a time period of your choice. Do the following.

-   Indicate which company you have chosen, the stock symbol, and the time period.
-   Create a time plot of the daily closing stock prices.
-   Produce a time plot of the differences in the daily closing stock prices.
-   Create a correlogram of the stock prices
-   Create a correlogram of the differences
-   Generate a histogram of the difference in the stock prices and superimpose the corresponding normal density.
-   Compute the variance of the differences
-   Compare your results with those from the other teams of students.

:::


```{r}
#| echo: false

# This is here so it will compile
# With the values commented out below, it would not compile without a model_hw object.
stock_ts0 <- ts(rnorm(50))
model_hw <- HoltWinters(stock_ts0, alpha = 1, gamma = FALSE)
```



<!-- This should be a stock where the ACF of the first differences is significant for k=1, but not for k>1. -->
<!-- This should be a stock where the ACF of the first differences is significant for k=1, but not for k>1. -->
<!-- This should be a stock where the ACF of the first differences is significant for k=1, but not for k>1. -->
<!-- ~~~ -->
<!--                                 This can be a little hard to find                                      -->
<!-- ~~~ -->
<!-- The correlogram for the residuals from the Holt-Winters model match the correlogram for the differences. Why?  -->
<!-- The correlogram for the residuals from the Holt-Winters model match the correlogram for the differences. Why?  -->
<!-- The correlogram for the residuals from the Holt-Winters model match the correlogram for the differences. Why?  -->
<!-- It's like there is no improvement in the fit of the model after doing Holt-Winters. -->
<!-- It's like there is no improvement in the fit of the model after doing Holt-Winters. -->
<!-- It's like there is no improvement in the fit of the model after doing Holt-Winters. -->


<!-- ## Class Activity: Integrated Autoregressive Model (Section 4.4.2) (15 min) -->

<!-- ```{r} -->
<!-- #| echo: false -->

<!-- # Set symbol and date range -->
<!-- symbol <- "BAC" -->
<!-- company <- "Bank of America" -->
<!-- date_start <- "2021-01-01" -->
<!-- date_end <- "2024-01-01" -->
<!-- ``` -->

<!-- This is a time plot of the daily closing price of `r company` stock (symbol: `r symbol`) from `r format(ymd(date_start), "%d %b %Y")` to `r format(ymd(date_end), "%d %b %Y")`. -->

<!-- ```{r} -->
<!-- #| echo: false -->

<!-- # Fetch stock prices (can be used to get new data) -->
<!-- stock_df <- tq_get(symbol, from = date_start, to = date_end, get = "stock.prices") -->

<!-- # Transform data into tibble -->
<!-- stock_ts <- stock_df %>% -->
<!--   mutate( -->
<!--     dates = date,  -->
<!--     value = adjusted -->
<!--   ) %>% -->
<!--   dplyr::select(dates, value) %>% -->
<!--   as_tibble() %>%  -->
<!--   arrange(dates) |> -->
<!--   mutate(diff = value - lag(value)) |> -->
<!--   as_tsibble(index = dates, key = NULL) -->

<!-- stock_ts |> -->
<!--   autoplot(.vars = value) +  -->
<!--     labs( -->
<!--       title = paste("Time Plot of", company, "Stock Prices"), -->
<!--       subtitle =  -->
<!--         paste0( -->
<!--           format(min(stock_ts$dates), "%d %b %Y"), -->
<!--             " - ", -->
<!--           format(max(stock_ts$dates), "%d %b %Y") -->
<!--           ), -->
<!--       x = "Date", -->
<!--       y = "Daily Closing Price", -->
<!--     ) + -->
<!--     theme( -->
<!--       plot.title = element_text(hjust = 0.5), -->
<!--       plot.subtitle = element_text(hjust = 0.5) -->
<!--     ) -->
<!-- ``` -->

<!-- Consider the correlogram of the first differences in the stock prices. -->

<!-- ```{r} -->
<!-- #| echo: false -->

<!-- acf(stock_ts$diff |> na.omit(), main = paste("ACF of First Difference of" ,company, "Stock Prices")) -->
<!-- ``` -->

<!-- Notice that for a lag of $k=1$, there is still a significant autocorrelation in the differences. This suggests that a more sophisticated model might be necessary. Note that for $k>1$, the autocorrelations in the differences are not significant. So, a random walk might be a good model for the differences.  -->

<!-- We can add an additional terms to the random walk model using the slope estimate from Holt-Winters. Assume the next term in the time series can be modeled as the previous term plus an estimated slope plus a white noise component. This gives us the first equation below. We will use the Holt-Winters update equation to estimate the slope. This is the second equation below. -->

<!-- $$ -->
<!-- \begin{cases} -->
<!-- ~~~x_t     = x_{t-1} + b_{t-1} + w_t \\ -->
<!--    b_{t-1} = \beta \left( x_{t-1} - x_{t-2} \right) + (1-\beta) b_{t-2}  -->
<!-- \end{cases} -->
<!-- $$ -->

<!-- We can constrain the Holt-Winters parameters $\alpha=1$ and $\gamma=0$, and fit $\beta$. (The time series of stock prices does not have values for weekends or holidays, so we will create a basic ts object and apply the base R HoltWinters() command.) -->

<!-- ```{r} -->
<!-- stock_ts0 <- ts(stock_ts$value) -->
<!-- model_hw <- HoltWinters(stock_ts0, alpha = 1, gamma = FALSE) -->
<!-- model_hw -->
<!-- ``` -->

<!-- This leads to the system of equations: -->

<!-- $$ -->
<!-- \begin{cases} -->
<!-- ~~~x_t     = x_{t-1} + b_{t-1} + w_t \\ -->
<!--    b_{t-1} = `r model_hw$beta |> round(3)` \left( x_{t-1} - x_{t-2} \right) + `r 1 - model_hw$beta |> round(3)` ~ b_{t-2}  -->
<!-- \end{cases} -->
<!-- $$ -->


<!-- <!-- Check Your Understanding --> -->

<!-- ::: {.callout-tip icon=false title="Check Your Understanding"} -->

<!-- Start with the fitted system of equations -->
<!-- $$ -->
<!-- \begin{cases} -->
<!-- ~~~x_t     = x_{t-1} + b_{t-1} + w_t \\ -->
<!--    b_{t-1} = `r model_hw$beta |> round(3)` \left( x_{t-1} - x_{t-2} \right) + `r 1 - model_hw$beta |> round(3)` ~ b_{t-2}  -->
<!-- \end{cases} -->
<!-- $$ -->
<!-- Show that it can be written as  -->
<!-- $$ -->
<!--   \left( 1 - \mathbf{B})^2 x_t = (1 - `r 1 - model_hw$beta |> round(3)` \mathbf{B} \right) w_t -->
<!-- $$ -->
<!-- by completing the following steps. -->

<!-- -   Write the system of equations in terms of the backward shift operator. -->
<!-- -   Solve for the term $\mathbf{B} b_t$ in the first equation and substitute the resulting expression into the second equation. -->
<!-- -   Combine like terms and simplify. -->

<!-- ::: -->

<!-- We will examine the residuals from the Holt-Winters model: -->
<!-- ```{r} -->
<!-- acf(resid(model_hw)) -->
<!-- ``` -->
<!-- Notice that there are no significant autocorrelations, so we conclude that the Holt-Winters model yields reasonable estimates of the time series. -->






## Class Activity: Random Walk with Drift (15 min)

<!-- This should be a stock with a fairly linear trend over time. -->
<!-- This should be a stock with a fairly linear trend over time. -->
<!-- This should be a stock with a fairly linear trend over time. -->
<!-- This should be a stock with a fairly linear trend over time. -->
<!-- This should be a stock with a fairly linear trend over time. -->
<!-- This should be a stock with a fairly linear trend over time. -->

We will now consider the daily closing price of Abercrombie & Fitch stock (Symbol = ANF). Here is a time series plot of the closing stock prices.

```{r}
#| code-fold: true
#| code-summary: "Show the code"

# Set symbol and date range
symbol <- "ANF"
date_start <- "2023-05-01"
date_end <- "2024-02-20"

# Fetch stock prices
df_stock <- tq_get(symbol, from = date_start, to = date_end, get = "stock.prices")

# Transform data into tibble
df_tsibble <- df_stock |>
  mutate(
    dates = date, 
    value = close
  ) |>
  dplyr::select(dates, value) |>
  as_tibble() |> 
  arrange(dates) |>
  as_tsibble(index = dates, key = NULL)

# Generate time series plot using plot_ly
plot_ly(df_tsibble, x = ~dates, y = ~value, type = 'scatter', mode = 'lines') |>
  layout(
    xaxis = list(title = "Date"),
    yaxis = list(title = "Value"),
    title = paste0("Time Plot of ", symbol, " Daily Closing Price (", format(ymd(date_start), "%d %b %Y"), " - ", format(ymd(date_end), "%d %b %Y"),")")
  )
```

We now generate a time plot and a correlogram of the differences. (No stock prices are recorded on weekends or holidays. Due to the gaps in the data, we will use the base R `acf` command, rather than the feasts `ACF` command.)

```{r}
#| code-fold: true
#| code-summary: "Show the code"

df_tsibble$diff = df_tsibble$value - lag(df_tsibble$value) 
df_tsibble |>
  na.omit() |>
  autoplot(.vars = diff) + 
    labs(
      title = "Time Plot of Differences in Daily ANF Stock Prices",
      subtitle = 
        paste0(
          format(ymd(date_start), "%d %b %Y"),
            " - ",
          format(ymd(date_end), "%d %b %Y")
          ),
      x = "Date",
      y = "Difference",
    ) +
    theme(
      plot.title = element_text(hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5)
    )
acf(df_tsibble$diff |> na.omit(), main = paste("ACF of First Difference of", symbol, "Stock Prices"))
```

There is no significant autocorrelation in the differences. They appear to be modeled reasonably well by white noise. 

We now compute the mean and standard deviation of the differences.

```{r}
mean_diff <- df_tsibble$diff |> mean(na.rm = TRUE)
sd_diff <- df_tsibble$diff |> sd(na.rm = TRUE)
n_diff <- df_tsibble$diff |> na.omit() |> length()
```

The mean of the differences is `r mean_diff |> round(3)`. The standard deviation of the differences is `r sd_diff |> round(3)`. There are `r n_diff` differences.

We can use the t-distribution to create a 95% confidence interval for the drift parameter. 
The critical $t$ value is given by
`qt(0.975, df = `r n_diff` - 1)`, yielding a value of $t^*_{0.975} = `r qt(0.975, df = n_diff - 1) |> round(3)`$.

So, our 95% confidence interval is computed as:
$$
  \left(
    \bar x - t^*_{0.975} \cdot \frac{s}{\sqrt{n}}
    , ~
    \bar x + t^*_{0.975} \cdot \frac{s}{\sqrt{n}}
  \right)
$$
$$
  \left(
    `r mean_diff |> round(3)` - `r qt(0.975, df = n_diff - 1) |> round(3)` \cdot  \frac{`r sd_diff |> round(3)`}{\sqrt{`r n_diff`}}
    , ~
    `r mean_diff |> round(3)` + `r qt(0.975, df = n_diff - 1) |> round(3)` \cdot  \frac{`r sd_diff |> round(3)`}{\sqrt{`r n_diff`}}
  \right)
$$
$$
  (`r (mean_diff - qt(0.975, df = n_diff - 1) * sd_diff / sqrt(n_diff)) |> round(3)`
  , ~
  `r (mean_diff + qt(0.975, df = n_diff - 1) * sd_diff / sqrt(n_diff)) |> round(3)`)
$$

This confidence interval does not contain 0, so we conclude that there is evidence of a positive drift in the price of Abercrombie & Fitch stock over this period.



<!-- Check Your Understanding -->

::: {.callout-tip icon=false title="Check Your Understanding"}

Use the McDonald's stock price data to do the following.

-   What is the estimate of the drift parameter?
-   What is the standard deviation of the differences?
-   What is the 95% confidence interval for the drift parameter?
-   Is there evidence to suggest that the time series can be modeled as a random walk with drift? Why or why not?

:::






## Homework Preview (5 min)

-   Review upcoming homework assignment
-   Clarify questions


::: {.callout-note icon=false icon=false}

## Download Homework

<a href="https://byuistats.github.io/timeseries/homework/homework_4_2.qmd" download="homework_4_2.qmd"> homework_4_2.qmd </a>

:::





<a href="javascript:showhide('Solutions2')"
style="font-size:.8em;">Backward Shift Operator</a>
  
::: {#Solutions2 style="display:none;"}

<center>
```{r, results='asis'}
#| echo: false

cat( paste( paste0("$x_{", d_operator$t, "} = ", d_operator$x, "$"), collapse = ",$~$ " ) ) 
```
</center>

Check Your Understanding Solutions:

-   $\mathbf{B} x_8 = x_7 = `r ts_val(7)`$
-   $\mathbf{B}^5 x_8 = x_3 = `r ts_val(3)`$
-   $(\mathbf{B}^5 - \mathbf{B} ) x_8 = x_3 - x_7 = `r ts_val(3)` - `r ts_val(7)` = `r ts_val(4) - ts_val(7)`$
-   $( \mathbf{B}^2 - 6 \mathbf{B} + 9 ) x_8 = x_6 - 6 (x_7) + 9 (x_{t}) = `r ts_val(6)` - 6 (`r ts_val(7)`) + 9 (`r ts_val(8)`) = `r ts_val(6) - 6 * ts_val(7) + 9 * ts_val(8)`$
-   $( (\mathbf{B} - 6) \mathbf{B} + 9 ) x_8 = `r ts_val(6) - 6 * ts_val(7) + 9 * ts_val(8)`$
-   $( \mathbf{B} - 3 )^2 x_8 = `r ts_val(6) - 6 * ts_val(7) + 9 * ts_val(8)`$
-   $( 1 - \frac{1}{2} B - \frac{1}{4} B^2 - \frac{1}{8} B^3 ) x_8 = x_8 - \frac{1}{2} x_7 - \frac{1}{4} x_6 - \frac{1}{8} x_5 = `r ts_val(8)` - \frac{1}{2} (`r ts_val(7)`) - \frac{1}{4} (`r ts_val(6)`) - \frac{1}{8} (`r ts_val(5)`) = `r ts_val(8) - ts_val(7) / 2 - ts_val(6) / 4 - ts_val(5) / 8`$

:::




<a href="javascript:showhide('Solutions3')"
style="font-size:.8em;">Difference Operator</a>
  
::: {#Solutions3 style="display:none;"}

<center>
```{r, results='asis'}
#| echo: false

cat( paste( paste0("$x_{", d_operator$t, "} = ", d_operator$x, "$"), collapse = ",$~$ " ) ) 
```
</center>

Check Your Understanding Solutions:

<!-- Beginning of two columns -->
::: columns
::: {.column width="45%"}

-   The values of the first differences are:
$$
\nabla x_t = 
\{
`r ts_val(2) - ts_val(1)`,
~~
`r ts_val(3) - ts_val(2)`,
~~
`r ts_val(4) - ts_val(3)`,
~~
`r ts_val(5) - ts_val(4)`,
~~
`r ts_val(6) - ts_val(5)`,
~~
`r ts_val(7) - ts_val(6)`,
~~
`r ts_val(8) - ts_val(7)`\}$$

-   The values of the second differences are:
$$
\nabla^2 x_t =
\{
`r (ts_val(3) - ts_val(2)) - (ts_val(2) - ts_val(1))`,
~~
`r (ts_val(4) - ts_val(3)) - (ts_val(3) - ts_val(2))`,
~~
`r (ts_val(5) - ts_val(4)) - (ts_val(4) - ts_val(3))`,
~~
`r (ts_val(6) - ts_val(5)) - (ts_val(5) - ts_val(4))`,
~~
`r (ts_val(7) - ts_val(6)) - (ts_val(6) - ts_val(5))`,
~~
`r (ts_val(8) - ts_val(7)) - (ts_val(7) - ts_val(6))`\}$$

- By substitution, we get:

\begin{align*}
(1-\mathbf{B} )^2 x_8 
&= (1-\mathbf{B} ) \left[ (1-\mathbf{B} ) x_8 \right] \\
&= (1-\mathbf{B} ) ( x_8 - x_7 ) \\
&= ( x_8 - x_7 ) - \mathbf{B} ( x_8 - x_7 ) \\
&= ( x_8 - x_7 ) - ( x_7 - x_6 ) \\
&= ( \nabla x_8 ) - ( \nabla x_7 ) \\
&= \nabla ( \nabla x_8 ) \\
&= \nabla^2 x_8 
\end{align*}


$$
  \nabla^2 x_8 
    = \nabla \left( \nabla x_8 \right)
    = ( x_8 - x_7 ) - ( x_7 - x_6 ) 
    = (`r ts_val(8)` - `r ts_val(7)`) - (`r ts_val(7)` - `r ts_val(6)`) \\
    = `r (ts_val(8) - ts_val(7)) - (ts_val(7) - ts_val(6))`
$$
This is the value of the last term in the sequence of second differences.


:::

::: {.column width="10%"}
<!-- empty column to create gap -->
:::

::: {.column width="45%"}

```{r}
#| echo: false

d_operator |>
  dplyr::select(-diff) |>
  mutate(diff1 = x - lag(x)) |>
  mutate(diff2 = diff1 - lag(diff1)) |>
  rename(
    "$$t$$" = t,
    "$$x_t$$" = x,
    "$$\\nabla x_t$$" = diff1,
    "$$\\nabla^2 x_t$$" = diff2
  ) |>
  replace_na_with_char("") |>
  # replace_cells_with_char(rows = 1:8, cols = 3:4) |>
  display_table("0.5in")
```


:::
:::
<!-- End of two columns -->



:::
<!-- End of solution -->




<!-- Solutions at the bottom of the page -->

<a href="javascript:showhide('Solutions5')"
style="font-size:.8em;">Computing Differences</a>

::: {#Solutions5 style="display:none;"}

#### Linear 

<!-- Beginning of two columns -->
::: columns
::: {.column width="45%"}

```{r}
#| echo: false
#| fig-asp: 1
#| fig-width: 4

linear_fcn <- function(t) { return(5 + 2.5 * t) }

df_linear <- data.frame(t = 1:9) |>
  mutate(x = linear_fcn(t)) |>
  mutate(diff1 = x - lag(x)) |>
  mutate(diff2 = diff1 - lag(diff1)) |>
  mutate(diff3 = diff2 - lag(diff2)) 

df_linear |>
  ggplot(aes(x = t, y = x)) +
  geom_function(fun = linear_fcn, color = "#D55E00", xlim=c(0.8, 9.2)) +
  geom_point() +
  scale_x_continuous(breaks=seq(1, 9, by = 1)) +
  scale_y_continuous(breaks=seq(0, 30, by = 5)) +
  coord_cartesian(ylim = c(0,30)) +
  theme_bw() +
  theme(panel.grid.minor = element_blank()) +
  labs(
    title = "Linear Function",
    subtitle = "x = 2.5 t + 5"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )
```

:::

::: {.column width="10%"}
<!-- empty column to create gap -->
:::

::: {.column width="45%"}

```{r}
#| echo: false

df_linear |>
  rename(
    "$$t$$" = t,
    "$$x_t$$" = x,
    "$$\\nabla x_t$$" = diff1,
    " " = diff2, 
    "  " = diff3
  ) |>
  replace_na_with_char("") |>
  replace_cells_with_char(rows = 1:9, cols = 4:5, new_char = "") |>
  display_table("0.5in")
```

:::
:::
<!-- End of two columns -->




 
#### Quadratic


<!-- Beginning of two columns -->
::: columns
::: {.column width="45%"}

```{r}
#| echo: false
#| fig-asp: 1
#| fig-width: 4

quadratic_fcn <- function(t) { return(2 * (t - 4.75)^2 - 13.125) }

df_quadratic <- data.frame(t = 1:9) |>
  mutate(x = quadratic_fcn(t)) |>
  mutate(diff1 = x - lag(x)) |>
  mutate(diff2 = diff1 - lag(diff1)) |>
  mutate(diff3 = diff2 - lag(diff2))

df_quadratic |>
  ggplot(aes(x = t, y = x)) +
  geom_function(fun = quadratic_fcn, color = "#D55E00", xlim=c(0.8, 9.2)) +
  geom_point() +
  scale_x_continuous(breaks=seq(1, 9, by = 1)) +
  scale_y_continuous(breaks=seq(-15, 30, by = 5)) +
  coord_cartesian(ylim = c(-14,26)) +
  theme_bw() +
  theme(panel.grid.minor = element_blank()) +
  labs(
    title = "Quadratic Function",
    subtitle = "x = 2 (t - 4.75)^2 - 13.125"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )
```

:::

::: {.column width="10%"}
<!-- empty column to create gap -->
:::

::: {.column width="45%"}

```{r}
#| echo: false

df_quadratic |>
  rename(
    "$$t$$" = t,
    "$$x_t$$" = x,
    "$$\\nabla x_t$$" = diff1,
    "$$\\nabla^2 x_t$$" = diff2,
    "  " = diff3
  ) |>
  replace_na_with_char("") |>
  replace_cells_with_char(rows = 1:9, cols = 5, new_char = "") |>
  display_table("0.5in")
```

:::
:::
<!-- End of two columns -->



#### Cubic


<!-- Beginning of two columns -->
::: columns
::: {.column width="45%"}

```{r}
#| echo: false
#| fig-asp: 1
#| fig-width: 4

cubic_fcn <- function(t) { return((t - 5.5) * (t - 2) * (t - 7) / 10) }

df_cubic <- data.frame(t = 1:9) |>
  mutate(x = cubic_fcn(t)) |>
  mutate(diff1 = x - lag(x)) |>
  mutate(diff2 = diff1 - lag(diff1)) |>
  mutate(diff3 = diff2 - lag(diff2)) 

df_cubic |>
  ggplot(aes(x = t, y = x)) +
  geom_function(fun = cubic_fcn, color = "#D55E00", xlim=c(0.8, 9.2)) +
  geom_point() +
  scale_x_continuous(breaks=seq(1, 9, by = 1)) +
  scale_y_continuous(breaks=seq(-5, 8, by = 1)) +
  theme_bw() +
  theme(panel.grid.minor = element_blank()) +
  labs(
    title = "Cubic Function",
    subtitle = "x = (t - 2)(t - 7)(t - 5.5)/10"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )
```

:::

::: {.column width="10%"}
<!-- empty column to create gap -->
:::

::: {.column width="45%"}

```{r}
#| echo: false

df_cubic |>
  rename(
    "$$t$$" = t,
    "$$x_t$$" = x,
    "$$\\nabla x_t$$" = diff1,
    "$$\\nabla^2 x_t$$" = diff2,
    "$$\\nabla^3 x_t$$" = diff3
  ) |>
  replace_na_with_char("") |>
  # replace_cells_with_char(rows = 1:9, cols = 3:5, new_char = "") |>
  display_table("0.5in")
```

:::
:::
<!-- End of two columns -->

:::
<!-- End of Solution -->







<!-- <!-- Solutions at the bottom of the page --> -->

<!-- <a href="javascript:showhide('SolutionsHW')" -->
<!-- style="font-size:.8em;">Integrated Autoregressive Model</a> -->

<!-- ::: {#SolutionsHW style="display:none;"} -->

<!-- <!-- Check Your Understanding --> -->

<!-- ::: {.callout-tip icon=false title="Check Your Understanding"} -->

<!-- Start with this fitted system of equations. Show that it can be written as  -->
<!-- $$ -->
<!--   (1 - \mathbf{B})^2 x_t = (1 - `r 1 - model_hw$beta |> round(3)`) w_t -->
<!-- $$ -->
<!-- by completing the following steps. -->

<!-- -   Write the system of equations in terms of the backward shift operator. -->
<!-- $$ -->
<!-- \begin{cases} -->
<!-- ~~~x_t     = x_{t-1} + b_{t-1} + w_t \\ -->
<!--    b_{t-1} = `r model_hw$beta |> round(3)` \left( x_{t-1} - x_{t-2} \right) + `r 1 - model_hw$beta |> round(3)` ~ b_{t-2}  -->
<!-- \end{cases} -->
<!-- $$ -->
<!-- $$ -->
<!-- \begin{cases} -->
<!-- ~~~x_t     = \mathbf{B} x_{t} + \mathbf{B} b_{t} + w_t \\ -->
<!--    \mathbf{B} b_{t} = `r model_hw$beta |> round(3)` \left( \mathbf{B} x_{t} - \mathbf{B}^2 x_{t} \right) + `r 1 - model_hw$beta |> round(3)` ~ \mathbf{B}^2 b_{t}  -->
<!-- \end{cases} -->
<!-- $$ -->

<!-- -   Solve for the term $\mathbf{B}b_t$ in the first equation and substitute the resulting expression into the second equation. -->

<!-- Solving for $\mathbf{B} b_{t}$: -->
<!-- $$ -->
<!-- \mathbf{B} b_{t} = x_t -\mathbf{B} x_{t} - w_t = (1-\mathbf{B}) x_t - w_t -->
<!-- $$ -->
<!-- First, we simplify the second equation: -->
<!-- \begin{align*} -->
<!--   \mathbf{B} b_{t}  -->
<!--     &= `r model_hw$beta |> round(3)` \left( \mathbf{B} x_{t} - \mathbf{B}^2 x_{t} \right) + `r 1 - model_hw$beta |> round(3)` ~ \mathbf{B}^2 b_{t}  \\ -->
<!--     &= `r model_hw$beta |> round(3)` \left(  1 - \mathbf{B} \right) \mathbf{B} x_{t}   + `r 1 - model_hw$beta |> round(3)` ~ \mathbf{B} \left[ \mathbf{B} b_{t} \right]   -->
<!-- \end{align*} -->
<!-- Substitutingthe expression for $\mathbf{B} b_{t}$ into the second equation: -->
<!-- \begin{align*} -->
<!--   (1-\mathbf{B}) x_t - w_t  -->
<!--     &= `r model_hw$beta |> round(3)` \left(  1 - \mathbf{B} \right) \mathbf{B} x_{t}   + `r 1 - model_hw$beta |> round(3)` ~ \mathbf{B} \left[ (1-\mathbf{B}) x_t - w_t \right]  \\ -->
<!--     &= `r model_hw$beta |> round(3)` \left(  1 - \mathbf{B} \right) \mathbf{B} x_{t}   + `r 1 - model_hw$beta |> round(3)` ~ \mathbf{B} (1-\mathbf{B}) x_t - `r 1 - model_hw$beta |> round(3)` ~ \mathbf{B} w_t \\ -->
<!--     &= \left(  1 - \mathbf{B} \right) \mathbf{B} x_{t} - `r 1 - model_hw$beta |> round(3)` ~ \mathbf{B} w_t   \\ -->
<!-- \end{align*} -->

<!-- -   Combine like terms and simplify. -->

<!-- Combining like terms: -->
<!-- \begin{align*} -->
<!--   (1-\mathbf{B}) x_t - w_t  -->
<!--     &= \left(  1 - \mathbf{B} \right) \mathbf{B} x_{t} - `r 1 - model_hw$beta |> round(3)` ~ \mathbf{B} w_t   \\ -->
<!--   (1-\mathbf{B}) x_t - \left(  1 - \mathbf{B} \right) \mathbf{B} x_{t} -->
<!--     &=  w_t - `r 1 - model_hw$beta |> round(3)` ~ \mathbf{B} w_t   \\ -->
<!--   \left( 1-\mathbf{B} \right) \left(  1 - \mathbf{B} \right) x_{t} -->
<!--     &=  \left( 1 - `r 1 - model_hw$beta |> round(3)` ~ \mathbf{B} \right) w_t   \\ -->
<!--   \left( 1-\mathbf{B} \right)^2 x_{t} -->
<!--     &=  \left( 1 - `r 1 - model_hw$beta |> round(3)` ~ \mathbf{B} \right) w_t   \\ -->
<!-- \end{align*} -->

<!-- This is a special case of the ARIMA models in Chapter 7. -->
<!-- ::: -->

<!-- ::: -->
<!-- <!-- End of solution for Holt Winters Fitting --> -->




<!-- Solutions at the bottom of the page -->

<a href="javascript:showhide('Solutions4')"
style="font-size:.8em;">McDonald's Drift Parameter</a>

::: {#Solutions4 style="display:none;"}

Solutions to Class Activity

<!-- Check Your Understanding -->

::: {.callout-tip icon=false title="Check Your Understanding"}

Use the McDonald's stock price data to do the following.

-   What is the estimate of the drift parameter?

As seen above, there is no significant autocorrelation in the differences of the McDonald's stock prices. They appear to be modeled well by white noise. 

We now compute the mean and standard deviation of the differences.

```{r}
#| echo: false

mean_diff <- stock_ts$diff |> mean(na.rm = TRUE)
sd_diff <- stock_ts$diff |> sd(na.rm = TRUE)
n_diff <- stock_ts$diff |> na.omit() |> length()
```

```{r}
mean(stock_ts$diff, na.rm = TRUE)
```

-   What is the standard deviation of the differences?

```{r}
sd(stock_ts$diff, na.rm = TRUE)
```

-   What is the 95% confidence interval for the drift parameter?

The number of differences, $n$, is given by:

```{r}
length(stock_ts$diff |> na.omit())
```

The mean of the differences is `r mean_diff |> round(3)`. The standard deviation of the differences is `r sd_diff |> round(3)`. There are `r n_diff` differences.

We can use the t-distribution to create a 95% confidence interval for the drift parameter. 
The critical $t$ value is given by
`qt(0.975, df = `r n_diff` - 1)`, yielding a value of $t^*_{0.975} = `r qt(0.975, df = n_diff - 1) |> round(3)`$.

So, our 95% confidence interval is computed as:
$$
  \left(
    \bar x - t^*_{0.975} \cdot \frac{s}{\sqrt{n}}
    , ~
    \bar x + t^*_{0.975} \cdot \frac{s}{\sqrt{n}}
  \right)
$$
$$
  \left(
    `r mean_diff |> round(3)` - `r qt(0.975, df = n_diff - 1) |> round(3)` \cdot  \frac{`r sd_diff |> round(3)`}{\sqrt{`r n_diff`}}
    , ~
    `r mean_diff |> round(3)` + `r qt(0.975, df = n_diff - 1) |> round(3)` \cdot  \frac{`r sd_diff |> round(3)`}{\sqrt{`r n_diff`}}
  \right)
$$
$$
  (`r (mean_diff - qt(0.975, df = n_diff - 1) * sd_diff / sqrt(n_diff)) |> round(3)`
  , ~
  `r (mean_diff + qt(0.975, df = n_diff - 1) * sd_diff / sqrt(n_diff)) |> round(3)`)
$$


-   Is there evidence to suggest that the time series can be modeled as a random walk with drift? Why or why not?

This confidence interval contains 0, so we cannot conclude that there is evidence of a positive drift in the price of McDonald's stock over this period.

:::


:::





