---
title: "Autoregressive Moving Average (ARMA) Models"
subtitle: "Chapter 6: Lesson 2"
format: html
editor: source
sidebar: false
---

```{r}
#| include: false
source("common_functions.R")
```

```{=html}
<script type="text/javascript">
 function showhide(id) {
    var e = document.getElementById(id);
    e.style.display = (e.style.display == 'block') ? 'none' : 'block';
 }
 
 function openTab(evt, tabName) {
    var i, tabcontent, tablinks;
    tabcontent = document.getElementsByClassName("tabcontent");
    for (i = 0; i < tabcontent.length; i++) {
        tabcontent[i].style.display = "none";
    }
    tablinks = document.getElementsByClassName("tablinks");
    for (i = 0; i < tablinks.length; i++) {
        tablinks[i].className = tablinks[i].className.replace(" active", "");
    }
    document.getElementById(tabName).style.display = "block";
    evt.currentTarget.className += " active";
 }    
</script>
```


## Learning Outcomes

{{< include outcomes/_chapter_6_lesson_2_outcomes.qmd >}}




## Preparation

-   Read Sections 6.5.1 and 6.6-6.7



## Learning Journal Exchange (10 min)

-   Review another student's journal

-   What would you add to your learning journal after reading another student's?

-   What would you recommend the other student add to their learning journal?

-   Sign the Learning Journal review sheet for your peer




## Class Activity: Introduction to Autoregressive Moving Average (ARMA) Models (10 min)

### Autoregressive (AR) Models

In [Chapter 4, Lesson 3](https://byuistats.github.io/timeseries/chapter_4_lesson_3.html#ARdefinition), we learned the definition of an AR model:

::: {.callout-note icon=false title="Definition of an Autoregressive (AR) Model"}

The time series $\{x_t\}$ is an **autoregressive process of order $p$**, denoted as $AR(p)$, if
$$
  x_t = \alpha_1 x_{t-1} + \alpha_2 x_{t-2} + \alpha_3 x_{t-3} + \cdots + \alpha_{p-1} x_{t-(p-1)} + \alpha_p x_{t-p} + w_t ~~~~~~~~~~~~~~~~~~~~~~~ (4.15)
$$

where $\{w_t\}$ is white noise and the $\alpha_i$ are the model parameters with $\alpha_p \ne 0$.

:::

The $AR(p)$ model can be expressed as:
$$
  \underbrace{\left( 1 - \alpha_1 \mathbf{B} - \alpha_2 \mathbf{B}^2 - \cdots - \alpha_p \mathbf{B}^p \right)}_{\theta_p \left( \mathbf{B} \right)} x_t = w_t
$$

### Moving Average (MA) Models

The definition of an $MA(q)$ model is:

::: {.callout-note icon=false title="Definition of a Moving Average (MA) Model"}

We say that a time series $\{x_t\}$ is a **moving average process of order $q$**, denoted as $MA(q)$, if each term in the time series is a linear combination of the current white noise term and the $q$ most recent past white noise terms. 

It is given as:
$$
  x_t = w_t + \beta_1 w_{t-1} + \beta_2 w_{t-2} + \beta_3 w_{t-3} + \cdots + \beta_{q-1} w_{t-(q-1)} + \beta_q w_{t-q}
$$

where $\{w_t\}$ is white noise with zero mean and variance $\sigma_w^2$, and the $\beta_i$ are the model parameters with $\beta_q \ne 0$.

:::

Written in terms of the backward shift operator, we have 
$$
  x_t = \underbrace{\left( 1 + \beta_1 \mathbf{B} + \beta_2 \mathbf{B}^2  + \beta_3 \mathbf{B}^3  + \cdots + \beta_{q-1} \mathbf{B}^{q-1}  + \beta_q \mathbf{B}^{q} \right)}_{\phi_q(\mathbf{B})} w_t
$$

Putting the $AR$ and $MA$ models together, we get the $ARMA$ model. 

::: {.callout-note icon=false title="Definition of an Autogregressive Moving Average (ARMA) Model"}

A time series $\{ x_t \}$ follows an **autoregressive moving average (ARMA) model** of order $(p, q)$, which we write as $ARMA(p,q)$, if it can be written as:

$$
  x_t = 
    \underbrace{
      \alpha_1 x_{t-1} + \alpha_2 x_{t-2} 
      + \alpha_3 x_{t-3} 
      + \cdots 
      % + \alpha_{p-1} x_{t-(p-1)} + 
      \alpha_p x_{t-p}
    }_{
      AR(p) ~ \text{model}
    } 
    + \underbrace{
      w_t + \beta_1 w_{t-1} + \beta_2 w_{t-2} 
      + \beta_3 w_{t-3} 
      + \cdots 
      % + \beta_{q-1} w_{t-(q-1)} 
      + \beta_q w_{t-q}
    }_{
      MA(q) ~ \text{model}
    } 
$$
where $\{w_t\}$ is a white noise process.

:::

We can write this as:
$$
  \theta_p \left( \mathbf{B} \right) x_t
  =
  \phi_q \left( \mathbf{B} \right) w_t
$$

::: {.callout-caution icon=false title="Facts about ARMA Processes"}

The following facts are true for $ARMA(p,q)$ processes:

-   The ARMA process is stationary if all the roots of $\theta_p \left( \mathbf{B} \right)$ are greater than 1 in absolute value.
-   The ARMA process is invertible if all the roots of $\phi_q \left( \mathbf{B} \right)$ are greater than 1 in absolute value.
-   The special case $ARMA(p,0)$ is the $AR(p)$ model.
-   The special case $ARMA(0,q)$ is the $MA(q)$ model.
-   An $ARMA$ model will usually require fewer parameters than a single $MA$ or $AR$ model. This is called *parameter parsimony*.
-   If $\theta$ and $\phi$ have a common factor, a stationary model can be simplified. This is called *parameter redundancy*.
As an example, the model 
$$
  \left( 1 - \frac{1}{2} \mathbf{B} \right)\left( 1 - \frac{1}{3} \mathbf{B} \right) x_t 
  = 
  \left( 1-\frac{1}{2} \mathbf{B} \right)\left( 1 - \frac{1}{4} \mathbf{B} \right) w_t
$$ 
is the same as the model
$$
  \left( 1 - \frac{1}{3} \mathbf{B} \right) x_t 
  = 
  \left( 1 - \frac{1}{4} \mathbf{B} \right) w_t
$$

:::


### Comparison of AR and MA Models
::: {.callout-note icon=false title="ACF and PACF of an $AR(p)$ Process"}

We can use the pacf and acf plots to assess if an $AR(p)$ or $MA(q)$ model is appropriate. 
For an $AR(p)$ or $MA(q)$ process, we observe the following:

<center>

|      | AR(p)                  | MA(q)                  | ARMA(p,q)                |
|------|------------------------|------------------------|--------------------------|
| ACF  | Tails off              | Cuts off after lag $q$ | Tails off                |
| PACF | Cuts off after lag $p$ | Tails off              | Tails off                |

</center>

<!-- https://people.cs.pitt.edu/~milos/courses/cs3750/lectures/class16.pdf -->

:::


### Analyzing Time Series with a Regular Seasonal Pattern

Here are some steps you can use to guide your work as you analyze time series with a regular seasonal pattern. Even though these are presented linearly, you may find it helpful to iterate between some of the steps as needed.

::: {.callout-warning icon=false title="Steps in a Time Series Analysis"}

-   Create a time plot of the series
-   Determine an appropriate model for the trend and seasonality; call this Model 1
    -   Regression
    -   Holt-Winters
    -   Additive / Multiplicative Decomposition
    -   Other Techniques
-   Use AIC/AICc/BIC (or other metrics) and your reasoning skills to choose the best model
-   Obtain the residuals
-   Check the residuals of Model 1 for evidence of autocorrelation and non-stationarity
    -   Time plot of the residuals
    -   ACF plot
    -   PACF plot
-   Fit an AR/MA/ARMA model to the (stationary) residuals; call this Model 2
-   Check the residuals of Model 2 for evidence of autocorrelation and normality
    -   ACF plot
    -   PACF plot
    -   Histogram
-   Forecast the trend and seasonality of the time series using Model 1
-   Forecast the residuals of Model 1 using Model 2
-   Add the two forecasts together
-   Plot and summarize the resulting forecast
-   Refine the models above as needed

:::

## Class Activity: Model for the Residuals from the Rexburg Weather Model (15 min)

### Review

We now review the model we built in [Chapter 5 Lesson 3](https://byuistats.github.io/timeseries/chapter_5_lesson_3.html#weather) for the monthly average of the daily high temperature in Rexburg, Idaho.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false
#| label: fig-WeatherTimePlot
#| fig-cap: "Time plot of the monthly mean daily high temperatures in Rexburg Idaho, in degrees Fahrenheit"

weather_df <- rio::import("https://byuistats.github.io/timeseries/data/rexburg_weather_monthly.csv") |>
  mutate(dates = my(date_text)) |>
  filter(dates >= my("1/2008") & dates <= my("12/2023")) |>
  rename(x = avg_daily_high_temp) |>
  mutate(TIME = 1:n()) |>
  mutate(
    cos1 = cos(2 * pi * 1 * TIME/12),
    cos2 = cos(2 * pi * 2 * TIME/12),
    cos3 = cos(2 * pi * 3 * TIME/12),
    cos4 = cos(2 * pi * 4 * TIME/12),
    cos5 = cos(2 * pi * 5 * TIME/12),
    cos6 = cos(2 * pi * 6 * TIME/12),
    sin1 = sin(2 * pi * 1 * TIME/12),
    sin2 = sin(2 * pi * 2 * TIME/12),
    sin3 = sin(2 * pi * 3 * TIME/12),
    sin4 = sin(2 * pi * 4 * TIME/12),
    sin5 = sin(2 * pi * 5 * TIME/12),
    sin6 = sin(2 * pi * 6 * TIME/12)) |>
  mutate(zTIME = (TIME - mean(TIME)) / sd(TIME)) |>
  as_tsibble(index = TIME)

weather_df |>
  as_tsibble(index = dates) |>
  autoplot(.vars = x) +
  geom_smooth(method = "lm", se = FALSE, color = "#F0E442") +
    labs(
      x = "Month",
      y = "Mean Daily High Temperature (Fahrenheit)",
      title = "Time Plot of Mean Daily Rexburg High Temperature by Month",
      subtitle = paste0("(", format(weather_df$dates %>% head(1), "%b %Y"), endash, format(weather_df$dates %>% tail(1), "%b %Y"), ")")
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5)
    )
```


We chose the "Reduced Linear 5" model. For convenience, we reprint the coefficients here.

```{r}
#| label: weather22
#| code-fold: true
#| code-summary: "Show the code"

reduced5_linear_lm <- weather_df |>
  model(reduced_linear_5  = TSLM(x ~ zTIME + sin1 + cos1 + sin2 + cos2))

reduced5_linear_lm |>
  tidy() |>
  mutate(sig = p.value < 0.05)

r5lin_coef_unrounded <- reduced5_linear_lm |>
  tidy() |>
  select(term, estimate, std.error)

r5lin_coef <- r5lin_coef_unrounded |>
  round_df(3)

stats_unrounded <- weather_df |>
  as_tibble() |>
  dplyr::select(TIME) |>
  summarize(mean = mean(TIME), sd = sd(TIME))

stats <- stats_unrounded |>
  round_df(3)
```

The fitted model is:

\begin{align*}
  x_t 
      &= \hat \beta_0 + \hat \beta_1 \left( \frac{t - \bar t}{s_t} \right)  \\
      & ~~~~~~~~~~ + \hat \beta_2 \sin \left( \frac{2\pi \cdot 1 t}{12} \right) 
            + \hat \beta_3 \cos \left( \frac{2\pi \cdot 1 t}{12} \right) \\
      & ~~~~~~~~~~ + \hat \beta_4 \sin \left( \frac{2\pi \cdot 2 t}{12} \right) 
            + \hat \beta_5 \cos \left( \frac{2\pi \cdot 2 t}{12} \right) 
      \\
      &= `r r5lin_coef$estimate[1]` 
            + `r r5lin_coef$estimate[2]` \left( \frac{t - `r stats$mean`}{`r stats$sd`} \right) \\ 
      & ~~~~~~~~~~~~~~~~~ + (`r r5lin_coef$estimate[3]`) \sin \left( \frac{2\pi \cdot 1 t}{12} \right) 
            + (`r r5lin_coef$estimate[4]`) \cos \left( \frac{2\pi \cdot 1 t}{12} \right) \\
      & ~~~~~~~~~~~~~~~~~ + `r r5lin_coef$estimate[5]` \sin \left( \frac{2\pi \cdot 2 t}{12} \right) 
            + (`r r5lin_coef$estimate[6]`) \cos \left( \frac{2\pi \cdot 2 t}{12} \right) 
      \\
\end{align*}

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| label: fig-WeatherTimePlotWithForecast
#| fig-cap: "Time plot of the monthly mean daily high temperatures in Rexburg Idaho, in degrees Fahrenheit; the fitted values from the regression model are given in orange"

num_months <- weather_df |> 
  as_tibble() |> 
  dplyr::select(TIME) |> 
  tail(1) |> 
  pull()

df <- tibble( TIME = seq(1, num_months, 0.01) ) |>
  mutate(
    cos1 = cos(2 * pi * 1 * TIME/12),
    cos2 = cos(2 * pi * 2 * TIME/12),
    cos3 = cos(2 * pi * 3 * TIME/12),
    cos4 = cos(2 * pi * 4 * TIME/12),
    cos5 = cos(2 * pi * 5 * TIME/12),
    cos6 = cos(2 * pi * 6 * TIME/12),
    sin1 = sin(2 * pi * 1 * TIME/12),
    sin2 = sin(2 * pi * 2 * TIME/12),
    sin3 = sin(2 * pi * 3 * TIME/12),
    sin4 = sin(2 * pi * 4 * TIME/12),
    sin5 = sin(2 * pi * 5 * TIME/12),
    sin6 = sin(2 * pi * 6 * TIME/12)) |>
  mutate(zTIME = (TIME - mean(TIME)) / sd(TIME)) |>
  as_tsibble(index = TIME)

linear5_ts <- reduced5_linear_lm |>
  forecast(df) |>
  as_tibble() |>
  dplyr::select(TIME, .mean) |>
  rename(value = .mean) |>
  mutate(Model = "Linear 5")

data_ts <- weather_df |> 
  as_tibble() |>
  rename(value = x) |>
  mutate(Model = "Data") |>
  dplyr::select(TIME, value, Model)

combined_ts <- bind_rows(data_ts, linear5_ts) 
point_ts <- combined_ts |> filter(TIME == floor(TIME))

okabe_ito_colors <- c("#000000", "#E69F00")

combined_ts |>
  ggplot(aes(x = TIME, y = value, color = Model)) +
  geom_line() +
  geom_point(data = point_ts, alpha = 0.5) +
  labs(
      x = "Month Number",
      y = "Temperature (Fahrenheit)",
      title = "Monthly Average of Daily High Temperatures in Rexburg",
      subtitle = paste0("(", format(weather_df$dates %>% head(1), "%b %Y"), endash, format(weather_df$dates %>% tail(1), "%b %Y"), ")")
  ) +    
  scale_color_manual(
    values = okabe_ito_colors[1:nrow(combined_ts |> as_tibble() |> select(Model) |> unique())], 
    name = ""
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    legend.position = "top", # Position the legend at the top
    legend.direction = "horizontal" # Set the legend direction to horizontal
  )
```


### Fitting an ARMA(p,q) Model

First, we create an acf and pacf plot of the residuals from the model above.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| label: fig-acfResidWeather
#| fig-cap: "ACF Plot of the Residuals from the Rexburg Weather Model"
#| warning: false

reduced5_linear_lm |>
  residuals() |>
  ACF() |> 
  autoplot(var = .resid)
```


```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| label: fig-pacfResidWeather
#| fig-cap: "PACF Plot of the Residuals from the Rexburg Weather Model"
#| warning: false

reduced5_linear_lm |> 
  residuals() |>
  PACF() |>
  autoplot(var = .resid)
```


<!-- Check Your Understanding -->

::: {.callout-tip icon=false title="Check Your Understanding"}

-   What ARMA model does the combined information in the acf and pacf plots suggest would be appropriate for the residuals of the Rexburg weather model?

:::

Here are summaries of some ARMA models that could be constructed.

```{r}
#| output: false

model_resid <- reduced5_linear_lm |>
  residuals() |>
  select(-.model) |>
  model(
    auto = ARIMA(.resid ~ 1 + pdq(0:2,0,0:2) + PDQ(0, 0, 0)),
    a000 = ARIMA(.resid ~ 1 + pdq(0,0,0) + PDQ(0, 0, 0)),
    a001 = ARIMA(.resid ~ 1 + pdq(0,0,1) + PDQ(0, 0, 0)),
    a002 = ARIMA(.resid ~ 1 + pdq(0,0,2) + PDQ(0, 0, 0)),
    a100 = ARIMA(.resid ~ 1 + pdq(1,0,0) + PDQ(0, 0, 0)),
    a101 = ARIMA(.resid ~ 1 + pdq(1,0,1) + PDQ(0, 0, 0)),
    a102 = ARIMA(.resid ~ 1 + pdq(1,0,2) + PDQ(0, 0, 0)),
    a200 = ARIMA(.resid ~ 1 + pdq(2,0,0) + PDQ(0, 0, 0)),
    a201 = ARIMA(.resid ~ 1 + pdq(2,0,1) + PDQ(0, 0, 0)),
    a202 = ARIMA(.resid ~ 1 + pdq(2,0,2) + PDQ(0, 0, 0))) 

model_resid |>
  glance() 
```


```{r}
#| echo: false

# model_best |>
#   residuals() |>
#   ACF() |>
#   autoplot()


# new_data <- tibble(
#   date = seq(
#     max(cbe_ts$date) + months(1),
#       max(cbe_ts$date) + months(36),
#       by = "1 months"),
#   month = tsibble::yearmonth(date),  
#   time = seq(nrow(cbe_ts), length = 36),
#   imth = rep(1:12, 3)) |>
#   as_tsibble(index = month)
# 
# plot_dat <- new_data |>
#   mutate(
#     residuals = forecast(
#       model_best,
#       h = "3 years") |>
#       pull(.mean),
#     expected_log = forecast(
#       elec_lm,
#       new_data = new_data) |>
#       pull(.mean),
#     expected = exp(expected_log + residuals)  
#   )
# 
# ggplot() +
#   geom_line(data = cbe_ts, aes(x = date, y = elec)) +
#   geom_line(data = plot_dat, aes(x = date, y = expected), linetype = 2)
```




```{r}
#| label: tbl-ModelComparison
#| tbl-cap: "Comparison of the AIC, AICc, and BIC values for the models fitted to the Rexburg weather data"
#| echo: false

combined_models <- glance(model_resid) |> 
  select(.model, AIC, AICc, BIC)

minimum <- combined_models |>
  reframe(
    AIC = which(min(AIC)==AIC),
    AICc = which(min(AICc)==AICc),
    BIC = which(min(BIC)==BIC)
  )

combined_models |>
  rename(Model = ".model") |>
  round_df(1) |>
  format_cells(rows = minimum$AIC, cols = 2, "bold") |>
  format_cells(rows = minimum$AICc, cols = 3, "bold") |>
  format_cells(rows = minimum$BIC, cols = 4, "bold") |>
  display_table()
```

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false

model_resid |>
  tidy() |>
  filter(.model == "a101") 
```

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false
#| label: fig-ACFofResiduals
#| fig-cap: "ACF plot of the residuals from the $ARMA(1,1)$ model"

model_resid |>
  select(a101) |>
  residuals() |>
  ACF() |>
  autoplot(var = .resid)
```

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false
#| label: fig-PACFofResiduals
#| fig-cap: "PACF plot of the residuals from the $ARMA(1,1)$ model"

model_resid |>
  select(a101) |>
  residuals() |>
  PACF() |>
  autoplot(var = .resid)
```









## Small-Group Activity: Industrial Electricity Consumption in Texas (20 min)




These data represent the amount of electricity used each month for industrial applications in Texas.

```{r}
#| code-fold: true
#| code-summary: "Show the code"

elec_ts <- rio::import("https://byuistats.github.io/timeseries/data/electricity_tx.csv") |>
  dplyr::select(-comments) |>
  mutate(month = my(month)) |>
  mutate(
    t = 1:n(),
    std_t = (t - mean(t)) / sd(t)
  ) |>
  mutate(
    cos1 = cos(2 * pi * 1 * t / 12),
    cos2 = cos(2 * pi * 2 * t / 12),
    cos3 = cos(2 * pi * 3 * t / 12),
    cos4 = cos(2 * pi * 4 * t / 12),
    cos5 = cos(2 * pi * 5 * t / 12),
    cos6 = cos(2 * pi * 6 * t / 12),
    sin1 = sin(2 * pi * 1 * t / 12),
    sin2 = sin(2 * pi * 2 * t / 12),
    sin3 = sin(2 * pi * 3 * t / 12),
    sin4 = sin(2 * pi * 4 * t / 12),
    sin5 = sin(2 * pi * 5 * t / 12)
  ) |>
  as_tsibble(index = month)

elec_plot_raw <- elec_ts |>
    autoplot(.vars = megawatthours) +
    labs(
      x = "Month",
      y = "Megawatt-hours",
      title = "Texas' Industrial Electricity Use"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5)
    )

elec_plot_log <- elec_ts |>
    autoplot(.vars = log(megawatthours)) +
    labs(
      x = "Month",
      y = "log(Megwatt-hours)",
      title = "Log of Texas' Industrial Electricity Use"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5)
    )

elec_plot_raw | elec_plot_log
```

<!-- Check Your Understanding -->

::: {.callout-tip icon=false title="Check Your Understanding"}

Use the Texas industrial electricity consumption data to do the following.

-   Select an appropriate fitted model using the AIC, AICc, or BIC critera.
-   Use the residuals to determine the appropriate correction for the data.
<!-- -   Forecast the data for the next 5 years. -->
<!-- -   Apply the appropriate correction to the forecasted values. -->
<!-- -   Plot the fitted (forecasted) values along with the time series. -->

:::
















## Homework Preview (5 min)

-   Review upcoming homework assignment
-   Clarify questions




::: {.callout-note icon=false}

## Download Homework

<a href="https://byuistats.github.io/timeseries/homework/homework_6_2.qmd" download="homework_6_2.qmd"> homework_6_2.qmd </a>

:::



