---
title: "Resources"
listing:
  contents: resources
  fields: [image, title]
  type: grid
  categories: true
---

::: supplement
::: supplement-header
**Tidyverse Conversion**
:::

::: supplement-container
### [Supplement to Introductory Time Series with R](https://byuistats.github.io/timeseries_supplement/)
:::
:::

::: {.callout-note title="Common Packages" icon="false"}
```{r message=FALSE}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, 
               tsibble, fable,
               feasts, tsibbledata,
               fable.prophet,
               patchwork,
               lubridate, 
               rio 
)
```

Full list of libraries used throughout the lessons. If you are having issues with a function not getting called correctly, read through the comments and see if anything matches, if not pull packages out of the pacman and individually load them in through library(). Some of the most likely error causeing libraries are tagged in the comments with a note, so start with those unless you have a better idea of what package might be causing the issue.
```{r message=FALSE}
pacman::p_load(
  # Interactive plots
  plotly, # Interactive visualizations, loaded before tidyverse so it overwrite dplyr::select(). Note: High conflict Potential
    
  # Core packages
  MASS, # MVNorm, loaded before tidyverse so it doesn't overwrite dplyr::select()
  tidyverse, # This will also load the dependencies; dplyr, readr, stringr, tibble, tidyr, purrr, forcats, gglot2, & lubridate
  
  # Data manipulation
  tsibble, # Tidyverse Temporal data
  tsibbledata, # Sample Tsibble datasets

  
  # Statistical modeling (GLS - Chpt 6-7) 
  nlme, # loaded before feasts to avoid ACF() conflict
  tidymodels, # for GLS, This will also load the dependencies; broom, rsample, dials, tune, infer, workflows, modeldata, workflowsets, parsnip, yardstick, & recipies. Note: High conflict Potential
  multilevelmod, # for GLS
  broom.mixed, # for GLS
  
  # TS modeling and forecasting
  fable,# Forecasting Models for Tidy Time Series, Note: High conflict Potential
  feasts, # collection of features, decomposition methods, statistical summaries and graphics for tsibble data, Loaded after nlme to avoid ACF() conflict
  fable.prophet, # Converts prophet (forecasting) package for fable workflow
  
  # Data exploration & visualization
  patchwork, # Multiple plot outputs
  ggthemes, # Plot styling
  see,  # okabeito color scheme
  ggokabeito,  # colorblind palette

  
  # Reporting & output
  kableExtra, # Create nice-looking tables from data.frames
  rio, # Easy import/export of data between R and other software
  gt, # Grammar of Tables for advanced table creation
  quarto, # For generating reports in LaTeX format

  # Additional packages
  tidyquant # Quantitative analysis tools using tidyverse principles, This will also load the dependencies; PerformanceAnalytics, xts, & zoo. Important Masks: ‘package:base’: as.Date, as.Date.numeric. Note: High conflict Potential
)
```
:::
