---
title: "Resources"
listing:
  contents: resources
  fields: [image, title]
  type: grid
  categories: true
execute:
  eval: false
---

::: supplement
::: supplement-header
**Tidyverse Conversion**
:::

::: supplement-container
### [Supplement to Introductory Time Series with R](https://byuistats.github.io/timeseries_supplement/)
:::
:::

## R Packages

These are the most common packages...

```{r message=FALSE}
#| file: "_common_packages.R"
```

...And below are all the packages used in this course (including for backend lesson content)

::: {.callout-note title=NOTE icon=false}
If you are having issues with a function not getting called correctly, read through the comments and see if anything matches your situation. If not, pull packages out of `pacman::p_load()` and individually load them in through `library()`. Some of the most likely error-causing libraries are tagged in the comments with a note, so start with those unless you have a better idea of what package might be causing the issue.
:::

```{r}
#| file: "_all_packages.R"
```

## Download Textbook

If you would like access to the GitHub repository, you can find it at https://github.com/byuistats/timeseries