---
title: "TS_HW1.2"
format: html
editor: visual
toc: true
---

## Learning Objectives

1.1 Use technical language to describe the main features of time series data

1.3 Use R to describe key features of time series data.

1.4 Plot time series data to visualize trends, seasonal patterns, and potential outliers.

Chapter 1 Outcome Map [HERE](https://docs.google.com/spreadsheets/d/1iYftJo380M2d9rV-KexgEaCjkKbqK3Ip7uEdn-Y15B4/edit?usp=sharing)

## Packages

```{r}
#| echo: false
library(rio)

```

## Data

### Google Trends Data: chocolate searches

```{r}

chocolate <- rio::import("https://byuistats.github.io/timeseries/data/chocolate.csv")

```

### Macroeconomic Data: unemployment rate

```{r}

unemp_rate <- rio::import("https://byuistats.github.io/timeseries/data/unemp_rate.csv")
```

### Financial Data: term premium (10 year)

```{r}

term_prem10 <- rio::import("https://byuistats.github.io/timeseries/data/term_prem10.csv")
```

## Data Background

The first part of any time series analysis is context. You can't analyze data without knowing what the data is measuring. Without context the most simple features of data appear obscure and inscrutable. This homework assignment will center around three data series, the US unemployment series, the term premium series, and the Google Trends series of the search term "chocolate."

### Question 1 - Context and Measurement

Please research them to understand the data collection process, unit of analysis, and meaning of each observation. Click on each of the subheadings below for a link to a source to get you started.

#### [US unemployment](https://www.bls.gov/cps/cps_htgm.htm)

::: {.callout-note title="Answer" icon="false"}
Your answer HERE
:::

#### [The Term Premium](https://www.newyorkfed.org/research/data_indicators/term-premia-tabs#/overview)

::: {.callout-note title="Answer" icon="false"}
Your answer HERE
:::

#### [Chocolate Trends](https://support.google.com/trends/answer/4365533?hl=en)

::: {.callout-note title="Answer" icon="false"}
Your answer HERE
:::

## Plots, Trends, and Seasonal Variation

### Question 2 - Statistical and Graphical Description

Please *contrast* *and compare* the trends and seasonal features of the US unemployment, term premium, and the chocolate Google search series. Please include illustrations.

```{r}

# Your code HERE
```

::: {.callout-note title="Answer" icon="false"}
Your answer HERE
:::

## Rubric

+---------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+:-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------:+
| **Criteria**                                                  | **Ratings**                                                                                                                                                                                                                                                                       | **Total Points**                                                                                                                                                                                                    |
+---------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
|                                                               | **Mastery (20)**                                                                                                                                                                                                                                                                  | **Incomplete (0)**                                                                                                                                                                                                  |
+---------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| #### **Question 1: Context and Measurement**                  | The student demonstrates a clear understanding of the context for each data series (US unemployment, term premium, and Google Trends for "chocolate"). The explanation includes details about the data collection process, unit of analysis, and the meaning of each observation. | The student fails to provide a clear understanding of the context for one or more data series.                                                                                                                      |
+---------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
|                                                               | **Mastery (20)**                                                                                                                                                                                                                                                                  | **Incomplete (0)**                                                                                                                                                                                                  |
+---------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| #### **Question 2: Technical Language**                       | The student uses appropriate technical language to describe the main features of time series data (sampling interval, time series trend, seasonal variation, cycle, deterministic and stochastic trends) and provides accurate definitions. Outcome 1.1                           | The student does not use technical language to describe the main features of time series data or does not define one or more specified terms. The student fails to present a decomposition of time series data.     |
+---------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
|                                                               | **Mastery (20)**                                                                                                                                                                                                                                                                  | **Incomplete (0)**                                                                                                                                                                                                  |
+---------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| #### **Question 2: R Programming Usage**                      | The student effectively utilizes R to describe key features of time series data. Code snippets or outputs are appropriately included to support the analysis. Outcome 1.3                                                                                                         | The student does not use R to describe key features of time series data or does not demonstrate proficiency in one or more specified R functions. The student fails to present a decomposition of time series data. |
+---------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
|                                                               | **Mastery (20)**                                                                                                                                                                                                                                                                  | **Incomplete (0)**                                                                                                                                                                                                  |
+---------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| #### **Question 2: Data Visualization**                       | The student creates clear and informative plots to visualize trends, seasonal patterns, and potential outliers for each data series. The illustrations enhance the overall understanding of the time series data. Outcome 1.4, Outcome 1.3.4                                      | The student does not include visualizations or the visualizations provided are unclear and do not effectively represent the data, including the decomposition.                                                      |
+---------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
|                                                               | **Mastery (20)**                                                                                                                                                                                                                                                                  | **Incomplete (0)**                                                                                                                                                                                                  |
+---------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| #### **Question 2: Contrast of Trends and Seasonal Features** | The student provides a thoughtful and detailed contrast of the trends and seasonal features among the US unemployment, term premium, and chocolate Google search series. Outcome 1.1, Outcome 1.3                                                                                 | The student does not provide a clear contrast of trends and seasonal features for the specified data series.                                                                                                        |
+---------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
|                                                               |                                                                                                                                                                                                                                                                                   |                                                                                                                                                                                                                     |
+---------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
