---
title: "Time Series Homework: Chapter 1 Lesson 2"
subtitle: "Please_put_your_name_here"
format: 
  html:
    embed-resources: true
    toc: true
editor: source
---

```{r}
#| message: false
#| echo: false

# Loading R packages
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, 
               tsibble, fable,
               feasts, tsibbledata,
               fable.prophet,
               patchwork,
               lubridate, 
               rio 
)
```

<details>

<summary>Use R to describe key features of time series data</summary>

-   Import CSV data and convert to tsibble format

</details>

<details>

<summary>Decompose time series into trends, seasonal variation, and residuals</summary>

-   Implement additive decomposition
-   Explain how to remove seasonal variation using an estimate for seasonal component of a time series
-   Compute the estimators of seasonal variation for an additive model
-   Calculate the random component for an additive model
-   Compute a seasonally-adjusted time series based on an additive model

</details>

<!-- Assignment Begins Here-->

## Data

```{r}
# Weather data for Rexburg
rex_temp <- rio::import("https://byuistats.github.io/timeseries/data/rexburg_weather.csv")
```

## Questions

### Question 1 - Context and Measurement (10 points)

The first part of any time series analysis is context. You cannot properly analyze data without knowing what the data is measuring. Without context, the most simple features of data can be obscure and inscrutable. This homework assignment will center around the series below.

Please research the time series. In the spaces below, give the data collection process, unit of analysis, and meaning of each observation for the series.

#### a) Rexburg, ID Daily High Temperatures

::: {.callout-note title="Answer" icon="false"}
<!-- Delete this and put your answer here. The three colons mark the end of your answer. -->
:::

### Question 2 - Key Definitions (10 points)

Define the terms "additive decomposition," "estimators of seasonal variation for an additive model," and "seasonally adjusted series."

::: {.callout-note title="Answer" icon="false"}
-   **Additive Decomposition**: <!-- Delete this and put your answer here. -->

-   **Estimators of seasonal variation for an additive model**: <!-- Delete this and put your answer here. -->

-   **Seasonally Adjusted Series**: <!-- Delete this and put your answer here. -->
:::

### Question 3 -Visualization (20 points)

Please plot the Rexburg Daily Temperature series choosing the range manually to illustrate the data in the most readable format. Use the appropriate axis labels, units, and captions.

::: {.callout-note title="Answer" icon="false"}
```{r}
# Please provide your code here

```
:::

### Question 4 Seasonally Adjusted Series - Calculation (40 points)

Without using an R built-in function, manually calculate the seasonally adjusted series for the Rexburg Daily Temperature series. Make sure that your code is clearly commented to highlight each step: calculate the centered moving average, monthly additive effect, seasonal adjusted mean, and seasonally adjusted series. Illustrate the original and the seasonally adjusted series in the same plot.

::: {.callout-note title="Answer" icon="false"}
```{r}
# Please provide your code here

```
:::

### Question 5 Seasonally Adjusted Series - Random Component (20 points)

Using your estimates from the previous question, calculate and plot the random component of the Rexburg Daily Temperature series.

::: {.callout-note title="Answer" icon="false"}
```{r}
# Please provide your code here

```
:::

### Question 6 Seasonally Adjusted Series - Analysis (20 points)

1.  Justify why we use the additive decomposition model to seasonally adjust the Rexburg Daily Temperature series.

::: {.callout-note title="Answer" icon="false"}
<!-- Delete this and put your written answer here. The three colons mark the end of your answer. -->
:::

2.  Describe any patterns you observe in the random component. Propose an explanation.

::: {.callout-note title="Answer" icon="false"}
<!-- Delete this and put your written answer here. The three colons mark the end of your answer. -->
:::

<!-- Try not to touch the table markdown below -->

## Rubric

+--------------------------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **Criteria**                                                       | **Mastery (10)**                                                                                                                                                                                                                                                                                             | **Incomplete (0)**                                                                                                                                                                                                                                                                                                                            |
+--------------------------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| #### **Question 1: Context and Measurement**                       | The student thoroughly researches the data collection process, unit of analysis, and meaning of each observation for both the requested time series. Clear and comprehensive explanations are provided.                                                                                                      | The student does not adequately research or provide information on the data collection process, unit of analysis, and meaning of each observation for the specified series.                                                                                                                                                                   |
+--------------------------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
|                                                                    | **Mastery (10)**                                                                                                                                                                                                                                                                                             | **Incomplete (0)**                                                                                                                                                                                                                                                                                                                            |
+--------------------------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| #### **Question 2: Definitions**                                   | The student correctly defined each of the terms and included mathematical expressions or illustration if available in the text or the Time Series Notebook                                                                                                                                                   | The student did not provide a correct definition for one or more of the terms.                                                                                                                                                                                                                                                                |
+--------------------------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
|                                                                    | **Mastery (20)**                                                                                                                                                                                                                                                                                             | **Incomplete (0)**                                                                                                                                                                                                                                                                                                                            |
+--------------------------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| #### **Question 3: Visualization**                                 | Chooses a reasonable manual range for the Rexburg Daily Temperature series, providing a readable plot that captures the essential data trends. Creates a plot with accurate and clear axis labels, appropriate units, and a caption that enhances the understanding of the Rexburg Daily Temperature series. | Attempts manual range selection, but with significant issues impacting the readability of the plot. The chosen range may obscure important data trends, demonstrating a limited understanding of graphical representation.Fails to include, axis labels, units, or captions, leaving the visual representation and interpretation incomplete. |
+--------------------------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
|                                                                    | **Mastery (20)**                                                                                                                                                                                                                                                                                             | **Incomplete (0)**                                                                                                                                                                                                                                                                                                                            |
+--------------------------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| #### **Question 4: Seasonally Adjusted Series - Calculation**      | The student correctly calculates every step to seasonally adjust the time series. The code is easy to read and well commented. The illustration shows the original and seasonally series in one plot.                                                                                                        | The student incorrectly calculates one or more steps to seasonally adjust the time series. The code is difficult to read and poorly commented. The illustration fails show one or more components, labels, or captions.                                                                                                                       |
+--------------------------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
|                                                                    | **Mastery (20)**                                                                                                                                                                                                                                                                                             | **Incomplete (0)**                                                                                                                                                                                                                                                                                                                            |
+--------------------------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| #### **Question 5: Seasonally Adjusted Series - Random Component** | Correctly calculates the random component of the Rexburg Daily Temperature series. The code is clear, well-commented, and easy to understand. Creates a visually appealing plot that effectively illustrates the random component, including accurate axis labels, units, and a caption.                     | Attempts to calculate the random component, but with significant inaccuracies in the methodology. The code is unclear or poorly commented. The plot may lack clarity in illustrating the random component, with issues in axis labels, units, or caption.                                                                                     |
+--------------------------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
|                                                                    | **Mastery (10)**                                                                                                                                                                                                                                                                                             | **Incomplete (0)**                                                                                                                                                                                                                                                                                                                            |
+--------------------------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| #### **Question 6: Analysis Part 1**                               | Offers a solid explanation for choosing the additive decomposition model, demonstrating a good understanding of the data characteristics and the rationale behind the selection. The explanation is clear and logically presented.                                                                           | Attempts to justify the use of the additive decomposition model but with significant inaccuracies or unclear reasoning. Shows a limited understanding of the data characteristics and the rationale for choosing the additive model.                                                                                                          |
+--------------------------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
|                                                                    | **Mastery (10)**                                                                                                                                                                                                                                                                                             | **Incomplete (0)**                                                                                                                                                                                                                                                                                                                            |
+--------------------------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| #### **Question 6: Analysis Part 2**                               | Provides an insightful description of patterns observed in the random component of the Rexburg Daily Temperature series. Demonstrates a keen understanding of statistical patterns and trends. The description is clear, detailed, and supported by relevant observations.                                   | Attempts to describe patterns in the random component, but with significant inaccuracies or unclear observations. Shows a limited understanding of statistical patterns, and the description may lack coherence.                                                                                                                              |
+--------------------------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **Total Points**                                                   | **100**                                                                                                                                                                                                                                                                                                      |                                                                                                                                                                                                                                                                                                                                               |
+--------------------------------------------------------------------+--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
