---
title: "Time Series Homework: Chapter 3 Lesson 3"
subtitle: "Please_put_your_name_here"
format: 
  html:
    embed-resources: true
    toc: true
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}

# Loading R packages
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, 
               tsibble, fable,
               feasts, tsibbledata,
               fable.prophet,
               patchwork,
               lubridate, 
               rio,
               ggplot2,
               kableExtra
               )
```

## Data

```{r, message=FALSE}

unratensa <- rio::import()

avgkwhr <- rio::import()

```

## Questions

### Question 1 - Context and Measurement (10 points)

The first part of any time series analysis is context. You cannot properly analyze data without knowing what the data is measuring. Without context, the most simple features of data can be obscure and inscrutable. This homework assignment will center around the series below.

Please research the time series. In the spaces below, give the data collection process, unit of analysis, and meaning of each observation for the series.

#### a) **Average Price of Electricity per Kilowatt-Hour in U.S.: City Average**

<https://fred.stlouisfed.org/series/APU000072610>

::: {.callout-note title="Answer" icon="false"}
<!-- Delete this and put your answer here. The three colons mark the end of your answer. -->
:::

### Question 2 - Key Definitions (10 points)

Answer the prompt to the learning outcome below. Include any mathematical expressions or illustrations that may accompany the definitions and ideas if available.

::: {.callout-note title="Answer" icon="false"}
-   **Explain the purpose of the Holt-Winters parameters** $\alpha$, $\beta$, **and** $\gamma$ : <!-- Delete this and put your answer here. -->

-   **Explain the Holt-Winters forecasting equation for additive decomposition models, Equation (3.22)** <!-- Delete this and put your answer here. -->
:::

### Question 3 - US Unemployment Rate: Additive Holt-Winters Forecasting (10 points)

We have used a measure of US unemployment rates before. This series imported above has not been seasonally adjusted.

a)  Please use the Holt-Winters smoothing method to the series.

::: {.callout-note title="Answer" icon="false"}
```{r}



```
:::

b)  What parameters values did you choose for $\alpha$, $\beta$, and $\gamma$. Justify your choice.

::: {.callout-note title="Answer" icon="false"}
<!-- Delete this and put your answer here. The three colons mark the end of your answer. -->
:::

c)  Please plot the Holt-Winters forecast of the series for the next 12 months superimposed against the original series. Please see [Figure 7](https://byuistats.github.io/timeseries/chapter_3_lesson_3.html#fig-crime-hw-forecast) in Chapter 3: Lesson 3

::: {.callout-note title="Answer" icon="false"}
```{r}



```
:::

d)  Is the trend in the US unemployment series deterministic or stochastic? What is the basis of your evaluation?

::: {.callout-note title="Answer" icon="false"}
<!-- Delete this and put your answer here. The three colons mark the end of your answer. -->
:::

### Question 4 - US Average Price of Electricity: Additive Holt-Winters Forecasting (25 points)

a)  Please use the Holt-Winters smoothing method to the series.

::: {.callout-note title="Answer" icon="false"}
```{r}



```
:::

b)  What parameters values did you choose for $\alpha$, $\beta$, and $\gamma$. Justify your choice.

::: {.callout-note title="Answer" icon="false"}
<!-- Delete this and put your answer here. The three colons mark the end of your answer. -->
:::

c)  Please plot the Holt-Winters forecast of the series for the next 12 months superimposed against the original series. Please see [Figure 7](https://byuistats.github.io/timeseries/chapter_3_lesson_3.html#fig-crime-hw-forecast) in Chapter 3: Lesson 3

::: {.callout-note title="Answer" icon="false"}
```{r}



```
:::

d)  Is the trend in the US unemployment series deterministic or stochastic? What is the basis of your evaluation?

::: {.callout-note title="Answer" icon="false"}
<!-- Delete this and put your answer here. The three colons mark the end of your answer. -->
:::

### Question 5 - Additive Holt-Winters Forecasting: Evaluation (45 points)

You used the same algorithm to forecast the US Unemployment Rate and the Energy Price series. However, underneath, the processes that generated the two series are very different. Whether a series has a deterministic or stochastic trend has implications for forecasting. The parameter values used in the algorithm and their justification matters for the result. Please contrast the complexity or ease of modeling, and the uncertainty of the forecast between series with deterministic and stochastic trends.

a)  What is the reason behind the difference in the erase of modeling between the US Unemployment Rate and the Energy Price series?

::: {.callout-note title="Answer" icon="false"}
<!-- Delete this and put your answer here. The three colons mark the end of your answer. -->
:::

b)  Based on your forecast of the US Unemployment Rate and the Energy Price series, which are you the most confident with? Why?

::: {.callout-note title="Answer" icon="false"}
<!-- Delete this and put your answer here. The three colons mark the end of your answer. -->
:::

### Rubric
