---
title: "Time Series Homework: Chapter 1 Lesson 3"
subtitle: "Please_put_your_name_here"
format: 
  html:
    embed-resources: true
    toc: true
editor: source
---

```{r}
#| message: false
#| echo: false

# Loading R packages
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, 
               tsibble, fable,
               feasts, tsibbledata,
               fable.prophet,
               patchwork,
               lubridate, 
               rio 
)
```

<details>

<summary>Decompose time series into trends, seasonal variation, and residuals</summary>
-   Define smoothing
-   Compute the centered moving average for a time series
-   Estimate the trend component using moving averages

</details>


<details>

<summary>Plot time series data to visualize trends, seasonal patterns, and potential outliers</summary>
-   Plot the estimated trend of a time series using a moving average
-   Make box plots to examine seasonality
-   Interpret the trend and seasonal pattern observed in a time series

</details>


<!-- Assignment Begins Here-->

## Data

```{r}
# Macroeconomic Data: unemployment rate
unemp_rate <- rio::import("https://byuistats.github.io/timeseries/data/unemp_rate.csv")

```

## Questions

### Question 1 - Context and Measurement (10 points)

The first part of any time series analysis is context. You cannot properly analyze data without knowing what the data is measuring. Without context, the most simple features of data can be obscure and inscrutable. This homework assignment will center around the US unemployment series.

Please research the time series. The subheadings below has a link to a source to help you get started. In the spaces below, give the data collection process, unit of analysis, and meaning of each observation for the US unemployment time series.

#### a) US unemployment

<https://www.bls.gov/cps/cps_htgm.htm>

::: {.callout-note title="Answer" icon="false"}
<!-- Delete this and put your answer here. The three colons mark the end of your answer. -->
:::

### Question 2 - Key Definitions (10 points)

Define the terms below and include any mathematical expressions or illustrations that may accompany the definition if available.

::: {.callout-note title="Answer" icon="false"}
-   **Smoothing**: <!-- Delete this and put your answer here. -->

-   **Centered Moving Average**: <!-- Delete this and put your answer here. -->

-   **Box Plot**: <!-- Delete this and put your answer here. -->
:::

### Question 3 - Estimating the Trend: Centered Moving Average (20 points)

Please plot the US Unemployment time series and superimpose the centered moving average series $\hat{m}_{t}$ in the same graph. Don't use an R command, rather do it by coding $\hat{m}_{t}$ like in Chapter 1: Lesson 3. Use the appropriate axis labels, units, and captions.

::: {.callout-note title="Answer" icon="false"}
```{r}
# Please provide your code here

```

<!-- Delete this and put your written answer here. The three colons mark the end of your answer. -->
:::

### Question 4 - Estimating the Seasonal Effect: Side-by-Side Box Plots by Month (20 points)

Please illustrate the seasonality in the US Unemployment time series by creating a box plot. Use the appropriate axis labels, units, and captions.

::: {.callout-note title="Answer" icon="false"}
```{r}
# Please provide your code here

```

<!-- Delete this and put your written answer here. The three colons mark the end of your answer. -->
:::

### Question 5 - Estimating the Seasonal Effect: Analysis (20 points)

1.  Describe the seasonality in US Unemployment time series . Comment on the series' periods of highest and lowest unemployment. Are there any notable outliers?

::: {.callout-note title="Answer" icon="false"}
<!-- Delete this and put your written answer here. The three colons mark the end of your answer. -->
:::

2.  Please explain the patterns you found? Include information from your prior research on the series.

::: {.callout-note title="Answer" icon="false"}
<!-- Delete this and put your written answer here. The three colons mark the end of your answer. -->
:::









<!-- Try not to touch the table markdown below -->

## Rubric

+----------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **Criteria**                                 | **Mastery (10)**                                                                                                                                                                                                                                                                                | **Incomplete (0)**                                                                                                                                                                                                                                                                                  |
+----------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| #### **Question 1: Context and Measurement** | The student thoroughly researches the data collection process, unit of analysis, and meaning of each observation for the US Unemployment data. Clear and comprehensive explanations are provided.                                                                                               | The student does not adequately research or provide information on the data collection process, unit of analysis, and meaning of each observation for one or both of the specified series.                                                                                                          |
+----------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
|                                              | **Mastery (10)**                                                                                                                                                                                                                                                                                | **Incomplete (0)**                                                                                                                                                                                                                                                                                  |
+----------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| #### **Question 2: Definitions**             | The student correctly defined each of the terms and included mathematical expressions or illustrations if available in the text or the Time Series Notebook                                                                                                                                     | The student did not provide a correct definition for one or more of the terms.                                                                                                                                                                                                                      |
+----------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
|                                              | **Mastery (20)**                                                                                                                                                                                                                                                                                | **Incomplete (0)**                                                                                                                                                                                                                                                                                  |
+----------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| #### **Question 3: Centered Moving Average** | The student correctly employs the centering procedure to seasonally adjust the US unemployment series.                                                                                                                                                                                          | The student does not employ the centering procedure to seasonally adjust the US unemployment series.                                                                                                                                                                                                |
+----------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
|                                              | **Mastery (20)**                                                                                                                                                                                                                                                                                | **Incomplete (0)**                                                                                                                                                                                                                                                                                  |
+----------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| #### **Question 4: Box Plot**                | Creates a clear and well-constructed box plot that effectively illustrates seasonality in the US Unemployment time series. Labels both the x and y-axes, including appropriate units. Ensures clarity and accuracy in axis labeling, contributing to the overall understanding of the box plot. | There are mistakes in the plot. Fails to include any axis labels or units, resulting in an incomplete representation that lacks essential context.                                                                                                                                                  |
+----------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
|                                              | **Mastery (20)**                                                                                                                                                                                                                                                                                | **Incomplete (0)**                                                                                                                                                                                                                                                                                  |
+----------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| #### **Question 5: Seasonality Analysis**    | Provides an accurate description of the seasonality in the US Unemployment time series, capturing key patterns and trends effectively. Shows a good understanding of the recurring cycles.                                                                                                      | Attempts to identify peaks and troughs but with significant inaccuracies or lack of clarity in the commentary. Shows a limited understanding of the variations. Fails to identify or comment on notable outliers, providing no insight into unusual data points in the US Unemployment time series. |
+----------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **Total Points**                             | **80**                                                                                                                                                                                                                                                                                          |                                                                                                                                                                                                                                                                                                     |
+----------------------------------------------+-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
