---
title: "Chapter 5: Lesson 3"
subtitle: "Title Goes Here"
format: html
editor: source
sidebar: false
---

```{r}
#| include: false
source("../common_functions.R")
```

```{=html}
<script type="text/javascript">
 function showhide(id) {
    var e = document.getElementById(id);
    e.style.display = (e.style.display == 'block') ? 'none' : 'block';
 }
 
 function openTab(evt, tabName) {
    var i, tabcontent, tablinks;
    tabcontent = document.getElementsByClassName("tabcontent");
    for (i = 0; i < tabcontent.length; i++) {
        tabcontent[i].style.display = "none";
    }
    tablinks = document.getElementsByClassName("tablinks");
    for (i = 0; i < tablinks.length; i++) {
        tablinks[i].className = tablinks[i].className.replace(" active", "");
    }
    document.getElementById(tabName).style.display = "block";
    evt.currentTarget.className += " active";
 }    
</script>
```

## Learning Outcomes

<details>
<summary>Interpret the key statistics used to describe sample data</summary>
    -   Interpret the sample mean
    -   Interpret the sample variance
    -   Interpret the sample standard deviation
</details>
<details>
<summary>Compute the key statistics used to describe the linear relationship between two variables</summary>
    -   Compute the sample mean
    -   Compute the sample variance
    -   Compute the sample standard deviation

</details>


## Preparation

-   Read Sections 2.1-2.2.2 and 2.2.4

## Learning Journal Exchange (10 min)

-   Review another student's journal
-   What would you add to your learning journal after reading your partner's?
-   What would you recommend your partner add to their learning journal?
-   Sign the Learning Journal review sheet for your peer

## Class Activity: Variance and Standard Deviation (15 min)

We will explore the variance and standard deviation in this section.

-   What do the standard deviation and the variance measure?

The following code simulates observations of a random variable. We will use these data to explore the variance and standard deviation.

```{r}

```

