---
title: "Time Series Assessment: Final W24"
subtitle: "Please_put_your_name_here"
format: 
  html:
    embed-resources: true
    toc: true
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}

# Loading R packages

if (!require("pacman")) install.packages("pacman")
pacman::p_load(nlme,tsibble, fable,feasts, tsibbledata,
               fable.prophet, patchwork,lubridate, 
               rio,ggplot2,kableExtra,tidyquant, 
               dplyr , plotly,tidymodels, stats,
               multilevelmod, broom.mixed,
               forecast
)
```

## Data

```{r, message=FALSE}

non_store_ts <- rio::import("https://byuistats.github.io/timeseries/data/retail_by_business_type.parquet") |>
  filter(naics == 454) |>
  as_tsibble(index = month)

term_prem10 <- rio::import("https://byuistats.github.io/timeseries/data/term_prem10.csv")


```

## Questions

### Question 1 - Non- Store Retailers (200 points)

Over the last 20 years, the part of the retail world that doesn't use traditional stores, classified under the NAICS code 454, has grown significantly. This includes online shopping, mail orders, vending machines, and home deliveries, etc. Among these, online shopping has become especially important, leading the growth in this sector. The series `non-store_ts` shows the sales in millions for this category starting in the early 90's. Your client is a start-up company entering the space, and needs a forecast of the category for the next 10 years. The audience will be primarily venture capital firms looking to invest in the start-up. However, your report will also serve as the main source of information for the structuring of the company's supply chain risk assessment and purchasing strategy. Assume that half of your audience will have similar or higher technical expertise than you, and the rest of your audience has an MBA. Your report should contain a discussion of [all]{.underline} the features of the series and analysis of the random component of the series in the context of uncertainty in business planning. Your analysis should use [all]{.underline} the tools taught in the course that are appropriate for the setting. Please include a [justification]{.underline} of model selection and a discussion of the strengths and limitations of your forecast.

::: {.callout-note title="Answer" icon="false"}
```{r}



```

<!-- Delete this and put your answer here. The three colons mark the end of your answer. -->
:::

### Question 2 - (150 points)

The term premium refers to the extra yield that investors require to commit to holding a longer-term bond, such as a 10-year Treasury, instead of rolling over shorter-term securities over the same period. Essentially, it compensates investors for the additional risks associated with longer maturities, including interest rate risk, inflation risk, and the uncertainty over the future path of short-term interest rates. Analyzing the term premium provides valuable insights into investor expectations regarding future economic conditions, monetary policy actions, and other factors that influence the yield curve. For you, this exploration offers a practical application of time series analysis techniques to real-world economic and financial data. For this section, pretend you are a financial analyst who has applied to a job with a financial analytics firm. They have provided the `term_prem10` series and your task is to explore the term premium's temporal structure. Your discussion should center around which stochastic-process model is the most appropriate for the realizations of the time series. Your audience is a potential employer who is trying to asses your understanding of time series models, such as autoregressive (AR), moving average (MA), ARMA and autoregressive integrated moving average (ARIMA).

::: {.callout-note title="Answer" icon="false"}
```{r}



```

<!-- Delete this and put your answer here. The three colons mark the end of your answer. -->
:::
