---
title: "Time Series Homework: Chapter 3 Lesson 4"
subtitle: "Please_put_your_name_here"
format: 
  html:
    embed-resources: true
    toc: true
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}

# Loading R packages
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, 
               tsibble, fable,
               feasts, tsibbledata,
               fable.prophet,
               patchwork,
               lubridate, 
               rio,
               kableExtra
               )
```

## Data

```{r, message=FALSE}

avgkwhr <- rio::import("https://byuistats.github.io/timeseries/data/USENERGYPRICE.csv")
avgkwhr <- read_csv("../../data/USENERGYPRICE2.csv")

```

## Questions

### Question 1 - Context and Measurement (10 points)

The first part of any time series analysis is context. You cannot properly analyze data without knowing what the data is measuring. Without context, the most simple features of data can be obscure and inscrutable. This homework assignment will center around the series below.

Please research the time series. In the spaces below, give the data collection process, unit of analysis, and meaning of each observation for the series.

**Average Price of Electricity per Kilowatt-Hour in U.S.: City Average**

<https://fred.stlouisfed.org/series/APU000072610>

::: {.callout-note title="Answer" icon="false"}
#### Data Collection Process

The data is sourced from the U.S. Bureau of Labor Statistics (BLS) under the release “Average Price: Electricity per Kilowatt-Hour in U.S. City Average. The prices are collected monthly from 75 urban areas across the United States using mail questionnaires managed by the Department of Energy. All reported prices include Federal, State, and local taxes, as well as fuel and purchased gas adjustments for natural gas and electricity.

#### Unit of Analysis

The primary unit is the U.S. City Average electricity price (\$) per kilowatt-hour (kWh) across the 75 urban areas. Prices are measured in U.S. Dollars (USD) and are not seasonally adjusted, reflecting actual monthly prices without adjustments for seasonal variations.

#### Meaning of Each Observation

Each data point represents the average electricity price per kilowatt-hour for a specific month in a U.S. city.
:::

### Question 2 - US Average Price of Electricity: Additive Holt-Winters Forecasting (25 points)

#### a) Please use the Holt-Winters smoothing method to the series.

::: {.callout-caution collapse="true"}
## Missing data

We will discuss this more in class but for those who are curious in this dataset originally there was no value for September 1985. In this case with only one seemingly random value missing from our data the implications of how we fill the value are less impactful, regardless the method chosen here was to average the values from $t_{-1}$ & $t_{+1}$ then use the resulting value to fill the missing data. This can be accomplished quickly using the following base R code:

```{r}
#| eval: false
#| warning: false
# Average the values from the periods before(82) and after(84) the missing period(83).
value <- (avgkwhr_tsbl$usenergyprice[82]+avgkwhr_tsbl$usenergyprice[84])/2
# this evaluates to 0.835

# Fill missing value with average of surrounding values
avgkwhr_tsbl$usenergyprice[83] <- value


```

For your convience the data provided for this Homework assignment has already been filled.

:::

::: {.callout-note title="Answer" icon="false"}
```{r}
## Wrangling
avgkwhr$yearmonth <- yearmonth(lubridate::mdy(avgkwhr$date))
avgkwhr_tsbl <- as_tsibble(avgkwhr, index=yearmonth)

# Fill missing value with average of surrounding values
avgkwhr_tsbl$usenergyprice[83] <- 0.0835

# First the series itself was plotted, you can use any method/function for this but before attempting a Holt-Winters decomposition we want to visualize the series and deterine roughly what trend, seasonality, & cycles look like in the series.
autoplot(avgkwhr_tsbl)

avgkwhr_hw <- avgkwhr_tsbl |>
  model(Additive = ETS(usenergyprice ~
        trend("A", alpha = 0.82, beta = 0.01) +
        error("A") +
        season("A", gamma = 0.02),
        opt_crit = "amse", nmse = 1))
report(avgkwhr_hw)
```

```{r}
autoplot(components(avgkwhr_hw))

augment(avgkwhr_hw) |>
  ggplot(aes(x = yearmonth, y = usenergyprice)) +
    #coord_cartesian(xlim = yearmonth(c("1978 Nov","1990 Dec"))) +
    coord_cartesian(xlim = yearmonth(c("2000 Nov","2016 Dec"))) +
    geom_line() +
    geom_line(aes(y = .fitted, color = "Fitted")) +
    labs(color = "")
```
:::

```{r}
avgkwhr_hw <- avgkwhr_tsbl |>
  model(Additive = ETS(usenergyprice ~
        trend("A", alpha = 0.01, beta = 0.0001) +
        error("A") +
        season("A", gamma = 0.02),
        opt_crit = "amse", nmse = 1))
report(avgkwhr_hw)

autoplot(components(avgkwhr_hw))
augment(avgkwhr_hw) |>
  ggplot(aes(x = yearmonth, y = usenergyprice)) +
    #coord_cartesian(xlim = yearmonth(c("1978 Nov","1984 Dec"))) +
    #coord_cartesian(xlim = yearmonth(c("2010 Nov","2016 Dec"))) +
    geom_line() +
    geom_line(aes(y = .fitted, color = "Fitted")) +
    labs(color = "")
```



#### b) What parameters values did you choose for $\alpha$, $\beta$, and $\gamma$. Justify your choice.

::: {.callout-note title="Answer" icon="false"}
<!-- Delete this and put your answer here. The three colons mark the end of your answer. -->
:::

#### c) Please plot the Holt-Winters forecast of the series for the next 36 months superimposed against the original series. Please see [Figure 7](https://byuistats.github.io/timeseries/chapter_3_lesson_4.html#fig-crime-hw-forecast) in Chapter 3: Lesson 4

::: {.callout-note title="Answer" icon="false"}
```{r}

avgkwhr_forecast <- avgkwhr_hw |>
  forecast(h = "3 years") 
avgkwhr_forecast |>
  autoplot(avgkwhr_tsbl, level = 95) +
  #coord_cartesian(ylim = c(0,5500)) +
  geom_line(aes(y = .fitted, color = "Fitted"),
    data = augment(avgkwhr_hw)) +
  scale_color_discrete(name = "")

```
:::

#### d) Is the trend in the US Average Price of Electricity series deterministic or stochastic? What is the basis of your evaluation?

:::: {.callout-note title="Answer" icon="false"}
Evaluting the series as a whole, it contains more deterministic characteristics than stochastic characteristics.

##### Trend

A loosely constant positive upwards trend is apparent, meaning that overtime the price of electricity is rising. When we examine this with context however it seems strange to say that the cost of electricity itself is rising, in theory it would decrease overtime as we would expect technological advancements to make it easier and more efficient to produce electricity. *However in practice this is somewhat offset by pressure to shift towards renewables which are more expensive*. The trend we observe is more reasonably linked to the economy, when the price of a good rises overtime independent of changes in supply/demand we are observing inflation. So the trend in this series is not actually a result of energy prices themselves but rather the inflation in the economy the prices were collected from.

##### Seasonality

The seasonality appears very constant and deterministic, with an annual period, cycling every 12 months, with high values in the summer (everyone turns on the A/C) and lower values in the winter.

##### Cycles & 'Random' Shocks/Events

However stochastic cycles or rather level shocks are observed, ex: 1987, 1998, 2008-2009, & 2014. There's lots of factors that could be at play here so this is an excellent time to try and harness AI for some help.

Using this prompt and providing a screenshot of the timeseries plot the following *potentially related* context was obtained.

*"I have this data on the average price of energy in the usa from 1980-2020. Im specifically interested in the context and reason that I can observe cycles/level shocks around these time periods: 1987, 1998, 2008-2009, 2014. Can you tell me about world event, political events, policy changes, technological advancements that happened around these timeframes that would be the explanation for these occurrences?"*

::: {.panel-tabset .nav-pills}

## 1987

During the early/mid 1980's prices were rapidly on the rise, roughly around 1987 we observe a drop and then a slower trend upwards until the next shock in 1998.

*Event:* Black Monday stock market crash (October 19, 1987).

*Context:* The Dow Jones Industrial Average fell by 22.6%, the largest single-day drop in history. Worldwide losses were estimated at US$1.71 trillion.

*Economic Factors:* Concerns over inflation, a weakening dollar, rising oil and commodity prices, protectionism, tensions between the United States and Iran, and the potential end to a long economic expansion contributed to investor jitters.

*Policy/Technological Factors:* Deregulation of wellhead prices for natural gas and the pipeline industry were unleashing market forces, making natural gas and gas-fired power less expensive.

## 1998

Around 1997-1998, we can observe a steep drop in prices follwed by roughly a decade of slowed price growth. World events impacted US policy in favor of spending/investment, most importantly, federal institutions reorganized electrical markets breaking apart the industry and promoting increased competition. This drove prices down in both the short and long-term.

*Event:* Russian Financial Crisis (Ruble Crisis) and Asian Financial Crisis.

*Context:* The Russian government devalued the ruble and defaulted on its debt in August 1998. The Asian Financial Crisis had begun in 1997.

*Economic Factors:* The Russian crisis had severe impacts on neighboring economies and shook investor confidence in global financial markets. International stock markets hit record lows.

*Energy Sector Impact:* Economic downturns in Russia and Asia may have reduced energy demand in those regions, potentially impacting global prices. In the U.S., the Federal Reserve cut interest rates to safeguard the economy against systemic financial crises.

*Policy/Technological Factors:* FERC finalized orders to significantly restructure natural gas markets. Policy makers were also incentivizing alternative technologies, such as wind power.

## 2008-2009

The housing crash and subsequent bank failures lead to a severe resession in the US (and world).

????????????????????????????????????????????????????????????????
WHY DID PRICES RISE? not fall...?

**EDIT** I dug around for like 2 hours yesterday and tried a bunch of different series in FRED.
The best thing I was able to find Is that in 2005 the US suffered a few major hurricanes that destroyed a substaintial amount of NG infrastrcutre in the south eastern USA, somewhat connected to this, the prices of NG rose significantly at the end of 2005, however the quickly fell back to more normal levels during 2006/2007 before rising again during the 2008/2009 crisis. 2006 saw a large level rise in electricity expenses *so a 6 month lagged effect from NG prices in 2005?* and then prices also rose during the 2008/2009 crisis. A similar 6 month lagged affect was observed in NG prices rising in Jan 2001 with electric prices rising in Jul 2001. 
Additionally during 2005, the Energy Policy Act was passed, it was lack luster and disappointing for the NG industry who was hoping for improved feasability for LNG terminal construction (Liquid Nat Gas), which may have lowered outlooks in the industry significantly in a somewhat more long-term effect... I'm far from an expert though.

*Event:* Global Financial Crisis/Great Recession.

*Context:* This was a major worldwide economic crisis that began in 2007 and extended through 2009. The crisis stemmed from predatory lending, a U.S. housing bubble, and excessive risk-taking by financial institutions.

*Economic Factors:* The crisis led to bank failures, plummeting stock and commodity prices, and a global economic slowdown. Unemployment soared.

*Energy Sector Impact:* The recession caused a sharp drop in energy demand and a collapse in oil prices. Financial institutions sank, and the US Government was forced to offer bailouts.

*Policy/Technological Factors:* The American Recovery and Reinvestment Act of 2009 included provisions for investments in renewable energy and energy efficiency, helping to stabilize markets in the early 2010's.

## 2014

In the early 2010's we can observe a slow but distinguishable upwards trend, however around 2014 there is a spike and then the upwards trend all but disappears leading up to the present day. This is primarily a result of the US Shale revolution, leading to a large & long term increase in natural gas supply.

Event: Drop in Oil Prices, Geopolitical Tensions in the middle east

Context: Oil prices experienced a significant decline in 2014.

Economic Factors: Several factors contributed to this, including increased U.S. oil production due to fracking. OPEC's decision not to cut production in response to falling prices exacerbated the oversupply and further contributed to the price decline in late 2014.

Energy Sector Impact: The drop in oil prices affected energy markets globally.

Policy/Technological Factors: The U.S. shale energy revolution increased domestic production. The US lifted its crude export ban in 2015.

:::

::::

### Question 3 - Real US Average Price of Electricity: Additive Holt-Winters Forecasting (25 points)

The upward trend of the series is mostly due to inflation, the generalized increase in prices throughout the economy. One way to quantify inflation is to use a price index, like the Personal Consumption Expenditures Deflator (PCE).The series [HERE](https://fred.stlouisfed.org/series/PCEPILFE) shows that prices in the US have climbed steadily over the last 60 years.Because energy is an important part of the economy, it's likely that energy prices have followed a similar pattern. Adjusting a series with nominal prices, like the price of electricity, to real prices that account for inflation is simple, divide the original series by the price index. The data set imported below is the real price of electricity, which is the US Average Price of Electricity divided by the PCE index excluding food and energy prices (PCEPILFE). Repeat steps a) to d) for the updated series.

::: {.callout-note title="Answer" icon="false"}
```{r, message=FALSE}

real_avgkwhr <- rio::import("https://byuistats.github.io/timeseries/data/USENERGYPRICE_Real.csv")
real_avgkwhr <- read_csv("../../data/USENERGYPRICE_Real2.csv")

```
:::

#### a) Please use the Holt-Winters smoothing method to the series.

::: {.callout-note title="Answer" icon="false"}
```{r}

real_avgkwhr$yearmonth <- yearmonth(lubridate::mdy(real_avgkwhr$date))
real_avgkwhr_tsbl <- as_tsibble(real_avgkwhr, index=yearmonth)

autoplot(real_avgkwhr_tsbl)

real_avgkwhr_hw <- real_avgkwhr_tsbl |>
  model(Additive = ETS(realprices ~
        trend("A", alpha = 0.3, beta = 0.1) +
        error("A") +
        season("A", gamma = 0.1),
        opt_crit = "amse", nmse = 1))
report(real_avgkwhr_hw)
```

```{r}
autoplot(components(real_avgkwhr_hw))
augment(real_avgkwhr_hw) |>
  ggplot(aes(x = yearmonth, y = realprices)) +
    #coord_cartesian(xlim = yearmonth(c("1978 Nov","1990 Dec"))) +
    coord_cartesian(xlim = yearmonth(c("2014 Nov","2016 Dec"))) +
    geom_line() +
    geom_line(aes(y = .fitted, color = "Fitted")) +
    labs(color = "")
```
:::

#### b) What parameters values did you choose for $\alpha$, $\beta$, and $\gamma$. Justify your choice.

::: {.callout-note title="Answer" icon="false"}

In comparison to the original series, the real prices seem to be fluctuating around and average value (roughly $0.0014), additionally the cycles/level shifts around the previously mention periods are now much clearer.

So we can set a Low alpha/beta to ignore the cycles? or high alpha/beta because we want to follow them?

When examining the seasonality of the series somewhere around 2020 there seems to be a significant shift in the pattern
:::

#### c) Please plot the Holt-Winters forecast of the series for the next 12 months superimposed against the original series. Please see [Figure 7](https://byuistats.github.io/timeseries/chapter_3_lesson_3.html#fig-crime-hw-forecast) in Chapter 3: Lesson 3

::: {.callout-note title="Answer" icon="false"}
```{r}



```
:::

#### d) Is the trend in the US Average Real Price of Electricity series deterministic or stochastic? What is the basis of your evaluation?

::: {.callout-note title="Answer" icon="false"}
<!-- Delete this and put your answer here. The three colons mark the end of your answer. -->
:::

### Rubric

|  |  |  |
|-----------------|----------------------------|---------------------------|
| **Criteria** | **Mastery (10)** | **Incomplete (0)** |
| **Question 1: Context and Measurement** | The student thoroughly researches the data collection process, unit of analysis, and meaning of each observation for both the requested time series. Clear and comprehensive explanations are provided. | The student does not adequately research or provide information on the data collection process, unit of analysis, and meaning of each observation for the specified series. |
|  | **Mastery (5)** | **Incomplete (0)** |
| **Question 3a: HW Smoothing** | Demonstrate the implementation of the Holt-Winters smoothing method in R, providing well-commented code that clearly explains each step of the algorithm. They correctly specify the necessary parameters, including trend and seasonality components. | Students encounter difficulties in accurately implementing the Holt-Winters smoothing method in R. Their code may lack sufficient comments or clarity, making it challenging to understand the implementation process. Additionally, they may overlook important parameters or make errors in the application of the method, leading to inaccuracies in the results. |
|  | **Mastery (10)** | **Incomplete (0)** |
| **Question 3b: Parameter Choice** | Responses not only specify the chosen parameter values for \$\\alpha\$, \$\\beta\$, and \$\\gamma\$ in the context of the Holt-Winters smoothing method but also correctly identify the purpose of each parameter in their explanation. They provide a thorough justification for each parameter choice, considering factors such as the data characteristics, seasonality patterns, and the desired level of smoothing | Student struggles to clearly specify the chosen parameter values for \$\\alpha\$, \$\\beta\$, and \$\\gamma\$. It's no clear that they understand the purpose of each parameter in their explanation. They may provide limited or vague justification for each parameter choice, lacking consideration of important factors such as data characteristics or seasonality patterns. |
|  | **Mastery (5)** | **Incomplete (0)** |
| **Question 3c: Forecast Plot** | Responses effectively create a plot of the Holt-Winters forecast for the next 24 months superimposed against the original series in R. The forecasted values align with the original series and display relevant trends and seasonality patterns. Additionally, they appropriately label the axes, title the plot, and provide a clear legend to distinguish between the original series and the forecast. The plot closely resembles Figure 7 in the Time Series Notebook | Student encounter challenges in creating a plot of the Holt-Winters forecast. They may struggle with accurately implementing the plotting code, resulting in inaccuracies or inconsistencies in the plotted forecast. Additionally, their plot may lack proper labeling of the axes, a title, or a legend, making it difficult to interpret the information presented. Furthermore, their plot may deviate significantly from Figure 7 in the Time Series Notebook. |
|  | **Mastery (5)** | **Incomplete (0)** |
| **Question 3d: Trend Evaluation** | The submission demonstrate a clear understanding of the distinction between deterministic and stochastic trends and provide a reasoned argument for their assessment based on the observed data properties. They provide anevaluation of the data characteristics, considering factors such as the presence of consistent patterns or irregular fluctuations over time. Analyses involve visual inspections to identify any discernible patterns or randomness in the trend. | Student offers limited insights into the data characteristics, lacking consideration of relevant factors such as patterns or fluctuations over time. Additionally, their evaluation may lack depth or coherence. No plots drawn to evaluate the trend. |
|  | **Mastery (5)** | **Incomplete (0)** |
| **Question 3a: HW Smoothing** | Demonstrate the implementation of the Holt-Winters smoothing method in R, providing well-commented code that clearly explains each step of the algorithm. They correctly specify the necessary parameters, including trend and seasonality components. | Students encounter difficulties in accurately implementing the Holt-Winters smoothing method in R. Their code may lack sufficient comments or clarity, making it challenging to understand the implementation process. Additionally, they may overlook important parameters or make errors in the application of the method, leading to inaccuracies in the results. |
|  | **Mastery (10)** | **Incomplete (0)** |
| **Question 3b: Parameter Choice** | Responses not only specify the chosen parameter values for \$\\alpha\$, \$\\beta\$, and \$\\gamma\$ in the context of the Holt-Winters smoothing method but also correctly identify the purpose of each parameter in their explanation. They provide a thorough justification for each parameter choice, considering factors such as the data characteristics, seasonality patterns, and the desired level of smoothing | Student struggles to clearly specify the chosen parameter values for \$\\alpha\$, \$\\beta\$, and \$\\gamma\$. It's no clear that they understand the purpose of each parameter in their explanation. They may provide limited or vague justification for each parameter choice, lacking consideration of important factors such as data characteristics or seasonality patterns. |
|  | **Mastery (5)** | **Incomplete (0)** |
| **Question 3c: Forecast Plot** | Responses effectively create a plot of the Holt-Winters forecast for the next 24 months superimposed against the original series in R. The forecasted values align with the original series and display relevant trends and seasonality patterns. Additionally, they appropriately label the axes, title the plot, and provide a clear legend to distinguish between the original series and the forecast. The plot closely resembles Figure 7 in the Time Series Notebook | Student encounter challenges in creating a plot of the Holt-Winters forecast. They may struggle with accurately implementing the plotting code, resulting in inaccuracies or inconsistencies in the plotted forecast. Additionally, their plot may lack proper labeling of the axes, a title, or a legend, making it difficult to interpret the information presented. Furthermore, their plot may deviate significantly from Figure 7 in the Time Series Notebook. |
|  | **Mastery (5)** | **Incomplete (0)** |
| **Question 3d: Trend Evaluation** | The submission demonstrate a clear understanding of the distinction between deterministic and stochastic trends and provide a reasoned argument for their assessment based on the observed data properties. They provide anevaluation of the data characteristics, considering factors such as the presence of consistent patterns or irregular fluctuations over time. Analyses involve visual inspections to identify any discernible patterns or randomness in the trend. | Student offers limited insights into the data characteristics, lacking consideration of relevant factors such as patterns or fluctuations over time. Additionally, their evaluation may lack depth or coherence. No plots drawn to evaluate the trend. |
| **Total Points** | **60** |  |
